### 2021-01-19 README.Rmd
init
revisiting the repo after some time

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# settlements

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/petrpajdla/settlements/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Trampota F., Pajdla P., (2021). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

The simplest way to explore the text, code and data is to click on
[binder](https://mybinder.org/v2/gh/petrpajdla/settlements/master?urlpath=rstudio)
to open an instance of RStudio in your browser, which will have the
compendium files ready to work with. Binder uses rocker-project.org
Docker images to ensure a consistent and reproducible computational
environment. These Docker images can also be used locally.

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2021-02-10 analysis/code/dataprep.R
update dataprep pipeline

@@ -1,319 +1,51 @@
 # Project "Settlements"
 # Script nr. 0
-# DATA PREPARATION
+# DATA PREPARATION II
 # author: Petr Pajdla
-# Prepares data for the analysis from raw spreadsheet formats
+# Downloads updated data from GD
 
 library(here)
 library(tidyverse)
-library(readxl)
 library(sf)
+library(googledrive)
 
-dt_raw <- "analysis/data/raw_data"
+# paths
 dt_der <- "analysis/data/derived_data"
+gd_path <- "~/settlements/data/database/"
 
-# funs --------------------------------------------------------------------
+# dir.create(here(dt_der, "input_gd"))
 
-indicator_to_logical <- function(x) {
-  x <- if_else(str_detect(x, "1"), "TRUE", x)
-  x <- if_else(str_detect(x, "0"), "FALSE", x)
-  x <- if_else(str_detect(x, "\\?"), "TRUE", x)
-  return(as.logical(x))
+# get updated data from drive
+files_gd <- drive_ls(gd_path)
+
+for (i in seq_along(files_gd$id)) {
+  drive_download(as_id(files_gd[i, ]),
+                 here(dt_der, "input_gd", files_gd$name)[i],
+                 type = "csv", overwrite = TRUE)
 }
 
-# data input --------------------------------------------------------------
 
-in_morava <- read_excel(here(dt_raw, "morava_pp.xlsx"))
+# prepare derived data ----------------------------------------------------
 
-in_cechy <- read_excel(here(dt_raw, "cechy_pp.xlsx"))
 
-# morava ------------------------------------------------------------------
-m <- in_morava %>%
-  rowid_to_column() %>%
-  # correct phase labels
-  rename(SBK_Late = SBK_IV,
-         'TRB_Boleraz/Saalzm' = TRB_Boleraz) %>%
-  unite(col = "ENE_Proto", LgK_EPI, Ludanice, na.rm = TRUE) %>%
-  unite(col = "SBK_Early", `SBK_II-III`, SBK_III, na.rm = TRUE) %>%
-  unite(col = "LgK_II_b", LgK_II_b, Brodzany, na.rm = TRUE) %>%
-  unite(col = "TRB_Baalberg", TRB_Baalberg, TRB_I, na.rm = TRUE) %>%
-  mutate(across(.cols = c(SBK:LgK_II_a), as.character),
[…]

### 2021-02-10 analysis/code/dataprep_hydro.R

@@ -1,0 +1,49 @@
+
+library(sf)
+library(here)
+library(tidyverse)
+
+# read data
+dt_hydro <- "analysis/data/raw_data/hydro/"
+
+# hydrology
+cz <- st_read(here(dt_hydro, "hydro_czechia"))
+
+au <- st_read(here(dt_hydro, "hydro_austria"))
+
+sk <- st_read(here(dt_hydro, "hydro_slovakia"))
+
+# project mask
+mask <- st_read(here(here("analysis/data/derived_data/maps/", "mask.geojson")))
+
+# transform to S-JTSK and crop to mask
+au_crop <- au %>%
+  st_zm() %>%
+  st_transform(st_crs(mask)) %>%
+  st_intersection(mask)
+
+sk_crop <- sk %>%
+  st_transform(st_crs(mask)) %>%
+  st_intersection(mask)
+
+cz_crop <- cz %>%
+  st_intersection(mask)
+
+# control plot
+ggplot() +
+  geom_sf(data = au_crop, color = "green") +
+  geom_sf(data = cz_crop, color = "blue") +
+  geom_sf(data = sk_crop, color = "orange") +
+  theme_void()
+
+# join layers
+temp <- st_join(st_as_sf(st_geometry(au_crop)),
+                st_as_sf(st_geometry(sk_crop)))
+x <- st_join(temp, st_as_sf(st_geometry(cz_crop)))
+
+
+ggplot() +
+  geom_sf(data = x) +
+  theme_void()
+
+st_union()

### 2021-02-10 analysis/code/dataprep_init.R

@@ -1,0 +1,349 @@
+# # Project "Settlements"
+# # Script nr. 0
+# # DATA PREPARATION
+# # author: Petr Pajdla
+# # Prepares data for the analysis from raw spreadsheet formats
+#
+# library(here)
+# library(tidyverse)
+# library(sf)
+# library(googledrive)
+#
+# dt_raw <- "analysis/data/raw_data"
+# dt_der <- "analysis/data/derived_data"
+#
+# # funs --------------------------------------------------------------------
+#
+# # 1 / 0 indicators to logical TRUE / FALSE
+# indicator_to_logical <- function(x) {
+#   x <- as.character(x)
+#   x <- if_else(str_detect(x, "1"), "TRUE", x)
+#   x <- if_else(str_detect(x, "0"), "FALSE", x)
+#   x <- if_else(str_detect(x, "\\?"), "TRUE", x)
+#   return(as.logical(x))
+# }
+#
+# # creates chronologies from indicators
+# create_chrono <- function(x, chrono) {
+#   stopifnot(chrono %in% c(1, 2, 3))
+#
+#   if (chrono == 1) {
+#     chrono_cols <- c("SBK", "LgK", "ENE", "TRB")
+#   } else if (chrono == 2) {
+#     chrono_cols <- c("SBK1", "SBK2", "LgK1", "LgK2", "ENE0",
+#                      "TRB1", "TRB2")
+#   } else if (chrono == 3) {
+#     chrono_cols = c("ENE0_Epilengyel", "ENE0_Ludanice",
+#                     "ENE0_Schussenried", "ENE0_Michelsberg", "ENE0_Jordanow")
+#   }
+#
+#   x %>%
+#     select(id, any_of(chrono_cols)) %>%
+#     pivot_longer(-id,
+#                  names_to = "chrono",
+#                  values_to = "chrono_dubious", values_drop_na = TRUE) %>%
+#     # filter only certain sites
+#     filter(!str_detect(chrono_dubious, "\\?")) %>%
+#     select(-chrono_dubious)
+# }
+#
+# # create dataset
+# create_data_set <- function(x, pad) {
+#   stopifnot(pad %in% c("M", "B"))
+#
+#   res <- list(
+#     data = NA,
+#     chrono1 = NA,
+#     chrono2 = NA
+#   )
+#
+#   # unique ids
+#   dt <- x %>% mutate(id = str_pad(rowid, width = 4, side = "left", pad = "0"),
[…]

### 2021-02-10 analysis/code/dataprep_map.R

@@ -4,9 +4,10 @@
 # author: Petr Pajdla
 # Prepare background data for maps (crop, transform etc.)
 # Data from Natural Earth Project using rnaturalearth package
+# Own data for raw materials
 
-library(tidyverse)
 library(here)
+library(tidyverse)
 library(sf)
 library(rnaturalearth)
 
@@ -64,6 +65,28 @@ admin1 <- ne_load(scale = 10,
                   returnclass = "sf") %>%
   crop_and_transform()
 
+
+# raw materials -----------------------------------------------------------
+# lithics rm
+si_pts <- st_read(dsn = here(dt_raw, "suroviny/"), layer = "surovina") %>%
+  st_set_crs(5514) %>%
+  rename(rm = Surovina) %>%
+  mutate(orig = "l")
+si_lne <- st_read(dsn = here(dt_raw, "suroviny/"), layer = "SGS") %>%
+  rename(rm = surovina) %>%
+  mutate(orig = "l")
+
+# polished stone rm
+bi_pts <- st_read(dsn = here(dt_raw, "suroviny/"), layer = "surovinaBI") %>%
+  rename(rm = surovina) %>%
+  mutate(orig = "p")
+bi_lne <- st_read(dsn = here(dt_raw, "suroviny/"), layer = "surovinaBI_linie") %>%
+  rename(rm = surovina) %>%
+  mutate(orig = "p")
+
+rm_pts <- bind_rows(si_pts, bi_pts)
+rm_lne <- bind_rows(si_lne, bi_lne)
+
 # plots -------------------------------------------------------------------
 
 ggplot() +
@@ -75,6 +98,8 @@ ggplot() +
   # geom_rect(aes(xmin = bbox[1], xmax = bbox[3],
   #               ymin = bbox[2], ymax = bbox[4]), fill = NA, color = "black") +
   geom_sf(data = bbox, fill = NA) +
+  geom_sf(data = rm_pts, shape = 4) +
+  geom_sf(data = rm_lne) +
   theme_void()
 
 # write derived datasets --------------------------------------------------
@@ -92,3 +117,6 @@ geojson(admin1)
 geojson(mask)
 geojson(bbox_large)
 geojson(bbox)
+
+st_write(rm_lne, here(dt_der, "rm_lines.geojson"), delete_dsn = TRUE)
+st_write(rm_pts, here(dt_der, "rm_points.geojson"), delete_dsn = TRUE)

### 2021-02-10 analysis/paper/paper.Rmd
continue R/ and paper/

---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
#  "This is abstract"
keywords: |
  keyword 1; keyword 2; keyword 3
---

Keywords: `r rmarkdown::metadata$keywords`

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages}
devtools::load_all(".")
library(tidyverse)
library(sf)
```

# Introduction

The analysis is divided into two 'time' scales:

* Pottery traditions
* Pottery groups

The study area is broadly defined as *Easter part of Bohemia* (abbreviated *B*) 
and *Morava river catchment* (*M*).
It spans across the area of the Czech Republic and bordering regions of 
Austria and Slovakia.

<!-- ======================================================================= -->

# Background

```{r get-data, cache = FALSE}
derived_data <- "analysis/data/derived_data"
path_maps <- here::here(derived_data, "maps/")
path_data <- here::here(derived_data)
path_temp <- here::here(derived_data, "temp/")

# read data ---------------------------------------------------------------
# mask
mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)

# rivers + admin for plots
admin0 <- st_read(paste0(path_maps, "admin0.geojson"), quiet = TRUE)
admin1 <- st_read(paste0(path_maps, "admin1.geojson"), quiet = TRUE)
rivers0 <- st_read(paste0(path_maps, "rivers0.geojson"), quiet = TRUE)
rivers1 <- st_read(paste0(path_maps, "rivers1.geojson"), quiet = TRUE)

# labels
labs_chrono <- read_rds(here::here(derived_data, "chrono_labels.RDS"))

# database
set_base <- read_rds(here::here(derived_data, "settlements.RDS"))

set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"), 
                    quiet = TRUE)

settlements1 <- set_spat %>%
  right_join(set_base$period1, by = "id")
settlements2 <- set_spat %>%
  right_join(set_base$period2, by = "id")
```

<!--
```{r base-tab}
# set_base$base %>%
#   DT::datatable(rownames = FALSE)
```
-->

## Pottery traditions

```{r tab-set1-count}
set_base$period1 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set1-count, fig.cap = 'Settlements classified by general pottery traditions across time periods.'}
settlements1 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

<!-- ----------------------------------------------------------------------- -->

## Pottery groups

```{r tab-set2-count}
set_base$period2 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set2-count, fig.cap = 'Settlements classified by pottery groups across time periods.'}
settlements2 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

<!-- ======================================================================= -->

# Settlement density

Settlement density is estimated using KDE.
Kernel size of 4 km is used. 
This value is completely arbitrary but is selected because settlement clusters 
are nicely highlighted at this scale.

## Pottery traditions

```{r set1-kde}
kde1 <- kdestimate(settlements1,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), kde)
```

```{r plot-set1-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements1,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

<!-- ----------------------------------------------------------------------- -->

## Pottery groups

```{r set2-kde}
kde2 <- kdestimate(settlements2,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), kde)
```

```{r plot-set2-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements2,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

<!-- ======================================================================= -->

# Distance to raw material sources

## Raw materials distribution

Raw material sources are defined as either points or lines.
Each raw material can be identified by single or multiple simple features.
If the location of source used in the Neolithic period is unknown or several sources or even
modes of procurement (e. g. from river bed etc.) are possible, the most probable 
are included.

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
```

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```
```{r rm-dist, cache = FALSE}
# calculated separately in a script `rm_distance.R`
# dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
# dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))
```

```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
rm_dist %>%
  ggplot(aes(x = min, y = rm, fill = region)) +
  geom_violin() +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~type, scales = "free_y") +
  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

<!-- ======================================================================= -->

<!--\@ref(fig:demo-plot)-->

<!-- The following line inserts a page break -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2021-02-10 R/distances.R

@@ -1,0 +1,58 @@
+#' Distance to a source of a raw material
+#'
+#' For simple feature geometries in projected coordinate reference system,
+#' euclidean distance (in km) is returned between features of \code{from}
+#' and \code{to} categorized by levels specified in \code{rm}.
+#'
+#' @param from \code{sf} object of geometry type \code{POINTS} in projected
+#'     CRS (settlements).
+#' @param to \code{sf} object of geometry type \code{POINT} or \code{LINESTRING}
+#'     in projected CRS (raw material locations).
+#'     Each raw material can have multiple locations, same raw materials are
+#'     identified by identical label in column specified by \code{rm} argument.
+#' @param rm Column name of \code{to} to use as a grouping variable
+#'     (identical raw material).
+#' @param id Column name of \code{from} containing unique identifiers.
+#'
+#' @return A \code{tibble} containing min, max, mean and median distance
+#'     for each raw material (level of \code{rm} column) to an observation
+#'     specified in \code{from} (settlement) is returned.
+#'
+dist_to_rm <- function(from, to, rm, id = "id") {
+
+  # set units to km
+  to_km <- function(x) {
+    units(x) <- "km"
+    return(x)
+  }
+
+  lvls <- levels(factor(pull(to, {{ rm }} )))
+
+  res <- vector("list", length(lvls))
+  names(res) <- lvls
+
+  for (i in seq_along(lvls)) {
+
+    rm_flt <- filter(to, .data[[rm]] == lvls[i])
+
+    res[[lvls[i]]] <- st_distance(from, rm_flt) %>%
+      as_tibble() %>%
+      # first col must contain ids
+      bind_cols(st_drop_geometry(select(from, all_of(id)))) %>%
+      pivot_longer(-id) %>%
+      # to kilometers
+      mutate(value = to_km(value)) %>%
+      group_by(id) %>%
+      summarize(min = min(value),
+                mean = mean(value),
+                median = median(value),
+                max = max(value), .groups = "drop")
+
+    # print(paste(lvls[i], "done,", i, "of", length(lvls)))
+
+  }
+
+  res %>%
+    bind_rows(.id = "rm")
+
+}

### 2021-02-10 R/kdestimate.R

@@ -1,0 +1,81 @@
+#' Kernel density estimation for spatial point pattern
+#'
+#' Wrapper around \code{spatstat} functions. Kernel density is estimated for
+#' a point pattern in a given window for different levels of a categorical
+#' variable.
+#'
+#' @param x \code{sf} object of geometry type \code{POINTS} in projected CRS.
+#' @param mask Single \code{sf} \code{polygon}, the study window.
+#' @param markscol Column name of \code{x} to use as grouping variable.
+#' @param lvls Vector of levels of \code{markscol}.
+#' @param sigma Standard deviation of isotropic smoothing kernel.
+#'     Numeric value, passed to \code{spatstat::density.splitppp} function.
+#'     Scale depends on coordinate reference system, for S-JTSK (EPSG 5514)
+#'     it is in meters and defaults to 4 km.
+#' @param points Logical, return value for points or whole surface as raster image.
+#'     Defaults to \code{TRUE}.
+#'
+#' @return With \code{points} set to \code{TRUE} (default) returns a tibble with
+#'     KDE estimated at given points. If set to \code{FALSE}, a tibble with KDE
+#'     values in a grid is returned to be plotted using \code{ggplot2}.
+#'
+kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
+  stopifnot(any(class(x) == "sf"))
+  stopifnot(all(sf::st_geometry_type(x) == "POINT"))
+  stopifnot(all_of(markscol) %in% colnames(x))
+
+  coords <- x %>% dplyr::filter(!is.na(markscol)) %>%
+    sf::st_coordinates(x)
+  mask <- spatstat::as.owin(mask)
+  marks <- sf::st_drop_geometry(x) %>%
+    dplyr::filter(!is.na(markscol)) %>%
+    dplyr::pull(markscol) %>%
+    factor(levels = lvls)
+
+  pppset <- spatstat::ppp(x = coords[, 1], y = coords[, 2],
+                          window = mask,
+                          marks = marks)
+  splitpppset <- spatstat::split.ppp(pppset)
+
+  if (points) {
+    dens <- spatstat::density.splitppp(splitpppset,
+                                       at = "points",
+                                       sigma = sigma) %>%
+      purrr::map(as_tibble) %>%
[…]

### 2021-02-10 analysis/code/kde.R
add functions code/

@@ -1,122 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# GET AND PLOT KDE VALUES FOR DIFFERENT SETTLEMENTS
-# author: Petr Pajdla
-#
-
-library(tidyverse)
-library(sf)
-library(here)
-library(spatstat)
-
-dt_der <- "analysis/data/derived_data"
-dt_maps <- here(dt_der, "/maps/")
-
-pth_plots <- here("analysis/figures/temp/")
-
-# read data ---------------------------------------------------------------
-# mask
-mask <- st_read(paste0(dt_maps, "mask.geojson"))
-
-# chrono labels
-labs_chrono <- read_rds(here(dt_der, "chrono_labels.RDS"))
-
-# settlements
-set_base <- read_rds(here(dt_der, "settlements.RDS"))
-
-set1<- st_read(here(dt_der, "settlements_sf.geojson")) %>%
-  right_join(set_base$chrono1, by = "rowid")
-
-set2<- st_read(here(dt_der, "settlements_sf.geojson")) %>%
-  right_join(set_base$chrono2, by = "rowid")
-
-# kde ---------------------------------------------------------------------
-
-kdestimate <- function(x, mask, markscol = "label", plot = FALSE) {
-  coords <- st_coordinates(x)
-  mask <- as.owin(mask)
-  marks <- select(st_drop_geometry(x), all_of(markscol))
-
-  pppset <- ppp(x = coords[, 1], y = coords[, 2],
-                window = mask,
-                marks = marks)
-
-  splitpppset <- split.ppp(pppset)
-
-  if (plot) {
-    plot(density.splitppp(splitpppset), main = "")
-  } else {
-    dens <- density.splitppp(splitpppset, at = "points") %>%
-      map(as_tibble) %>%
-      bind_rows(.id = "label") %>%
-      group_by(label) %>%
-      nest()
-
-    set <- x %>%
-      st_drop_geometry() %>%
-      as_tibble() %>%
-      select(rowid, label) %>%
-      group_by(label) %>%
-      nest()
-
-    left_join(set, dens, by = "label") %>%
-      mutate(data = pmap(list(data.x, data.y), bind_cols)) %>%
-      select(label, data) %>%
-      unnest(data) %>%
-      mutate(label = factor(label, levels = levels(marks[, 1]))) %>%
[…]

### 2021-02-10 analysis/code/map_chrono.R

@@ -1,101 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# PLOT SETTLEMENTS
-# author: Petr Pajdla
-# Plot settlement distribution in different periods
-
-library(tidyverse)
-library(sf)
-library(here)
-
-# dt_raw <- "analysis/data/raw_data"
-dt_der <- "analysis/data/derived_data"
-dt_maps <- here(dt_der, "maps/")
-
-pth_plots <- "analysis/figures/temp"
-
-# read data ---------------------------------------------------------------
-# mask
-mask <- st_read(paste0(dt_maps, "mask.geojson"))
-bbox <- st_read(paste0(dt_maps, "bbox.geojson"))
-
-# rivers + admin
-admin0 <- st_read(paste0(dt_maps, "admin0.geojson"))
-admin1 <- st_read(paste0(dt_maps, "admin1.geojson"))
-rivers0 <- st_read(paste0(dt_maps, "rivers0.geojson"))
-rivers1 <- st_read(paste0(dt_maps, "rivers1.geojson"))
-
-# chrono labels
-labs_chrono <- read_rds(here(dt_der, "chrono_labels.RDS"))
-
-set_base <- read_rds(here(dt_der, "settlements.RDS"))
-
-settlements1 <- st_read(here(dt_der, "settlements_sf.geojson")) %>%
-  right_join(set_base$chrono1, by = "rowid")
-
-settlements2 <- st_read(here(dt_der, "settlements_sf.geojson")) %>%
-  right_join(set_base$chrono2, by = "rowid")
-
-# maps - point distribution -----------------------------------------------
-
-# 1
-ch1_n <- set_base$chrono1 %>%
-  group_by(label) %>%
-  summarize(n = n()) %>%
-  mutate(n = str_c("n = ", n))
-
-settlements1 %>%
-  ggplot() +
-  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
-  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
-  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
-  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
[…]

### 2021-02-10 analysis/code/percolation.R

@@ -1,0 +1,344 @@
+# Project "Settlements"
+# Script nr.
+# PERCOLATION ANALYSIS
+# author: Petr Pajdla
+# Find percolation distances
+
+library(tidyverse)
+library(igraph)
+library(sf)
+library(here)
+
+
+# functions ---------------------------------------------------------------
+
+# percolation function
+percolation <- function(x, lower, upper, step) {
+  stopifnot(any(class(x) == "sf"))
+  m <- st_coordinates(x)
+  rownames(m) <- x$id
+
+  dist <- as.matrix(stats::dist(m, method = "euclidean"))
+
+  # radius
+  # triangle <- dist[lower.tri(dist)]
+  # if (is.null(lower)) {
+  #   lower <- min(triangle)
+  # }
+  # if (is.null(upper)) {
+  #   upper <- max(triangle) / 4
+  # }
+  # if (is.null(step)) {
+  #   step <- (upper - lower) / 10
+  # }
+  radii <- seq(lower, upper, by = step)
+
+  res_lst <- vector("list", length(radii))
+  names(res_lst) <- radii
+
+  # distances
+  dist_df <- reshape2::melt(dist)
+  colnames(dist_df) <- c("from", "to", "value")
+  dist_df <- dist_df[!(dist_df$from == dist_df$to), ]
+
+  for (i in seq_along(radii)) {
+    index <- dist_df$value <= radii[i]
+    if (!any(index)) {
+      res_lst[[as.character(radii[i])]] <- matrix(
+        ncol = 2,
+        dimnames = list(NULL, c("id", "clust")))
+    } else {
+      nodes <- dist_df[index, c("from", "to")]
+      nodes <- sapply(nodes, as.character)
+      g <- graph_from_edgelist(nodes)
+
+      # get clusters
+      clust_memb <- clusters(g)$membership
+      # clust_memb <- clusters(g, mode = "weak")$membership
+      memb <- matrix(ncol = 2, nrow = length(clust_memb))
+      colnames(memb) <- c("id", "clust")
+      memb[, "clust"]  <- unname(clust_memb)
+      memb[, "id"] <- names(clust_memb)
+
+      # result - list
+      res_lst[[as.character(radii[i])]] <- memb
+    }
+  }
+
+  res_tbl <- lapply(res_lst, as_tibble) %>%
+    bind_rows(.id = "radius") %>%
+    filter(!is.na(id)) %>%
+    mutate(radius = as.numeric(radius))
+
+  # membership for different radii
+  res <- vector("list", 3)
[…]

### 2021-02-10 analysis/code/rm_distance.R

@@ -1,0 +1,46 @@
+# Project "Settlements"
+# Script nr.
+# DISTANCE TO RAW MATERIAL
+# author: Petr Pajdla
+# Calculated separately for the speed
+
+devtools::load_all(".")
+library(tidyverse)
+library(sf)
+
+
+derived_data <- "analysis/data/derived_data"
+temp_data <- here::here(derived_data, "temp")
+
+# database
+set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
+
+set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+# raw materials
+rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>%
+  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+
+rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>%
+  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+
+# distance
+dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
+
+dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")
+
+# polished stone tools raw materials
+pstrm <- c("AD", "AMFIBOLIT", "KULM", "MTB_JH", "PMD", "ZELESICE")
+
+dist <- bind_rows(p = dist_pts, l = dist_lns, .id = "shp") %>%
+  select(id, everything()) %>%
+  mutate(across(where(is.numeric), unclass),
+         type = if_else(rm %in% pstrm, "Polished", "Chipped"),
+         region = if_else(str_detect(id, "^B"), "Boh.", "Mor."))
+
+# save result
+dir.create(temp_data)
+
+write_csv(dist, paste0(temp_data, "rm_dist.csv"))
+

### 2021-02-11 R/kdestimate.R
add column chrono to kde

@@ -49,7 +49,7 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
     set <- x %>%
       sf::st_drop_geometry() %>%
       dplyr::as_tibble() %>%
-      dplyr::select(id, markscol) %>%
+      dplyr::select(id, chrono, markscol) %>%
       dplyr::group_by(!!as.symbol(markscol)) %>%
       tidyr::nest()
 
@@ -61,7 +61,7 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
       dplyr::mutate(name = factor(!!as.symbol(markscol),
                                   levels = levels(marks))) %>%
       dplyr::mutate(kde = value * 1e6) %>%
-      dplyr::select(id, name, kde)
+      dplyr::select(id, name, chrono, kde)
   } else {
     img <- spatstat::density.splitppp(splitpppset,
                                       sigma = sigma)

### 2021-02-11 analysis/paper/paper.Rmd
cont watercourse density

@@
 # abstract: |
-#  "This is abstract"
 keywords: |
   keyword 1; keyword 2; keyword 3
+editor_options: 
+  markdown: 
+    wrap: 80
 ---
@@
 devtools::load_all(".")
+library(sf)
 library(tidyverse)
-library(sf)
 ```
@@
 
-* Pottery traditions
-* Pottery groups
+-   Pottery traditions
+-   Pottery groups
 
 The study area is broadly defined as *Easter part of Bohemia* (abbreviated *B*)
-and *Morava river catchment* (*M*).
-It spans across the area of the Czech Republic and bordering regions of 
-Austria and Slovakia.
+and *Morava river catchment* (*M*). It spans across the area of the Czech
+Republic and bordering regions of Austria and Slovakia.
 
@@
 
-# read data ---------------------------------------------------------------
 # mask
@@
 ```
-
-<!--
-```{r base-tab}
-# set_base$base %>%
-#   DT::datatable(rownames = FALSE)
-```
--->
 
@@
 
-Settlement density is estimated using KDE.
-Kernel size of 4 km is used. 
-This value is completely arbitrary but is selected because settlement clusters 
-are nicely highlighted at this scale.
+Settlement density is estimated using KDE. Kernel size of 4 km is used. This
+value is completely arbitrary but is selected because settlement clusters are
+nicely highlighted at this scale.
 
@@
          period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  select(id, starts_with("period"), kde)
+  select(id, starts_with("period"), chrono, kde)
+
+# tibble with results
+res1_long <- kde1 %>% 
+  rename(settlements_kde = kde) %>% 
+  select(-period_label) %>% 
+  pivot_longer(col = settlements_kde, names_to = "variable")
 ```
@@
          period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  select(id, starts_with("period"), kde)
+  select(id, starts_with("period"), chrono, kde)
+
+# tibble with results
+res2_long <- kde2 %>% 
+  rename(settlements_kde = kde) %>% 
+  select(-period_label) %>% 
+  pivot_longer(col = settlements_kde, names_to = "variable")
 ```
@@
 
-Raw material sources are defined as either points or lines.
-Each raw material can be identified by single or multiple simple features.
-If the location of source used in the Neolithic period is unknown or several sources or even
-modes of procurement (e. g. from river bed etc.) are possible, the most probable 
-are included.
+Raw material sources are defined as either points or lines. Each raw material
+can be identified by single or multiple simple features. If the location of
+source used in the Neolithic period is unknown or several sources or even modes
+of procurement (e. g. from river bed etc.) are possible, the most probable are
+included.
+
+Minimum, mean, median and maximum distances are calculated, median is used 
+as a result.
 
@@
 ```
+
 ```{r rm-dist, cache = FALSE}
@@
 rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))
+
+rm_dist_median <- rm_dist %>% 
+  select(-region, -type, -shp, -min, -mean, -max) %>% 
+  group_by(id, rm) %>% 
+  summarise(dist_rm = mean(median)) %>% 
+  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
+  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
+  select(-rm)
+
+# update result tibbles
+res1_long <- set_base$period1 %>% 
+  select(id, period, chrono) %>% 
+  inner_join(rm_dist_median) %>%
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(id, period, chrono) %>% 
+  inner_join(rm_dist_median) %>%
+  bind_rows(res2_long)
 ```
@@
 
-<!--\@ref(fig:demo-plot)-->
+# Density of watercourses
+
+In settlement structure studies it is often assumed that past human settlements
+have some kind of a relation to the network of watercourses. Numerous problems
+are faced while assessing the relationship between settlements and close
+watercourse. Most importantly the man-made recent changes in the structure and
+intensity of watercourses (and landscape in general) are a major factor
+imapiring such analysis.
+
+Usually, the settlement - watercourse relationship is expressed in some form of
+a metric giving shortest distance or so. We choose an approach used to study
+road networks in modern urban areas [e. g. @lin2020].
+
+Most of the input data is derived from *DIBAVOD* data set A01 layer
+(<https://www.dibavod.cz/>). 
+Data for Slovakia are derived from OpenStreet map ? , data for Austria ???
+
+Density (kernel smoothed intensity) is estimated using `spatstat::density.psp`
+function with sigma 2000 m.
+
+```{r get-raster}
+hydro_dens <- raster::raster(
+  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
+  raster::rasterToPoints() %>%
+  as_tibble() %>%
+  rename(KDE = density_raster)
+```
+
+```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
+ggplot() +
+  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
+  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
+  theme_void()
+```
+
+# Results
+
+```{r results-export}
+res1_long %>% 
+  pivot_wider(names_from = variable, values_from = value) %>% 
+  arrange(id, period, chrono) %>% 
+  write_csv(here::here(derived_data, "results", "results1.csv"))
+
+res2_long %>% 
+  pivot_wider(names_from = variable, values_from = value) %>% 
+  arrange(id, period, chrono) %>% 
+  write_csv(here::here(derived_data, "results", "results2.csv"))
+```
 
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
@@
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+This report was generated on `r Sys.time()` using the following computational
+environment and dependencies:
 

### 2021-02-11 analysis/code/dataprep_hydro.R

@@ -1,7 +1,14 @@
+# Project "Settlements"
+# Script nr.
+# WATERCOURSE DENSITY
+# author: Petr Pajdla
+# Joins neccessary layers and calculates watercourse intensity (KDE)
 
 library(sf)
 library(here)
 library(tidyverse)
+library(spatstat)
+library(raster)
 
 # read data
 dt_hydro <- "analysis/data/raw_data/hydro/"
@@ -14,7 +21,7 @@ au <- st_read(here(dt_hydro, "hydro_austria"))
 sk <- st_read(here(dt_hydro, "hydro_slovakia"))
 
 # project mask
-mask <- st_read(here(here("analysis/data/derived_data/maps/", "mask.geojson")))
+mask <- st_read(here("analysis/data/derived_data/maps/", "mask.geojson"))
 
 # transform to S-JTSK and crop to mask
 au_crop <- au %>%
@@ -29,6 +36,8 @@ sk_crop <- sk %>%
 cz_crop <- cz %>%
   st_intersection(mask)
 
+rm(list = c("cz", "au", "sk"))
+
 # control plot
 ggplot() +
   geom_sf(data = au_crop, color = "green") +
@@ -36,14 +45,55 @@ ggplot() +
   geom_sf(data = sk_crop, color = "orange") +
   theme_void()
 
-# join layers
-temp <- st_join(st_as_sf(st_geometry(au_crop)),
-                st_as_sf(st_geometry(sk_crop)))
-x <- st_join(temp, st_as_sf(st_geometry(cz_crop)))
+# join layers?
+sk_geom <- st_combine(st_geometry(sk_crop))
+au_geom <- st_combine(st_geometry(au_crop))
+cz_geom <- st_combine(st_geometry(cz_crop))
 
+rm(list = c("cz_crop", "au_crop", "sk_crop"))
+
+# save temp files...
+# dt_hydro_derived <- "analysis/data/derived_data/hydro/"
+# dir.create(here(dt_hydro_derived))
+# st_write(sk_geom, here(dt_hydro_derived, "sk_prep.geojson"))
+# st_write(au_geom, here(dt_hydro_derived, "au_prep.geojson"))
+# st_write(cz_geom, here(dt_hydro_derived, "cz_prep.geojson"))
+#
+# rstudioapi::restartSession()
+#
+# sk_geom <- st_read(here(dt_hydro_derived, "sk_prep.geojson"))
[…]

### 2021-02-11 analysis/code/results_to_gd.R
script to push results to gd

@@ -1,0 +1,38 @@
+# Project "Settlements"
+# Script nr. +Inf
+# PUBLISH RESULTS DATA TO GD
+# author: Petr Pajdla
+# Pushes results csvs to GD
+
+library(here)
+library(tidyverse)
+library(googledrive)
+
+
+# upload results to GD -----------------------------------------------------
+
+# paths
+dt_result <- "analysis/data/derived_data/results"
+gd_path <- "~/settlements/data/results/"
+
+now <- paste("Updated", Sys.time())
+write_lines(now, here(dt_result, "README"), append = TRUE)
+
+# delete previous files
+drive_trash(filter(drive_ls(gd_path), str_detect(name, "^results.$")))
+
+# list files to upload
+files_local <- list.files(dt_result, pattern = "csv$", full.names = TRUE)
+readme <- list.files(dt_result, pattern = "^README$", full.names = TRUE)
+
+drive_results <- map(files_local, ~ drive_upload(.x, gd_path,
+                                           type = "spreadsheet",
+                                           overwrite = TRUE))
+drive_readme <- drive_upload(readme, gd_path, overwrite = TRUE)
+
+
+# upload SI to GD ---------------------------------------------------------
+
+si <- here("analysis/paper/paper.html")
+
+drive_si <- drive_upload(si, "~/settlements/SI/", overwrite = TRUE)

### 2021-04-13 R/kdestimate.R
update broken dependencies - spatstat split into child packages

@@ -26,19 +26,19 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
 
   coords <- x %>% dplyr::filter(!is.na(markscol)) %>%
     sf::st_coordinates(x)
-  mask <- spatstat::as.owin(mask)
+  mask <- spatstat.geom::as.owin(mask)
   marks <- sf::st_drop_geometry(x) %>%
     dplyr::filter(!is.na(markscol)) %>%
     dplyr::pull(markscol) %>%
     factor(levels = lvls)
 
-  pppset <- spatstat::ppp(x = coords[, 1], y = coords[, 2],
+  pppset <- spatstat.geom::ppp(x = coords[, 1], y = coords[, 2],
                           window = mask,
                           marks = marks)
-  splitpppset <- spatstat::split.ppp(pppset)
+  splitpppset <- spatstat.geom::split.ppp(pppset)
 
   if (points) {
-    dens <- spatstat::density.splitppp(splitpppset,
+    dens <- spatstat.core::density.splitppp(splitpppset,
                                        at = "points",
                                        sigma = sigma) %>%
       purrr::map(as_tibble) %>%
@@ -63,7 +63,7 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
       dplyr::mutate(kde = value * 1e6) %>%
       dplyr::select(id, name, chrono, kde)
   } else {
-    img <- spatstat::density.splitppp(splitpppset,
+    img <- spatstat.core::density.splitppp(splitpppset,
                                       sigma = sigma)
     rstr <- vector("list", length(img))
     names(rstr) <- lvls

### 2021-04-13 analysis/code/dataprep.R
update datapreparation pipelines

@@ -33,15 +33,20 @@ files_disk <- list.files(here(dt_der, "input_gd"), full.names = TRUE)
 
 set_base <- vector("list")
 
-set_base$base <- read_csv(str_subset(files_disk, "base"))
+set_base$base <- read_csv(str_subset(files_disk, "base")) %>%
+  filter(in_analysis)
 
-set_base$chrono1 <- read_csv(str_subset(files_disk, "chrono1"))
+set_base$chrono1 <- read_csv(str_subset(files_disk, "chrono1")) %>%
+  filter(id %in% set_base$base$id)
 
-set_base$chrono2 <- read_csv(str_subset(files_disk, "chrono2"))
+set_base$chrono2 <- read_csv(str_subset(files_disk, "chrono2")) %>%
+  filter(id %in% set_base$base$id)
 
-set_base$period1 <- read_csv(str_subset(files_disk, "period1"))
+set_base$period1 <- read_csv(str_subset(files_disk, "period1")) %>%
+  filter(id %in% set_base$base$id)
 
-set_base$period2 <- read_csv(str_subset(files_disk, "period2"))
+set_base$period2 <- read_csv(str_subset(files_disk, "period2")) %>%
+  filter(id %in% set_base$base$id)
 
 read_csv(str_subset(files_disk, "spatial")) %>%
   st_as_sf(coords = c("X", "Y")) %>%

### 2021-04-13 analysis/code/dataprep_hydro.R

@@ -21,7 +21,12 @@ au <- st_read(here(dt_hydro, "hydro_austria"))
 sk <- st_read(here(dt_hydro, "hydro_slovakia"))
 
 # project mask
-mask <- st_read(here("analysis/data/derived_data/maps/", "mask.geojson"))
+mask_orig <- st_read(here("analysis/data/derived_data/maps/", "mask.geojson"))
+
+mask <- st_buffer(mask_orig, 20000) %>% st_simplify()
+
+plot(st_geometry(mask))
+plot(st_geometry(mask_orig), add = TRUE)
 
 # transform to S-JTSK and crop to mask
 au_crop <- au %>%
@@ -75,7 +80,7 @@ full <- st_union(temp, cz_geom)
 # full <- st_read(here(dt_hydro_derived, "hydro_mask.geojson"))
 
 rm(list = c("cz_geom", "au_geom", "sk_geom", "temp"))
-rstudioapi::restartSession()
+# rstudioapi::restartSession()
 
 ggplot() +
   geom_sf(data = full) +
@@ -96,4 +101,5 @@ dens_raster <- dens_raster %>%
   mask(mask)
 
 # save resulting raster
-dens_raster %>% writeRaster(here(dt_hydro_derived, "density_raster.tif"))
+dens_raster %>% writeRaster(here(dt_hydro_derived, "density_raster.tif"),
+                            overwrite = TRUE)

### 2021-04-13 analysis/code/rm_distance.R

@@ -40,7 +40,11 @@ dist <- bind_rows(p = dist_pts, l = dist_lns, .id = "shp") %>%
          region = if_else(str_detect(id, "^B"), "Boh.", "Mor."))
 
 # save result
-dir.create(temp_data)
 
-write_csv(dist, paste0(temp_data, "rm_dist.csv"))
+if (!dir.exists(temp_data)) {
+  dir.create(temp_data)
+}
+
+
+write_csv(dist, paste0(temp_data, "/rm_dist.csv"))
 

### 2021-04-13 analysis/paper/paper.Rmd
rm dist correlation with kde

@@
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      toc: true
+      toc_depth: 3
       pandoc_args:
@@
 knitr::opts_chunk$set(
-  collapse = TRUE,
+  collapse = FALSE,
   warning = FALSE,
@@
 
-```{r packages}
+```{r packages, cache=FALSE}
 devtools::load_all(".")
@@
 library(tidyverse)
+library(spatstat)
 ```
@@
 rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
-  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+  mutate(label = if_else(orig == "l", "Chipped", "Polished"),
+         rm = str_to_lower(rm))
 rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
-  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+  mutate(label = if_else(orig == "l", "Chipped", "Polished"),
+         rm = str_to_lower(rm))
 ```
@@
 ```
+## Distances
 
@@
 
-rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))
+rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
+  mutate(rm = str_to_lower(rm))
 
-rm_dist_median <- rm_dist %>% 
-  select(-region, -type, -shp, -min, -mean, -max) %>% 
+# rm_dist_median <- rm_dist %>% 
+#   select(-region, -type, -shp, -min, -mean, -max) %>% 
+#   group_by(id, rm) %>% 
+#   summarise(dist_rm = mean(median)) %>% 
+#   pivot_longer(cols = dist_rm, names_to = "variable") %>% 
+#   mutate(variable = str_c(variable, rm, sep = "_")) %>% 
+#   select(-rm)
+
+rm_dist_mean <- rm_dist %>%
+  select(-region, -type, -shp, -min, -median, -max) %>%
   group_by(id, rm) %>%
-  summarise(dist_rm = mean(median)) %>% 
+  summarise(dist_rm = mean(mean)) %>%
   pivot_longer(cols = dist_rm, names_to = "variable") %>%
@@
   select(-rm)
-
-# update result tibbles
-res1_long <- set_base$period1 %>% 
-  select(id, period, chrono) %>% 
-  inner_join(rm_dist_median) %>%
-  bind_rows(res1_long)
-
-res2_long <- set_base$period2 %>% 
-  select(id, period, chrono) %>% 
-  inner_join(rm_dist_median) %>%
-  bind_rows(res2_long)
 ```
 
-```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
+```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."}
 rm_dist %>%
@@
 ```
+## Settlement density against distance to raw materials
+
+> Hypothesis: In Morava river catchment settlement position is directly associated with 
+  raw material source.
+
+We compare settlement density (KDE) with distance to chosen raw materials.
+There is no strong evidence that the distribution of settlements is associated with distance to 
+any of the key raw material sources occurring in the region in any of the studied periods.
+There are some mild negative correlations in the period 4.8 -- 4.2 k BCE for polished stone tools materials.
+Strongest connection is observed between given pottery groups in period 4.8 -- 4.6 k BCE in case of ad/pmd/zelesice raw material sources.
 
@@
   mutate(region = str_extract(id, "^.")) %>% 
-  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
+  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
          region == "M") %>% 
-  ggplot(aes(median, kde)) +
+  ggplot(aes(mean, kde)) +
   geom_point(alpha = 0.1) + 
@@
   facet_grid(vars(period_label), vars(rm)) +
-  labs(x = "Median distance to raw material source (km)",
+  labs(x = "Mean distance to raw material source (km)",
        y = "KDE") +
@@
   theme(strip.text.y = element_text(angle = 0))
+```
+
+```{r tab-rm-dist-cor-kde1}
+kde1 %>% select(id, period_label, kde) %>% 
+  left_join(rm_dist) %>% 
+  mutate(region = str_extract(id, "^.")) %>% 
+  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
+         region == "M") %>% 
+  group_by(period_label, rm) %>% 
+  summarise(cor = cor(min, kde)) %>% 
+  pivot_wider(id_cols = period_label, names_from = rm, values_from = cor) %>% 
+  rename(period = period_label) %>% 
+  knitr::kable(caption = "Table of correlations between minimal distance to raw material and settlement density, pottery traditions.") %>% 
+  kableExtra::kable_styling(full_width = FALSE)
 ```
@@
   theme(strip.text.y = element_text(angle = 0))
+```
+```{r tab-rm-dist-cor-kde2}
+kde2 %>% select(id, period_label, kde) %>% 
+  left_join(rm_dist) %>% 
+  mutate(region = str_extract(id, "^.")) %>% 
+  filter(rm %in% c("ad", "zelesice", "pmd", "kl"),
+         region == "M") %>% 
+  group_by(period_label, rm) %>% 
+  summarise(cor = cor(min, kde)) %>% 
+  pivot_wider(id_cols = period_label, names_from = rm, values_from = cor) %>% 
+  rename(period = period_label) %>% 
+  knitr::kable(caption = "Table of correlations between minimal distance to raw material and settlement density, pottery groups.") %>% 
+  kableExtra::kable_styling(full_width = FALSE)
+```
+
+## PCA for raw material distances
+
+Principal components analysis is used to reduce dimensionality.
+First three PCs are enough to cover more than 95% variability in the data set.
+
+```{r rm-dist-pca}
+scale_col <- function(x){
+  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
+}
+
+rm_dist_mean_wide <- rm_dist_mean %>%
+  mutate(variable = str_remove(variable, "dist_rm_")) %>% 
+  pivot_wider(names_from = variable, values_from = value) %>% 
+  ungroup() %>% 
+  mutate(across(where(is.numeric), scale_col))
+
+# matrix for pca to reduce dimensions
+rm_dist_mx <- rm_dist_mean_wide %>% 
+  select(-id) %>% 
+  as.matrix() 
+row.names(rm_dist_mx) <- rm_dist_mean_wide$id
+
+rm_pca <- prcomp(rm_dist_mx)
+summary(rm_pca)
+```
+```{r rm-corrplot, fig.cap="Correlation between principal components and given raw materials."}
+corrplot::corrplot(t(rm_pca$rotation[, 1:3]))
+```
+```{r rm-biplot, fig.cap="First thee principal components (green - East part of Bohemia, blue - Morava river catchment)."}
+mycol <- c("B" = "#7FC97F", "M" = "#BEAED4")
+pairs(rm_pca$x[, 1:3], col = unname(mycol[str_extract(rownames(rm_dist_mx), ".")]))
+```
+
+```{r update-result-rm}
+rm_dist_pca <- rm_pca$x[, 1:3] %>% 
+  as_tibble(rownames = "id") %>% 
+  rename("rm_pc1" = "PC1",
+         "rm_pc2" = "PC2",
+         "rm_pc3" = "PC3",) %>% 
+  pivot_longer(-id, "variable", "value")
+
+# update result tibbles
+res1_long <- set_base$period1 %>% 
+  select(id, period, chrono) %>% 
+  inner_join(rm_dist_pca) %>%
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(id, period, chrono) %>% 
+  inner_join(rm_dist_pca) %>%
+  bind_rows(res2_long)
+
+rm(list = ls(pattern = "^rm_dist_"))
 ```
@@
 (<https://www.dibavod.cz/>).
-Data for Slovakia are derived from OpenStreet map ? , data for Austria ???
 
@@
 hydro_dens <- raster::raster(
-  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
+  here::here(derived_data, "hydro", "density_raster.tif"))
+
+hydro_dens_at_points <- hydro_dens %>% 
+  raster::extract(set_spat) %>% 
+  as_tibble() %>% 
+  rename(hydro_kde = value) %>% 
+  bind_cols(select(st_drop_geometry(set_spat), id))
+
+res1_long <- set_base$period1 %>% 
+  select(-ends_with("label")) %>% 
+  left_join(hydro_dens_at_points) %>% 
+  pivot_longer(cols = hydro_kde) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-ends_with("label"), -facet) %>% 
+  left_join(hydro_dens_at_points) %>% 
+  pivot_longer(cols = hydro_kde) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res2_long)
+```
+
+```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites (raster clipped to research area extent)."}
+hydro_dens %>% 
+  raster::mask(mask) %>% 
   raster::rasterToPoints() %>%
   as_tibble() %>%
-  rename(KDE = density_raster)
-```
-
-```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
+  rename(KDE = density_raster) %>% 
   ggplot() +
-  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
+  geom_raster(aes(x, y, fill = KDE)) +
   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
@@
 ```
+
+# Terrain surrounding the settlements
+
+We focus on slope around the settlements
+
+# Synthesis
 
@@
 ```
-
-<!-- The following line inserts a page break -->
 

### 2021-04-14 analysis/code/dataprep.R
add references to sites

@@ -26,26 +26,34 @@ for (i in seq_along(files_gd$id)) {
 
 
 # prepare derived data ----------------------------------------------------
-
+labs_chrono <- read_rds(here::here(dt_der, "chrono_labels.RDS"))
 
 # list downloaded files
 files_disk <- list.files(here(dt_der, "input_gd"), full.names = TRUE)
 
+# prepare input files for analysis
 set_base <- vector("list")
 
 set_base$base <- read_csv(str_subset(files_disk, "base")) %>%
   filter(in_analysis)
 
-set_base$chrono1 <- read_csv(str_subset(files_disk, "chrono1")) %>%
-  filter(id %in% set_base$base$id)
-
-set_base$chrono2 <- read_csv(str_subset(files_disk, "chrono2")) %>%
-  filter(id %in% set_base$base$id)
+# set_base$chrono1 <- read_csv(str_subset(files_disk, "chrono1")) %>%
+#   filter(id %in% set_base$base$id)
+#
+# set_base$chrono2 <- read_csv(str_subset(files_disk, "chrono2")) %>%
+#   filter(id %in% set_base$base$id)
 
 set_base$period1 <- read_csv(str_subset(files_disk, "period1")) %>%
-  filter(id %in% set_base$base$id)
+  filter(id %in% set_base$base$id) %>%
+  mutate(period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
+         period = fct_relevel(period, names(labs_chrono$periods)))
 
 set_base$period2 <- read_csv(str_subset(files_disk, "period2")) %>%
+  filter(id %in% set_base$base$id) %>%
+  mutate(period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
+         period = fct_relevel(period, names(labs_chrono$periods)))
+
+set_base$references <- read_csv(str_subset(files_disk, "references")) %>%
   filter(id %in% set_base$base$id)
 
 read_csv(str_subset(files_disk, "spatial")) %>%

### 2021-04-14 analysis/code/dataprep_references.R

@@ -1,0 +1,117 @@
+# Project "Settlements"
+# Script nr. ?
+# PREPARE REFERENCES
+# author: Petr Pajdla
+# Corrected references are added to base database
+
+library(tidyverse)
+library(here)
+library(httr)
+
+# References
+refs <- readxl::read_excel(here("analysis/data/raw_data", "references.xlsx")) %>%
+  select(id, references) %>%
+  separate(references, into = str_c("r", 1:15), sep = ";") %>%
+  pivot_longer(-id) %>%
+  filter(!is.na(value)) %>%
+  mutate(value = str_trim(value)) %>%
+  select(id, reference = value)
+
+# PIAN to Akce
+pian <- refs %>%
+  mutate(pian = if_else(str_detect(reference, "^P-\\d+"), TRUE, FALSE)) %>%
+  filter(pian) %>%
+  select(-pian) %>%
+  mutate(pian1 = str_extract(reference, "^P-\\d+"),
+         pian2 = str_extract(reference, "\\d{5}$"),
+         pian = str_c(pian1, "-0", pian2)) %>%
+  select(-pian1, -pian2)
+
+# ask for password somwhere
+password <- rstudioapi::askForPassword()
+
+
+# query API ---------------------------------------------------------------
+
+#'@param child Child to select from the XML tree
+query_amcr_api <- function(amcr_ident, username, child) {
+
+  api_url <- "https://api.aiscr.cz/dapro/"
+  verb <- "oai?verb=GetRecord"
+  ident <- paste0("&identifier=https://api.aiscr.cz/id/", amcr_ident)
+  prefix <- "&metadataPrefix=oai_amcr"
+
+  resp <- GET(paste0(api_url, verb, ident, prefix),
+              authenticate(username, password))
+
+  res <- content(resp, "parsed", encoding = "UTF-8") %>%
+    rvest::html_elements(xpath = child) %>%
+    rvest::html_text()
+
+  return(res)
+}
+
+#'@param x vector of AMCR identifiers
+query_amcr_api_loop <- function(x, username, child) {
+
+  res <- vector("list", length(x))
+  names(res) <- x
+
+  for (i in seq_along(x)) {
+    Sys.sleep(0.1)
+    res[[i]] <- query_amcr_api(x[i], username, child)
+    print(i)
+  }
+
+  return(res)
+
+}
+
+
+# Queries -----------------------------------------------------------------
+
+# get dokumentační jednotky
[…]

### 2021-05-28 analysis/paper/paper.Rmd
finish dataprep, cont. analysis

@@
   echo = FALSE,
-  cache = TRUE,
+  cache = FALSE,
   comment = "#>",
@@
 
-```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."}
-rm_dist %>%
-  ggplot(aes(x = min, y = rm, fill = region)) +
-  geom_violin() +
-  scale_fill_brewer(palette = "Accent") +
-  facet_wrap(~type, scales = "free_y") +
-  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
+<!-- ```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."} -->
+<!-- rm_dist %>% -->
+<!--   ggplot(aes(x = min, y = rm, fill = region)) + -->
+<!--   geom_violin() + -->
+<!--   scale_fill_brewer(palette = "Accent") + -->
+<!--   facet_wrap(~type, scales = "free_y") + -->
+<!--   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") + -->
+<!--   theme_minimal() + -->
+<!--   theme(legend.position = "bottom",  -->
+<!--         panel.border = element_rect(color = "black", fill = NA)) -->
+<!-- ``` -->
+
 ## Settlement density against distance to raw materials
@@
   theme_minimal() +
-  theme(strip.text.y = element_text(angle = 0))
+  theme(strip.text.y = element_text(angle = 0),
+        panel.border = element_rect(color = "black", fill = NA))
 ```
@@
 
+## RM relevant to given period and/or region
+
+```{r rm-period}
+# 0 = not used
+# 0.5 = somewhat used
+# 1 = used
+rm_b <- readxl::read_xlsx(here::here("analysis/data/raw_data",
+                                     "sidliste_vs_zdroje.xlsx"), sheet = "B")
+rm_m <- readxl::read_xlsx(here::here("analysis/data/raw_data",
+                                     "sidliste_vs_zdroje.xlsx"), sheet = "M")
+
+rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
+  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
+  pivot_longer(starts_with("p"), names_to = "period", values_to = "relevance") %>% 
+  rename(rm = surovina)
+
+rm_dist_rel_p1 <- rm_dist %>% group_by(id, rm) %>% 
+  summarise(mean = mean(mean)) %>% 
+  mutate(reg = if_else(str_detect(id, "^B"), "b", "m")) %>% 
+  left_join(select(set_base$period1, period, id)) %>% 
+  left_join(rm_relevance) %>% 
+  filter(relevance > 0) %>% 
+  mutate(relevance = factor(relevance, levels = c("0", "0.5", "1")),
+         label = labs_chrono$periods[period],
+         label = factor(label, levels = labs_chrono$periods))
+
+rm_dist_rel_p2 <- rm_dist %>% group_by(id, rm) %>% 
+  summarise(mean = mean(mean)) %>% 
+  mutate(reg = if_else(str_detect(id, "^B"), "b", "m")) %>% 
+  left_join(select(set_base$period2, period, id)) %>% 
+  left_join(rm_relevance) %>% 
+  filter(relevance > 0) %>% 
+  mutate(relevance = factor(relevance, levels = c("0", "0.5", "1")),
+         label = labs_chrono$periods[period],
+         label = factor(label, levels = labs_chrono$periods))
+```
+
+```{r fun-plot-rm-reg-dist}
+plot_rm_reg_dist <- function(x, region) {
+  x %>% 
+    filter(reg == region) %>% 
+    ggplot(aes(label, mean, color = relevance)) +
+    geom_violin(show.legend = FALSE) +
+    geom_boxplot(width = 0.2, show.legend = FALSE) +
+    scale_color_manual(values = c("gray", "black")) +
+    facet_wrap(vars(rm), scales = "free_y") + 
+    labs(x = element_blank(), y = "Mean distance to raw material source (km)") +
+    theme_minimal() + 
+    theme(axis.text.x = element_text(angle = 80, hjust = 1))
+}
+```
+
+### RM relations in *general* pottery traditions
+
+#### Morava river catchment
+
+```{r rm-dist-m-p1, fig.cap="Distance to RM sources, Morava river catchment, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p1 %>% 
+  plot_rm_reg_dist("m")
+```
+
+#### Eastern part of Bohemia
+
+```{r rm-dist-b-p1, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p1 %>% 
+  plot_rm_reg_dist("b")
+```
+
+### RM relations in *detailed* pottery groups
+
+#### Morava river catchment
+
+```{r rm-dist-m-p2, fig.cap="Distance to RM sources, Morava river catchment, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p2 %>% 
+  plot_rm_reg_dist("m")
+```
+
+#### Eastern part of Bohemia
+
+```{r rm-dist-b-p2, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p2 %>% 
+  plot_rm_reg_dist("b")
+```
+
 <!-- ======================================================================= -->
@@
 ```{r get-raster}
-hydro_dens <- raster::raster(
-  here::here(derived_data, "hydro", "density_raster.tif"))
+# pkg raster
+# hydro_dens <- raster::raster(
+#   here::here(derived_data, "hydro", "density_raster.tif"))
+# 
+# hydro_dens_at_points <- hydro_dens %>% 
+#   raster::extract(set_spat) %>% 
+#   as_tibble() %>% 
+#   rename(hydro_kde = value) %>% 
+#   bind_cols(select(st_drop_geometry(set_spat), id))
+
+# pkg stars
+hydro_dens <- stars::read_stars(
+  here::here(derived_data, "hydro", "density_raster.tif")
+) %>% 
+  st_set_crs(value = st_crs(set_spat))
 
 hydro_dens_at_points <- hydro_dens %>% 
-  raster::extract(set_spat) %>% 
-  as_tibble() %>% 
-  rename(hydro_kde = value) %>% 
+  stars::st_extract(set_spat) %>% 
+  st_drop_geometry() %>% 
+  rename(hydro_kde = density_raster.tif) %>% 
   bind_cols(select(st_drop_geometry(set_spat), id))
 
+# bind to results
 res1_long <- set_base$period1 %>% 
   select(-ends_with("label")) %>% 
-  left_join(hydro_dens_at_points) %>% 
+  left_join(hydro_dens_at_points, by = "id") %>% 
   pivot_longer(cols = hydro_kde) %>% 
@@
   select(-ends_with("label"), -facet) %>% 
-  left_join(hydro_dens_at_points) %>% 
+  left_join(hydro_dens_at_points, by = "id") %>% 
   pivot_longer(cols = hydro_kde) %>% 
@@
 ```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites (raster clipped to research area extent)."}
-hydro_dens %>% 
-  raster::mask(mask) %>% 
-  raster::rasterToPoints() %>%
-  as_tibble() %>%
-  rename(KDE = density_raster) %>% 
+# hydro_dens %>% 
+#   raster::mask(mask) %>% 
+#   raster::rasterToPoints() %>%
+#   as_tibble() %>%
+#   rename(KDE = density_raster) %>% 
+#   ggplot() +
+#   geom_raster(aes(x, y, fill = KDE)) +
+#   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
+#   theme_void()
+
 ggplot() +
-  geom_raster(aes(x, y, fill = KDE)) +
+  stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
-  theme_void()
+  theme_void() +
+  labs(fill = "KDE")
 ```
+<!--- ====================================================================== -->
 
@@
 
-We focus on slope around the settlements
+We focus on slope around the settlements.
+
+```{r dem-data}
+dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))
+
+# bind to results
+res1_long <- set_base$period1 %>% 
+  select(-ends_with("label")) %>% 
+  left_join(dem_data, by = "id") %>% 
+  pivot_longer(cols = c(dem, slope)) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-ends_with("label"), -facet) %>% 
+  left_join(dem_data, by = "id") %>% 
+  pivot_longer(cols = c(dem, slope)) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res2_long)
+```
+
+```{r altitude-to-res}
+altitude <- set_spat %>% st_drop_geometry() %>% 
+  as_tibble() %>% 
+  select(id, altitude)
+
+# bind to results
+res1_long <- set_base$period1 %>% 
+  select(-ends_with("label")) %>% 
+  left_join(altitude, by = "id") %>% 
+  pivot_longer(cols = altitude) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-ends_with("label"), -facet) %>% 
+  left_join(altitude, by = "id") %>% 
+  pivot_longer(cols = altitude) %>% 
+  rename(variable = name) %>% 
+  bind_rows(res2_long)
+```
+
+```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
+set_spat %>% st_drop_geometry() %>% 
+  as_tibble() %>% 
+  select(id, altitude) %>% 
+  left_join(dem_data) %>% 
+  ggplot(aes(altitude, dem)) +
+  geom_point() +
+  theme_minimal()
+```
+
+<!--- ====================================================================== -->
+
+# Settlement 'hierarchy'
+
+Minimal distance to fortified settlements in given regions is explored.
+
+```{r dist-fenced-input}
+dist_fenced <- read_rds(here::here(derived_data, "fortif/distance_fortif.RDS"))
+
+# results 1
+dist_fenced_p1 <- dist_fenced$traditions %>% 
+  mutate(variable = "dist_fenced",
+         value = min_dist) %>% 
+  select(-min_dist, -reg) %>% 
+  left_join(select(set_base$period1, id, period, chrono))
+
+res1_long <- res1_long %>% 
+  bind_rows(dist_fenced_p1)
+
+# results 2
+dist_fenced_p2 <- dist_fenced$groups %>% 
+  mutate(variable = "dist_fenced",
+         value = min_dist) %>% 
+  select(-min_dist, -reg)
+
+res2_long <- res2_long %>% 
+  bind_rows(dist_fenced_p2)
+```
+
+<!--- ====================================================================== -->
+
+# Settlement 'continuity'
+
+Is the same location settled in the previous period?
+
+```{r continuity-calc}
+settlement_continuity <- function(x) {
+  x %>% select(-label, -period_label) %>%
+    distinct(id, period, chrono) %>%
+    mutate(pres = 1L) %>%
+    pivot_wider(names_from = period, values_from = pres, values_fill = 0L) %>%
+    pivot_longer(starts_with("p"), names_to = "period") %>%
+    mutate(period = factor(period, levels = levels(set_base$period1$period)),
+           period_label = labs_chrono$periods[period],
+           period_label = factor(period_label, labs_chrono$periods)) %>%
+    arrange(id, period, chrono) %>%
+    group_by(id, chrono) %>%
+    mutate(prev = if_else(value == 1, lag(value), 0L),
+           prev = if_else(is.na(prev), 0L, prev),
+           reg = if_else(str_detect(id, "^B"), "B", "M")) %>%
+    ungroup(id, chrono)
+}
+
+cont1 <- set_base$period1 %>%
+  settlement_continuity()
+
+cont2 <- set_base$period2 %>%
+  settlement_continuity()
+
+# update results
+cont1_res <- cont1 %>% select(-value, -reg, -period_label) %>% 
+  mutate(variable = "cont") %>% 
+  rename(value = prev) %>% 
+  filter(value != 0)
+
+res1_long <- res1_long %>% bind_rows(cont1_res)
+
+cont2_res <- cont2 %>% select(-value, -reg, -period_label) %>% 
+  mutate(variable = "cont") %>% 
+  rename(value = prev) %>% 
+  filter(value != 0)
+
+res2_long <- res2_long %>% bind_rows(cont2_res)
+
+rm(list = c("cont1_res", "cont2_res"))
+```
+
+```{r continuity-plot1, fig.cap="Continuity of settlements, pottery traditions."}
+plot_continuity <- function(x) {
+  x %>% 
+    group_by(period_label, reg) %>%
+    summarise(n = sum(value), prev = sum(prev)) %>%
+    mutate(perc = (prev / n) * 100) %>%
+    ggplot(aes(period_label, perc, fill = reg)) +
+    geom_col(show.legend = FALSE) +
+    scale_fill_manual(values = mycol) +
+    facet_wrap(vars(reg), ncol = 1) +
+    labs(x = "Period", y = "Continuity of settlement (%)") +
+    theme_minimal()
+}
+
+cont1 %>% plot_continuity()
+```
+```{r continuity-plot2, fig.cap="Continuity of settlements, pottery groups."}
+cont2 %>% plot_continuity()
+```
+
+<!--- ====================================================================== -->
+
+# Arrangement along natural lines
+
+Arrangement of settlements along natural 'terrain' lines or along waterways is explored.
+
+```{r along_lines}
+along_lines <- read_rds(here::here(derived_data, "lines/linear_arrangement.RDS"))
+
+along_lines1 <- along_lines$traditions %>% 
+  mutate(chrono = str_remove(chrono, ".$")) %>% 
+  group_by(id, chrono, period, variable) %>% 
+  summarise(value = max(value)) %>% 
+  ungroup(id, chrono, variable) %>% 
+  mutate(value = value / n())
+
+along_lines2 <- along_lines$groups %>% 
+  distinct(id, chrono, period, variable, value) %>% 
+  group_by(period) %>% 
+  mutate(value = value / n())
+
+# results 1
+res1_long <- res1_long %>% 
+  bind_rows(along_lines1)
+
+# results 2
+res2_long <- res2_long %>% 
+  bind_rows(along_lines2)
+```
+
+<!--- ====================================================================== -->
 
 # Synthesis
+
+```{r res-cleaning}
+na2null <- function(x) {
+  x[is.na(x)] <- 0
+  x
+}
+
+not_na <- function(x) {
+  !is.na(x)
+}
+
+# tradice
+res1_wide <- res1_long %>% 
+  group_by(id, period, variable) %>% 
+  summarise(value = mean(value), .groups = "drop") %>% 
+  pivot_wider(values_from = value, names_from = variable) %>%
+  mutate(across(c(cont, starts_with("line")), na2null)) %>%
+  filter(across(everything(), not_na))
+
+res1_long <- res1_wide %>% 
+  pivot_longer(-c(id, period), names_to = "variable", values_to = "value")
+
+# skupiny
+res2_wide <- res2_long %>% 
+  group_by(id, period, variable) %>% 
+  summarise(value = mean(value), .groups = "drop") %>% 
+  pivot_wider(values_from = value, names_from = variable) %>%
+  mutate(across(c(cont, starts_with("line")), na2null)) %>%
+  filter(across(everything(), not_na))
+
+res2_long <- res2_wide %>% 
+  pivot_longer(-c(id, period), names_to = "variable", values_to = "value")
+
+# scale cols
+res1_wide <- res1_wide %>% 
+  mutate(across(where(is.numeric), scale))
+
+res2_wide <- res2_wide %>% 
+  mutate(across(where(is.numeric), scale))
+```
+
+## Violin plots
+
+```{r variable-labels}
+labs_variables <- c(
+  "settlements_kde" = "Settlement density",
+  "dist_fenced" = "Min. dist. to fenced set. (km)",
+  "line_terrain" = "Along terrain line",
+  "line_water" = "Along water line",
+  "cont" = "Settlement continuity",
+  "hydro_kde" = "Watercourse density",
+  "dem" = "Altitude (DEM)", 
+  "altitude" = "Altitude (map)",
+  "slope" = "Slope (°)",
+  "rm_pc1" = "Distance to RM sources 1",
+  "rm_pc2" = "Distance to RM sources 2",
+  "rm_pc3" = "Distance to RM sources 3"
+) %>% 
+  map_chr(str_wrap, 12)
+```
+
+```{r syn-p1, fig.height=12}
+res1_violin <- res1_long %>% 
+  filter(variable != "cont") %>% 
+  mutate(reg = str_extract(id, "^."),
+         period = labs_chrono$periods[period],
+         period = fct_relevel(period, labs_chrono$periods),
+         variable = labs_variables[variable],
+         variable = fct_relevel(variable, labs_variables)) %>% 
+  ggplot(aes(reg, value)) +
+  geom_violin(fill = "white") +
+  geom_boxplot(aes(fill = reg), width = .2) +
+  scale_fill_manual(values = mycol, name = "Region") +
+  facet_grid(rows = vars(variable), cols = vars(period), scales = "free_y") +
+  theme_minimal() +
+  theme(legend.position = "bottom", 
+        panel.border = element_rect(color = "black", fill = NA), 
+        strip.text.y = element_text(angle = 0))
+
+res1_violin
+```
+
+```{r syn-p2, fig.height=12}
+res2_violin <- res2_long %>% 
+  filter(variable != "cont") %>% 
+  mutate(reg = str_extract(id, "^."),
+         period = labs_chrono$periods[period],
+         period = fct_relevel(period, labs_chrono$periods),
+         variable = labs_variables[variable],
+         variable = fct_relevel(variable, labs_variables)) %>% 
+  ggplot(aes(reg, value)) +
+  geom_violin(fill = "white") +
+  geom_boxplot(aes(fill = reg), width = .2) +
+  scale_fill_manual(values = mycol, name = "Region") +
+  facet_grid(rows = vars(variable), cols = vars(period), scales = "free_y") +
+  theme_minimal() +
+  theme(legend.position = "bottom", 
+        panel.border = element_rect(color = "black", fill = NA), 
+        strip.text.y = element_text(angle = 0))
+
+res2_violin
+```
+
+```{r res-violin-out}
+ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
+       width = 9, height = 12)
+ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
+       width = 9, height = 12)
+```
+
+## Line plots
+
+```{r syn-p1-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
+res1_line <- res1_long %>% 
+  mutate(reg = str_extract(id, "^."),
+         period = labs_chrono$periods[period],
+         period = fct_relevel(period, labs_chrono$periods),
+         variable = labs_variables[variable],
+         variable = fct_relevel(variable, labs_variables)) %>% 
+  group_by(period, variable, reg) %>% 
+  summarise(mean = mean(value),
+            median = median(value)) %>% 
+  ggplot(aes(x = period, color = reg, group = reg)) +
+  geom_line(aes(y = mean)) +
+  geom_line(aes(y = median), alpha = 0.4, size = 2) +
+  scale_color_manual(values = mycol, name = "Region") +
+  facet_wrap(vars(variable), scales = "free_y", ncol = 2) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 80, hjust = 1),
+        axis.title = element_blank())
+
+res1_line
+```
+
+```{r syn-p2-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
+res2_line <- res2_long %>% 
+  mutate(reg = str_extract(id, "^."),
+         period = labs_chrono$periods[period],
+         period = fct_relevel(period, labs_chrono$periods),
+         variable = labs_variables[variable],
+         variable = fct_relevel(variable, labs_variables)) %>% 
+  group_by(period, variable, reg) %>% 
+  summarise(mean = mean(value),
+            median = median(value), .groups = "drop") %>% 
+  ggplot(aes(x = period, color = reg, group = reg)) +
+  geom_line(aes(y = mean)) +
+  geom_line(aes(y = median), alpha = 0.4, size = 2) +
+  scale_color_manual(values = mycol, name = "Region") +
+  facet_wrap(vars(variable), scales = "free_y", ncol = 2) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 80, hjust = 1),
+        axis.title = element_blank())
+
+res2_line
+```
+
+```{r res-line-out}
+ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
+       width = 7, height = 7)
+ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
+       width = 7, height = 7)
+```
+
+## Analysing results
+
+```{r}
+pca2 <- res2_wide %>% 
+  group_by(period) %>% 
+  nest() %>% 
+  mutate(mx = map(data, function(x) select(x, -id)),
+         mx = map(mx, as.matrix),
+         pca = map(mx, prcomp, center = TRUE),
+         sdev = map(pca, function(x) tibble(sdev = x$sdev)),
+         sdev = map(sdev, function(x) mutate(x, 
+                                             var_prop = sdev^2 / sum(sdev^2),
+                                             cum_prop = cumsum(var_prop),
+                                             pc = row_number())),
+         scores = map(pca, function(x) as_tibble(x$x)),
+         rotation = map(pca, broom::tidy, matrix = "rotation"),
+         rotation = map(rotation, pivot_wider, names_from = "PC", names_prefix = "PC", values_from = "value")) %>%
+  select(-mx, -pca)
+
+pca2 %>% 
+  unnest(sdev) %>% 
+  ggplot(aes(pc, cum_prop)) +
+  geom_col() +
+  facet_wrap(~period)
+
+p_pca2_x <- pca2 %>% 
+  unnest(c(scores, data)) %>% 
+  mutate(reg = str_extract(id, "^.")) %>% 
+  ggplot(aes(PC1, PC2, color = reg)) +
+  geom_point(alpha = 0.1) +
+  facet_wrap(~period, ncol = 1) +
+  theme_minimal()
+
+p_pca2_a <- pca2 %>% 
+  unnest(c(rotation)) %>% 
+  ggplot(aes(PC1, PC2)) +
+  geom_segment(xend = 0, yend = 0, alpha = 0.4) +
+  geom_text(
+    aes(label = column),
+    hjust = 1, nudge_x = -0.02, 
+    color = "#904C2F",
+    size = 2.4
+  ) +
+  facet_wrap(~period, ncol = 1) +
+  theme_minimal()
+
+g_out <- gridExtra::grid.arrange(p_pca2_x, p_pca2_a, ncol = 2)
+
+ggsave(here::here("analysis/figures", "pca2.png"), g_out, height = 21)
+```
+
+## Nature vs Culture
+
+## Morava river valley vs East part of Bohemia
+
+<!--- ====================================================================== -->
 
@@
 res1_long %>% 
+  # distinct(id, chrono, period, variable, .keep_all = TRUE) %>% 
   pivot_wider(names_from = variable, values_from = value) %>% 
@@
 
+# res1_long %>% 
+#   add_count(id, chrono, period, variable) %>% 
+#   filter(n > 1)
+
 res2_long %>% 
+  # distinct(id, chrono, period, variable, .keep_all = TRUE) %>% 
   pivot_wider(names_from = variable, values_from = value) %>% 
@@
 ```
+
+<!--- ====================================================================== -->
 

### 2021-05-28 analysis/code/analysis_dem.R

@@ -1,0 +1,218 @@
+# Project "Settlements"
+# Script nr.
+# ANALYSIS DEM
+# author: Petr Pajdla
+# Getting values from DEM
+
+library(tidyverse)
+library(sf)
+library(stars)
+library(starsExtra)
+
+
+# data --------------------------------------------------------------------
+
+derived_data <- "analysis/data/derived_data"
+
+path_der <- here::here(derived_data, "dem/")
+path_maps <- here::here(derived_data, "maps/")
+
+# mask
+mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)
+# bboxes_m <- st_read(paste0(path_der, "bboxes_morava.geojson"))
+
+# dem
+# dem <- read_stars(paste0(path_der, "dem_aster_sjtsk.tif")) %>%
+#   st_set_crs(st_crs(bbox))
+# write_stars(dem, paste0(path_der, "dem_aster_sjtsk_stars.tif"))
+dem <- read_stars(paste0(path_der, "dem_aster_sjtsk_stars.tif"), proxy = FALSE)
+
+# settlements
+set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+# reduce dem ---------------------------------------------------------------
+
+# bbox_grid <- st_make_grid(bbox, cellsize = 5000)
+#
+# plot(bbox_grid)
+# plot(st_geometry(mask), add = TRUE)
+#
+# dem_grid <- aggregate(dem, bbox_grid, mean, na.rm = TRUE)
+#
+# dem_grid_sf <- st_as_sf(dem_grid)
+
+# st_write(dem_grid_sf, paste0(path_der, "dem_aster_sjtsk_sf.geojson"))
+dem_sf <- st_read(paste0(path_der, "dem_aster_sjtsk_sf.geojson")) %>%
+  rename(value = dem_aster_sjtsk.tif)
+
+
+# B -----------------------------------------------------------------------
+
+# bbox
+b_bbox <- set_spat %>%
+  filter(str_detect(id, "^B")) %>%
+  st_buffer(dist = 5e3) %>%
+  st_union() %>%
+  st_bbox()
+
+# crop
+b_dem <- dem[b_bbox]
+
+# slope
+b_slope <- slope(b_dem)
+
[…]

### 2021-05-28 analysis/code/dataprep_dem.R

@@ -1,0 +1,58 @@
+# Project "Settlements"
+# Script nr.
+# DEM PREP
+# author: Petr Pajdla
+# Preparation of rasters for slope calculation
+
+library(sf)
+library(raster)
+
+
+# data --------------------------------------------------------------------
+
+raw_data <- "analysis/data/raw_data"
+derived_data <- "analysis/data/derived_data"
+
+path_raw <- here::here(raw_data, "dem/")
+path_der <- here::here(derived_data, "dem/")
+path_maps <- here::here(derived_data, "maps/")
+
+if (!dir.exists(path_der)) {
+  dir.create(path_der)
+}
+
+# mask
+mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)
+
+# dem
+dem_aster <- raster(paste0(path_raw, "DEM_digital_elevation.tif"))
+
+# crop
+dem_aster_cropped <- dem_aster %>%
+  crop(st_transform(bbox, st_crs(dem_aster)))
+
+# reproject to sjtsk
+dem_aster_sjtsk <- dem_aster_cropped %>%
+  projectRaster(crs = crs(bbox))
+
+dem_aster_sjtsk_cropped <- dem_aster_sjtsk %>%
+  crop(bbox)
+
+# write raster for analysis
+writeRaster(dem_aster_sjtsk_cropped,
+            filename = paste0(path_der, "dem_aster_sjtsk.tif"),
+            overwrite = TRUE)
+
+
+# exploratory plots -------------------------------------------------------
+
+# plot(dem_aster_sjtsk_cropped)
+# plot(mask, col = NA, add = TRUE)
+#
+# res(dem_aster_sjtsk_cropped)
+#
+# plot(dem_aster_cropped)
+# plot(dem_aster_sjtsk)
+# plot(dem_aster_sjtsk_cropped)
+

### 2021-05-28 analysis/code/fenced_dist.R

@@ -1,0 +1,193 @@
+# Project "Settlements"
+# Script nr.
+# DISTANCE TO FENCED SETTLEMENT
+# author: Petr Pajdla
+# Calculated separately for the speed
+
+# devtools::load_all(".")
+library(tidyverse)
+library(sf)
+
+derived_data <- "analysis/data/derived_data"
+path_maps <- here::here(derived_data, "maps/")
+
+mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+
+# database
+set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
+
+set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+chrono_labels <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
+
+# settlements spatial by periods
+set_spat_1 <- set_spat %>%
+  right_join(set_base$period1) %>%
+  mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
+
+set_spat_2 <- set_spat %>%
+  right_join(set_base$period2) %>%
+  mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
+
+# fenced settlements
+b <- readxl::read_excel(
+  here::here("analysis/data/raw_data/ohrazena_sidliste.xlsx"),
+  sheet = "b")
+
+m <- readxl::read_excel(
+  here::here("analysis/data/raw_data/ohrazena_sidliste.xlsx"),
+  sheet = "m")
+
+# point layer
+# period I - pottery tradtions
+fortif_p1 <- bind_rows(b = b, m = m, .id = "reg") %>%
+  separate(abbrv_p2, into = c("p2_1", "p2_2"), sep = ";") %>%
+  pivot_longer(c(p2_1, p2_2), values_to = "abbrv_p2") %>%
+  select(-name) %>%
+  filter(!is.na(abbrv_p2)) %>%
+  mutate(abbrv_p2 = str_trim(abbrv_p2, "both"),
+         "p5.0" = if_else(abbrv_p1  == "SBK", TRUE, FALSE),
+         "p4.8" = if_else(abbrv_p1  %in% c("SBK", "LgK"), TRUE, FALSE),
+         "p4.6" = if_else(abbrv_p1  == "LgK", TRUE, FALSE),
[…]

### 2021-05-28 analysis/code/lines.R

@@ -1,0 +1,160 @@
+# Project "Settlements"
+# Script nr.
+# ARRANGEMENT ON LINES
+# author: Petr Pajdla
+#
+
+library(tidyverse)
+library(sf)
+
+derived_data <- "analysis/data/derived_data"
+path_maps <- here::here(derived_data, "maps/")
+
+mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+
+# database
+set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
+
+set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+chrono_labels <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
+
+# # settlements spatial by periods
+# set_spat_1 <- set_spat %>%
+#   right_join(set_base$period1) %>%
+#   mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
+#
+# set_spat_2 <- set_spat %>%
+#   right_join(set_base$period2) %>%
+#   mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
+
+# arrangement on lines
+ids <- set_base$base %>% pull(id)
+names(ids) <- set_base$base %>% pull(site)
+
+add_id <- function(x) {
+  x %>%
+    filter(site %in% set_base$base$site) %>%
+    mutate(id = unname(ids[site]))
+}
+
+pivot_terrain_and_water <- function(x) {
+  x %>%
+    pivot_longer(c(starts_with("w"), starts_with("t")), values_to = "line_len") %>%
+    filter(!is.na(line_len)) %>%
+    separate(name, into = c("type", "chrono"), sep = "_") %>%
+    group_by(type)
+}
+
+pivot_period <- function(x) {
+  x %>%
+    pivot_longer(cols = starts_with("p"), names_to = "period") %>%
+    filter(value) %>%
+    select(-value) %>%
+    mutate(period_label = chrono_labels$periods[period],
+           period_label = factor(period_label, levels = chrono_labels$periods))
+}
+
+# east art of bohemia
+# traditions
+b1 <- readxl::read_excel(
[…]

### 2021-05-28 analysis/code/results_to_gd.R

@@ -36,3 +36,12 @@ drive_readme <- drive_upload(readme, gd_path, overwrite = TRUE)
 si <- here("analysis/paper/paper.html")
 
 drive_si <- drive_upload(si, "~/settlements/SI/", overwrite = TRUE)
+
+
+# Figs to GD --------------------------------------------------------------
+
+figs <- here("analysis/figures")
+figs_up <- str_c(figs, "/", list.files(figs, pattern = "^res"))
+
+map(figs_up, ~ drive_upload(.x, "~/settlements/figures/",
+                                overwrite = TRUE))

### 2021-05-28 analysis/code/rm_distance.R

@@ -39,12 +39,12 @@ dist <- bind_rows(p = dist_pts, l = dist_lns, .id = "shp") %>%
          type = if_else(rm %in% pstrm, "Polished", "Chipped"),
          region = if_else(str_detect(id, "^B"), "Boh.", "Mor."))
 
+
 # save result
 
 if (!dir.exists(temp_data)) {
   dir.create(temp_data)
 }
-
 
 write_csv(dist, paste0(temp_data, "/rm_dist.csv"))
 

### 2021-09-15 R/continuity.R
update pckg functions all functions extracted from Rmd doc to separate scripts

@@ -1,0 +1,55 @@
+# Settlement continuity
+
+#' Derive settlement continuity across periods
+#'
+#' @param x Input period \code{tibble}.
+#'
+#' @return A \code{tibble}.
+#' @export
+#'
+#' @examples
+settlement_continuity <- function(x) {
+  x %>%
+    select(-period_label) %>%
+    distinct(id, period, chrono) %>%
+    mutate(pres = 1L) %>%
+    pivot_wider(names_from = period, values_from = pres, values_fill = 0L) %>%
+    pivot_longer(starts_with("p"), names_to = "period") %>%
+    mutate(period = factor(period, levels = levels(set_base$period1$period))) %>%
+    arrange(id, period, chrono) %>%
+    group_by(id, chrono) %>%
+    mutate(prev = if_else(value == 1, lag(value), 0L),
+           prev = if_else(is.na(prev), 0L, prev)) %>%
+    ungroup(id, chrono) %>%
+    select(-value, -chrono) %>%
+    mutate(variable = "cont") %>%
+    rename(value = prev) %>%
+    filter(value != 0)
+}
+
+#' Plot settlement continuity
+#'
+#' @param x Long results \code{tibble}.
+#' @param y A \code{tibble} with number of settlements.
+#'
+#' @return A \code{ggplot2} object.
+#' @export
+#'
+#' @examples
+plot_continuity <- function(x, y) {
+  x %>%
+    filter(variable == "cont") %>%
+    add_period_label() %>%
+    add_region() %>%
+    group_by(period_label, reg) %>%
+    summarise(n = sum(value)) %>%
+    full_join(y) %>%
+    mutate(n = if_else(is.na(n), 0, n),
+           perc = (n / sum) * 100) %>%
+    ggplot(aes(period_label, perc, fill = reg)) +
+    geom_col(show.legend = FALSE) +
+    scale_fill_manual(values = mycol) +
+    facet_wrap(vars(reg), ncol = 1) +
+    labs(x = "Period", y = "Continuity of settlement (%)") +
+    theme_minimal()
+}

### 2021-09-15 R/density.R

@@ -33,14 +33,14 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
     factor(levels = lvls)
 
   pppset <- spatstat.geom::ppp(x = coords[, 1], y = coords[, 2],
-                          window = mask,
-                          marks = marks)
+                               window = mask,
+                               marks = marks)
   splitpppset <- spatstat.geom::split.ppp(pppset)
 
   if (points) {
     dens <- spatstat.core::density.splitppp(splitpppset,
-                                       at = "points",
-                                       sigma = sigma) %>%
+                                            at = "points",
+                                            sigma = sigma) %>%
       purrr::map(as_tibble) %>%
       dplyr::bind_rows(.id = markscol) %>%
       dplyr::group_by(!!as.symbol(markscol)) %>%
@@ -64,7 +64,7 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
       dplyr::select(id, name, chrono, kde)
   } else {
     img <- spatstat.core::density.splitppp(splitpppset,
-                                      sigma = sigma)
+                                           sigma = sigma)
     rstr <- vector("list", length(img))
     names(rstr) <- lvls
 
@@ -78,4 +78,59 @@ kdestimate <- function(x, mask, markscol, lvls, sigma = 4e3, points = TRUE) {
       dplyr::mutate(kde = layer * 1e6) %>%
       dplyr::select(markscol, x, y, kde, -layer)
   }
+}
+
+
+#' Helper function to estimate KDE
+#'
+#' @param x Simple features object. Input data.
+#'
+#' @return A long \code{tibble} with kernel density estimate for each id and period pair.
+#' @export
+#'
+#' @examples
+est_kde <- function(x) {
+  x %>% group_by(id, period) %>%
+    mutate(n = row_number()) %>%
+    filter(n == 1L) %>%
+    select(-label) %>%
+    kdestimate(mask,
+               markscol = "period",
+               lvls = names(labs_chrono$period)) %>%
+    rename(period = name) %>%
+    mutate(period_label = labs_chrono$periods[period],
+           period_label = factor(period_label, levels = labs_chrono$periods)) %>%
+    select(id, starts_with("period"), kde) %>%
+    rename(settlements_kde = kde) %>%
+    select(-period_label) %>%
[…]

### 2021-09-15 R/distances.R

@@ -1,58 +1,0 @@
-#' Distance to a source of a raw material
-#'
-#' For simple feature geometries in projected coordinate reference system,
-#' euclidean distance (in km) is returned between features of \code{from}
-#' and \code{to} categorized by levels specified in \code{rm}.
-#'
-#' @param from \code{sf} object of geometry type \code{POINTS} in projected
-#'     CRS (settlements).
-#' @param to \code{sf} object of geometry type \code{POINT} or \code{LINESTRING}
-#'     in projected CRS (raw material locations).
-#'     Each raw material can have multiple locations, same raw materials are
-#'     identified by identical label in column specified by \code{rm} argument.
-#' @param rm Column name of \code{to} to use as a grouping variable
-#'     (identical raw material).
-#' @param id Column name of \code{from} containing unique identifiers.
-#'
-#' @return A \code{tibble} containing min, max, mean and median distance
-#'     for each raw material (level of \code{rm} column) to an observation
-#'     specified in \code{from} (settlement) is returned.
-#'
-dist_to_rm <- function(from, to, rm, id = "id") {
-
-  # set units to km
-  to_km <- function(x) {
-    units(x) <- "km"
-    return(x)
-  }
-
-  lvls <- levels(factor(pull(to, {{ rm }} )))
-
-  res <- vector("list", length(lvls))
-  names(res) <- lvls
-
-  for (i in seq_along(lvls)) {
-
-    rm_flt <- filter(to, .data[[rm]] == lvls[i])
-
-    res[[lvls[i]]] <- st_distance(from, rm_flt) %>%
-      as_tibble() %>%
-      # first col must contain ids
-      bind_cols(st_drop_geometry(select(from, all_of(id)))) %>%
-      pivot_longer(-id) %>%
-      # to kilometers
-      mutate(value = to_km(value)) %>%
-      group_by(id) %>%
-      summarize(min = min(value),
-                mean = mean(value),
-                median = median(value),
-                max = max(value), .groups = "drop")
-
-    # print(paste(lvls[i], "done,", i, "of", length(lvls)))
-
-  }
-
-  res %>%
-    bind_rows(.id = "rm")
-
-}

### 2021-09-15 R/minions.R

@@ -1,0 +1,125 @@
+#' NA to zero
+#'
+#' Changes NA value to a zero (0).
+#'
+#' @param x An atomic vector.
+#'
+#' @return An atomic numeric vector.
+#' @export
+#'
+#' @examples
+na2null <- function(x) {
+  x[is.na(x)] <- 0
+  x
+}
+
+# not_na <- function(x) {
+#   !is.na(x)
+# }
+
+
+# Labelling functions ---------------------------------------------------
+
+#' Add period labels
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A \code{tibble}.
+#' @export
+add_period_label <- function(x) {
+  x %>% mutate(period_label = unname(labs_chrono$periods[period]),
+               period_label = factor(period_label, levels = unname(labs_chrono$periods)))
+}
+
+#' Add region labels
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A \code{tibble}.
+#' @export
+add_region <- function(x) {
+  x %>% mutate(reg = str_extract(id, "^."),
+               reg = factor(reg, levels = c("B", "M")))
+}
+
+#' Add labels Neolithic B and C
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A \code{tibble}.
+#' @export
+#'
+#' @examples
+add_neo <- function(x) {
+  x %>% mutate(neo = case_when(
+    period %in% c("p5.0", "p4.8", "p4.6", "p4.4", "p4.2", "p4.0") ~ "NeoB",
+    period %in% c("p3.8", "p3.6", "p3.4") ~ "NeoC"
+  ),
+  neo = factor(neo, levels = c("NeoB", "NeoC")))
+}
+
+#' Abbreviate variable names
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A \code{tibble}.
+#' @export
+#'
+#' @examples
+relabel_variable <- function(x) {
+  x %>% mutate(var = case_when(
+    column == "altitude" ~ "alt.",
+    column == "cont" ~ "cont.",
+    column == "dist_fenced" ~ "d. fenced",
+    column == "hydro_kde" ~ "hydro.",
+    column == "line_terrain" ~ " t. line",
+    column == "settlements_kde" ~ "dens.",
+    column == "slope" ~ "slope",
+    column == "rm_dist_chipped" ~ "d. ch.",
+    column == "line_water" ~ "w. line",
+    column == "rm_dist_polished" ~ "d. pol.",
+  ))
+}
+
+
[…]

### 2021-09-15 R/output_fns.R

@@ -1,0 +1,95 @@
+# Output preparation functions --------------------------------------------
+
+# Functions with \code{prep_} prefix prepare the data into the long format for
+# binding with results \code{tibble}.
+
+
+#' Prepare density of watercourse result data
+#'
+#' Functions with \code{prep_} prefix prepare the data into the long format for
+#' binding with results \code{tibble}.
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A long results \code{tibble}.
+#' @export
+#'
+#' @examples
+prep_hydro_dens_res <- function(x) {
+  x %>% select(-ends_with("label"), -chrono) %>%
+    left_join(hydro_dens_at_points, by = "id") %>%
+    pivot_longer(cols = hydro_kde) %>%
+    rename(variable = name)
+}
+
+#' Prepare DEM result
+#'
+#' Functions with \code{prep_} prefix prepare the data into the long format for
+#' binding with results \code{tibble}.
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A long results \code{tibble}.
+#' @export
+#'
+#' @examples
+prep_dem_res <- function(x) {
+  x %>% select(-ends_with("label"), -chrono) %>%
+    left_join(dem_data, by = "id") %>%
+    pivot_longer(cols = c(dem, slope)) %>%
+    rename(variable = name)
+}
+
+#' Prepare altitude result
+#'
+#' Functions with \code{prep_} prefix prepare the data into the long format for
+#' binding with results \code{tibble}.
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A long results \code{tibble}.
+#' @export
+#'
+#' @examples
+prep_alt_res <- function(x) {
+  x %>% select(-ends_with("label"), -chrono) %>%
+    left_join(altitude, by = "id") %>%
+    pivot_longer(cols = altitude) %>%
+    rename(variable = name)
+}
+
+#' Prepare distance to fenced settlements result
+#'
+#' Functions with \code{prep_} prefix prepare the data into the long format for
+#' binding with results \code{tibble}.
+#'
+#' @param x A \code{tibble}.
+#'
+#' @return A long results \code{tibble}.
+#' @export
+#'
+#' @examples
+prep_dist_fenced <- function(x) {
+  x %>% mutate(variable = "dist_fenced",
+               value = min_dist) %>%
+    select(-min_dist, -reg)
+}
+
+#' Prepare wide results \code{tibble}
+#'
[…]

### 2021-09-15 R/pca.R

@@ -1,0 +1,100 @@
+#' Nested PCA
+#'
+#' @param x A \code{tibble} grouped according to chosen variables.
+#'
+#' @return A nested \code{tibble} with results of \code{prcomp} function.
+#' @export
+#'
+#' @examples
+nested_pca <- function(x) {
+  x %>%
+    mutate(
+      id = map(data, \(x) select(x, id)),
+      data = map(data, \(x) select(x, -id)),
+      mx = map(data, \(x) complete(x)),
+      mx = map(mx, \(x) select(x, where(~ sum(.x) != 0))),
+      mx = map(mx, \(x) select(x, where(~ sum(.x)/length(.x) != 1))),
+      mx = map(mx, \(x) mutate(x, across(where(is.numeric), scale))),
+      mx = map(mx, as.matrix),
+      pca = map(mx, prcomp),
+      sdev = map(pca, \(x) tibble(sdev = x$sdev)),
+      sdev = map(sdev, \(x) mutate(x,
+                                   var_prop = sdev^2 / sum(sdev^2),
+                                   cum_prop = cumsum(var_prop),
+                                   pc = row_number())),
+      x = map(pca, \(x) as_tibble(x$x)),
+      rotation = map(pca, broom::tidy, matrix = "rotation"),
+      rotation = map(rotation, pivot_wider,
+                     names_from = "PC", names_prefix = "PC",
+                     values_from = "value")
+    ) %>%
+    select(-mx, -pca)
+}
+
+#' Plot nested PCA results
+#'
+#' @param x Nested PCA object, output of \code{nested_pca} function.
+#' @param var Grouping variable visualize on.
+#' @param pc Principal components to show on axes x and y.
+#'     Defaults to PC1 and PC2.
+#'
+#' @return Returns a \code{patchwork} of data in selected PC space and
+#'     associated plots with rotation visualization.
+#' @export
+#'
+#' @examples
+plot_nested_pca <- function(x, var, pc = c("PC1", "PC2")) {
+  pca <- x %>%
+    select(reg, !!sym(var), x) %>%
+    unnest(x) %>%
+    ungroup(!!sym(var), reg) %>%
+    ggplot(aes(!!sym(pc[1]), !!sym(pc[2]))) +
+    geom_point(alpha = 0.2) +
[…]

### 2021-09-15 R/rm_distance.R

@@ -1,0 +1,227 @@
+#' Distance to a source of a raw material
+#'
+#' For simple feature geometries in projected coordinate reference system,
+#' euclidean distance (in km) is returned between features of \code{from}
+#' and \code{to} categorized by levels specified in \code{rm}.
+#'
+#' @param from \code{sf} object of geometry type \code{POINTS} in projected
+#'     CRS (settlements).
+#' @param to \code{sf} object of geometry type \code{POINT} or \code{LINESTRING}
+#'     in projected CRS (raw material locations).
+#'     Each raw material can have multiple locations, same raw materials are
+#'     identified by identical label in column specified by \code{rm} argument.
+#' @param rm Column name of \code{to} to use as a grouping variable
+#'     (identical raw material).
+#' @param id Column name of \code{from} containing unique identifiers.
+#'
+#' @return A \code{tibble} containing min, max, mean and median distance
+#'     for each raw material (level of \code{rm} column) to an observation
+#'     specified in \code{from} (settlement) is returned.
+#'
+dist_to_rm <- function(from, to, rm, id = "id") {
+
+  # set units to km
+  to_km <- function(x) {
+    units(x) <- "km"
+    return(x)
+  }
+
+  lvls <- levels(factor(pull(to, {{ rm }} )))
+
+  res <- vector("list", length(lvls))
+  names(res) <- lvls
+
+  for (i in seq_along(lvls)) {
+
+    rm_flt <- filter(to, .data[[rm]] == lvls[i])
+
+    res[[lvls[i]]] <- st_distance(from, rm_flt) %>%
+      as_tibble() %>%
+      # first col must contain ids
+      bind_cols(st_drop_geometry(select(from, all_of(id)))) %>%
+      pivot_longer(-id) %>%
+      # to kilometers
+      mutate(value = to_km(value)) %>%
+      group_by(id) %>%
+      summarize(min = min(value),
+                mean = mean(value),
+                median = median(value),
+                max = max(value), .groups = "drop")
+
+    # print(paste(lvls[i], "done,", i, "of", length(lvls)))
+
+  }
+
+  res %>%
+    bind_rows(.id = "rm")
+
+}
+
+#' Plot linear model for distance to raw material vs settlements kde
+#'
+#' @param x Long results object.
+#'
[…]

### 2021-09-15 R/synthesis.R

@@ -1,0 +1,63 @@
+
+# Synthesis plot functions ------------------------------------------------
+
+#' Summary violin/box plots
+#'
+#' @param x Long result \code{tibble}.
+#'
+#' @return A \code{ggplot2} object.
+#' @export
+#'
+#' @examples
+plot_violin <- function(x) {
+  x %>%
+    filter(variable != "cont",
+           variable != "dem") %>%
+    mutate(reg = str_extract(id, "^."),
+           period = unname(labs_chrono$periods[period]),
+           period = factor(period, levels = unname(labs_chrono$periods)),
+           variable = labs_variables[variable],
+           variable = fct_relevel(variable, labs_variables)) %>%
+    ggplot(aes(reg, value)) +
+    # geom_violin(fill = "white") +
+    geom_jitter(alpha = 0.04, width = 0.4) +
+    geom_boxplot(aes(fill = reg), width = 0.2) +
+    scale_fill_manual(values = mycol, name = "Region") +
+    facet_grid(rows = vars(variable), cols = vars(period), scales = "free_y") +
+    theme_minimal() +
+    theme(legend.position = "bottom",
+          panel.border = element_rect(color = "black", fill = NA),
+          plot.background = element_rect(fill = "white", color = "white"),
+          strip.text.y = element_text(angle = 0))
+}
+
+#' Summary line plots
+#'
+#' @param x Long result \code{tibble}.
+#'
+#' @return A \code{ggplot2} object.
+#' @export
+#'
+#' @examples
+plot_lines <- function(x) {
+  x %>%
+    filter(variable != "cont",
+           variable != "dem") %>%
+    mutate(reg = str_extract(id, "^."),
+           period = unname(labs_chrono$periods[period]),
+           period = factor(period, levels = unname(labs_chrono$periods)),
+           variable = labs_variables[variable],
+           variable = fct_relevel(variable, labs_variables)) %>%
+    group_by(period, variable, reg) %>%
+    summarise(mean = mean(value),
+              median = median(value)) %>%
+    ggplot(aes(x = period, color = reg, group = reg)) +
+    geom_line(aes(y = mean)) +
+    geom_line(aes(y = median), alpha = 0.4, size = 2) +
+    scale_color_manual(values = mycol, name = "Region") +
+    facet_wrap(vars(variable), scales = "free_y", ncol = 2) +
+    theme_minimal() +
+    theme(axis.text.x = element_text(angle = 80, hjust = 1),
[…]

### 2021-09-15 analysis/code/dataprep.R
check code prep steps for errors

@@ -28,6 +28,11 @@ for (i in seq_along(files_gd$id)) {
 # prepare derived data ----------------------------------------------------
 labs_chrono <- read_rds(here::here(dt_der, "chrono_labels.RDS"))
 
+# labs_chrono$periods["p5.0"] <- "4.9 – 4.8 k"
+# labs_chrono$periods["p3.4"] <- "3.4 – 3.3 k"
+#
+# write_rds(labs_chrono, here::here(dt_der, "chrono_labels.RDS"))
+
 # list downloaded files
 files_disk <- list.files(here(dt_der, "input_gd"), full.names = TRUE)
 
@@ -45,13 +50,33 @@ set_base$base <- read_csv(str_subset(files_disk, "base")) %>%
 
 set_base$period1 <- read_csv(str_subset(files_disk, "period1")) %>%
   filter(id %in% set_base$base$id) %>%
-  mutate(period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
-         period = fct_relevel(period, names(labs_chrono$periods)))
+  mutate(
+    period_label = labs_chrono$periods[period],
+    period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
+    period = fct_relevel(period, names(labs_chrono$periods)),
+    keep = if_else(str_detect(id, "^B") &
+                     period == "p4.8" &
+                     chrono == "LgK", FALSE, TRUE)
+  ) %>%
+  filter(keep) %>%
+  select(-keep)
 
 set_base$period2 <- read_csv(str_subset(files_disk, "period2")) %>%
   filter(id %in% set_base$base$id) %>%
-  mutate(period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
-         period = fct_relevel(period, names(labs_chrono$periods)))
+  mutate(
+    period_label = labs_chrono$periods[period],
+    period_label = fct_relevel(period_label, unname(labs_chrono$periods)),
+    period = fct_relevel(period, names(labs_chrono$periods)),
+    keep = if_else(str_detect(id, "^B") &
+                     period == "p4.8" &
+                     chrono == "LgK2", FALSE, TRUE),
+    keep = if_else(period == "p5.0" &
+                     chrono == "SBK2", FALSE, keep),
+    keep = if_else(period == "p3.6" &
+                     chrono == "TRB1", FALSE, keep)
+  ) %>%
+  filter(keep) %>%
[…]

### 2021-09-15 analysis/code/dataprep_hydro.R

@@ -58,7 +58,7 @@ cz_geom <- st_combine(st_geometry(cz_crop))
 rm(list = c("cz_crop", "au_crop", "sk_crop"))
 
 # save temp files...
-# dt_hydro_derived <- "analysis/data/derived_data/hydro/"
+dt_hydro_derived <- "analysis/data/derived_data/hydro/"
 # dir.create(here(dt_hydro_derived))
 # st_write(sk_geom, here(dt_hydro_derived, "sk_prep.geojson"))
 # st_write(au_geom, here(dt_hydro_derived, "au_prep.geojson"))

### 2021-09-15 analysis/code/lines.R

@@ -56,7 +56,7 @@ pivot_period <- function(x) {
            period_label = factor(period_label, levels = chrono_labels$periods))
 }
 
-# east art of bohemia
+# east part of bohemia
 # traditions
 b1 <- readxl::read_excel(
   here::here("analysis/data/raw_data/lines.xlsx"),
@@ -65,7 +65,7 @@ b1 <- readxl::read_excel(
   add_id() %>%
   pivot_terrain_and_water() %>%
   mutate("p5.0" = if_else(chrono  %in% c("SBK1", "SBK2"), TRUE, FALSE),
-         "p4.8" = if_else(chrono  %in% c("SBK1", "SBK2", "LgK2"), TRUE, FALSE),
+         "p4.8" = if_else(chrono  %in% c("SBK1", "SBK2"), TRUE, FALSE),
          "p4.6" = if_else(chrono  %in% c("LgK2"), TRUE, FALSE),
          "p4.4" = if_else(chrono  %in% c("LgK2", "ENE0"), TRUE, FALSE),
          "p4.2" = if_else(chrono  == "ENE0", TRUE, FALSE),
@@ -75,21 +75,21 @@ b1 <- readxl::read_excel(
          "p3.4" = if_else(chrono  %in% c("TRB1", "TRB2"), TRUE, FALSE)) %>%
   pivot_period()
 
-# grousp
+# groups
 b2 <- readxl::read_excel(
   here::here("analysis/data/raw_data/lines.xlsx"),
   sheet = "b") %>%
   select(site = NAZLOK, site2 = ALTERNAZ, starts_with("w"), starts_with("t")) %>%
   add_id() %>%
   pivot_terrain_and_water() %>%
-  mutate("p5.0" = if_else(chrono  %in% c("SBK1", "SBK2"), TRUE, FALSE),
-         "p4.8" = if_else(chrono  %in% c("SBK1", "SBK2", "LgK2"), TRUE, FALSE),
+  mutate("p5.0" = if_else(chrono  %in% c("SBK1"), TRUE, FALSE),
+         "p4.8" = if_else(chrono  %in% c("SBK1", "SBK2"), TRUE, FALSE),
          "p4.6" = if_else(chrono  %in% c("LgK2"), TRUE, FALSE),
          "p4.4" = if_else(chrono  %in% c("LgK2", "ENE0"), TRUE, FALSE),
[…]

### 2021-09-15 analysis/code/results_to_gd.R

@@ -26,8 +26,8 @@ files_local <- list.files(dt_result, pattern = "csv$", full.names = TRUE)
 readme <- list.files(dt_result, pattern = "^README$", full.names = TRUE)
 
 drive_results <- map(files_local, ~ drive_upload(.x, gd_path,
-                                           type = "spreadsheet",
-                                           overwrite = TRUE))
+                                                 type = "spreadsheet",
+                                                 overwrite = TRUE))
 drive_readme <- drive_upload(readme, gd_path, overwrite = TRUE)
 
 
@@ -41,7 +41,7 @@ drive_si <- drive_upload(si, "~/settlements/SI/", overwrite = TRUE)
 # Figs to GD --------------------------------------------------------------
 
 figs <- here("analysis/figures")
-figs_up <- str_c(figs, "/", list.files(figs, pattern = "^res"))
+figs_up <- str_c(figs, "/", list.files(figs, pattern = "^res|^pca"))
 
 map(figs_up, ~ drive_upload(.x, "~/settlements/figures/",
-                                overwrite = TRUE))
+                            overwrite = TRUE))

### 2021-09-15 analysis/paper/paper.Rmd
update paper, PCA, hclust + dendro

--- 2021-02-11 analysis/paper/paper.Rmd
@@
   echo = FALSE,
-  cache = TRUE,
+  cache = FALSE,
   comment = "#>",
-  fig.path = "../figures/",
+  # fig.path = "../figures/",
   dpi = 300
@@
 
-```{r packages}
-devtools::load_all(".")
+```{r packages, cache=FALSE}
 library(sf)
 library(tidyverse)
+library(spatstat)
+library(patchwork)
+devtools::load_all(".")
 ```
@@
 
-The analysis is divided into two 'time' scales:
+The analysis is divided into two scales:
 
@@
 settlements1 <- set_spat %>%
-  right_join(set_base$period1, by = "id")
+  right_join(set_base$period1, by = "id") %>% 
+  mutate(label = labs_chrono$chrono1[chrono])
 settlements2 <- set_spat %>%
-  right_join(set_base$period2, by = "id")
+  right_join(set_base$period2, by = "id") %>% 
+  mutate(label = labs_chrono$chrono2[chrono])
+
+# colors
+mycol <- c("B" = "#7FC97F", "M" = "#BEAED4", "b" = "#7FC97F", "m" = "#BEAED4"
+)
 ```
@@
 set_base$period1 %>%
-  group_by(period_label) %>%
-  summarize(n = n()) %>% 
-  rename('Period' = period_label) %>% 
-  knitr::kable(caption = "Number of settlements across time periods.") %>% 
-  kableExtra::kable_styling(full_width = FALSE)
+  table_nr_set()
 ```
@@
 settlements1 %>%
-  ggplot() +
-  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
-  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
-  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
-  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
-  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
-  geom_sf(alpha = 0.2, aes(color = label)) +
-  scale_color_brewer(palette = "Set1", name = "Tradition") +
-  scale_x_continuous(expand = c(0, 0)) +
-  scale_y_continuous(expand = c(0, 0)) +
-  facet_wrap(vars(period_label), nrow = 3) +
-  theme_void() +
-  theme(panel.border = element_rect(color = "black", fill = NA))
+  plot_nr_set()
 ```
@@
 set_base$period2 %>%
-  group_by(period_label) %>%
-  summarize(n = n()) %>% 
-  rename('Period' = period_label) %>% 
-  knitr::kable(caption = "Number of settlements across time periods.") %>% 
-  kableExtra::kable_styling(full_width = FALSE)
+  table_nr_set()
 ```
@@
 settlements2 %>%
-  ggplot() +
-  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
-  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
-  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
-  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
-  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
-  geom_sf(alpha = 0.2, aes(color = label)) +
-  scale_color_brewer(palette = "Set1", name = "Tradition") +
-  scale_x_continuous(expand = c(0, 0)) +
-  scale_y_continuous(expand = c(0, 0)) +
-  facet_wrap(vars(period_label), nrow = 3) +
-  theme_void() +
-  theme(panel.border = element_rect(color = "black", fill = NA))
+  plot_nr_set()
 ```
@@
 ```{r set1-kde}
-kde1 <- kdestimate(settlements1,
-                   mask,
-                   markscol = "period",
-                   lvls = names(labs_chrono$period)) %>%
-  rename(period = name) %>%
-  mutate(period_label = labs_chrono$periods[period],
-         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  select(id, starts_with("period"), chrono, kde)
-
-# tibble with results
-res1_long <- kde1 %>% 
-  rename(settlements_kde = kde) %>% 
-  select(-period_label) %>% 
-  pivot_longer(col = settlements_kde, names_to = "variable")
+res1_long <- settlements1 %>% 
+  est_kde()
 ```
@@
 ```{r plot-set1-kde, fig.cap = 'Kernel density estimation for different periods.'}
-kdestimate(settlements1,
-           mask,
-           markscol = "period",
-           lvls = names(labs_chrono$period),
-           points = FALSE) %>%
-  mutate(period_label = labs_chrono$periods[period],
-         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  ggplot(aes(x, y, fill = kde)) +
-  geom_raster() +
-  scale_fill_viridis_c(name = "KDE") +
-  facet_wrap(vars(period_label)) +
-  coord_fixed() +
-  theme_void()
+settlements1 %>% 
+  plot_kde()
 ```
@@
 ```{r set2-kde}
-kde2 <- kdestimate(settlements2,
-                   mask,
-                   markscol = "period",
-                   lvls = names(labs_chrono$period)) %>%
-  rename(period = name) %>%
-  mutate(period_label = labs_chrono$periods[period],
-         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  select(id, starts_with("period"), chrono, kde)
-
-# tibble with results
-res2_long <- kde2 %>% 
-  rename(settlements_kde = kde) %>% 
-  select(-period_label) %>% 
-  pivot_longer(col = settlements_kde, names_to = "variable")
+res2_long <- settlements2 %>% 
+  est_kde()
 ```
@@
 ```{r plot-set2-kde, fig.cap = 'Kernel density estimation for different periods.'}
-kdestimate(settlements2,
-           mask,
-           markscol = "period",
-           lvls = names(labs_chrono$period),
-           points = FALSE) %>%
-  mutate(period_label = labs_chrono$periods[period],
-         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
-  ggplot(aes(x, y, fill = kde)) +
-  geom_raster() +
-  scale_fill_viridis_c(name = "KDE") +
-  facet_wrap(vars(period_label)) +
-  coord_fixed() +
-  theme_void()
+settlements2 %>% 
+  plot_kde()
 ```
@@
 rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
-  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+  add_rm_type()
+
 rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
-  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
+  add_rm_type()
 ```
@@
 
+## Distances
+
 ```{r rm-dist, cache = FALSE}
@@
 
-rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))
+rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv")) %>% 
+  mutate(rm = str_to_lower(rm))
 
-rm_dist_median <- rm_dist %>% 
-  select(-region, -type, -shp, -min, -mean, -max) %>% 
-  group_by(id, rm) %>% 
-  summarise(dist_rm = mean(median)) %>% 
-  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
-  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
-  select(-rm)
+# rm_dist_mean <- rm_dist %>%
+#   select(-region, -type, -shp, -min, -median, -max) %>%
+#   group_by(id, rm) %>%
+#   summarise(dist_rm = mean(mean)) %>%
+#   pivot_longer(cols = dist_rm, names_to = "variable") %>%
+#   mutate(variable = str_c(variable, rm, sep = "_")) %>%
+#   select(-rm)
+```
 
-# update result tibbles
-res1_long <- set_base$period1 %>% 
-  select(id, period, chrono) %>% 
-  inner_join(rm_dist_median) %>%
+<!-- ```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."} -->
+<!-- rm_dist %>% -->
+<!--   ggplot(aes(x = min, y = rm, fill = region)) + -->
+<!--   geom_violin() + -->
+<!--   scale_fill_brewer(palette = "Accent") + -->
+<!--   facet_wrap(~type, scales = "free_y") + -->
+<!--   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") + -->
+<!--   theme_minimal() + -->
+<!--   theme(legend.position = "bottom",  -->
+<!--         panel.border = element_rect(color = "black", fill = NA)) -->
+<!-- ``` -->
+
+## Settlement density against distance to raw materials
+
+> Hypothesis: In Morava river catchment settlement position is directly associated with 
+raw material source.
+
+We compare settlement density (KDE) with distance to chosen raw materials.
+There is no strong evidence that the distribution of settlements is associated with distance to 
+any of the key raw material sources occurring in the region in any of the studied periods.
+There are some mild negative correlations in the period 4.8 -- 4.2 k BCE for polished stone tools materials.
+Strongest connection is observed between given pottery groups in period 4.8 -- 4.6 k BCE in case of ad/pmd/zelesice raw material sources.
+
+```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
+res1_long %>% plot_dist_lm()
+```
+
+```{r tab-rm-dist-cor-kde1}
+res1_long %>% tab_dist_lm()
+```
+
+```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
+res2_long %>% plot_dist_lm()
+```
+
+```{r tab-rm-dist-cor-kde2}
+res2_long %>% tab_dist_lm()
+```
+
+<!--
+## PCA for raw material distances
+
+Principal components analysis is used to reduce dimensionality.
+First three PCs are enough to cover more than 95% variability in the data set.
+
+```{r rm-dist-pca}
+# scale_col <- function(x){
+#   (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
+# }
+# 
+# rm_dist_mean_wide <- rm_dist_mean %>%
+#   mutate(variable = str_remove(variable, "dist_rm_")) %>% 
+#   pivot_wider(names_from = variable, values_from = value) %>% 
+#   ungroup() %>% 
+#   mutate(across(where(is.numeric), scale_col))
+# 
+# # matrix for pca to reduce dimensions
+# rm_dist_mx <- rm_dist_mean_wide %>% 
+#   select(-id) %>% 
+#   as.matrix() 
+# row.names(rm_dist_mx) <- rm_dist_mean_wide$id
+# 
+# rm_pca <- prcomp(rm_dist_mx)
+# summary(rm_pca)
+```
+
+```{r rm-corrplot, fig.cap="Correlation between principal components and given raw materials."}
+# corrplot::corrplot(t(rm_pca$rotation[, 1:3]))
+```
+```{r rm-biplot, fig.cap="First thee principal components (green - East part of Bohemia, blue - Morava river catchment)."}
+# pairs(rm_pca$x[, 1:3], col = unname(mycol[str_extract(rownames(rm_dist_mx), ".")]))
+```
+
+```{r update-result-rm}
+# rm_dist_pca <- rm_pca$x[, 1:3] %>% 
+#   as_tibble(rownames = "id") %>% 
+#   rename("rm_pc1" = "PC1",
+#          "rm_pc2" = "PC2",
+#          "rm_pc3" = "PC3",) %>% 
+#   pivot_longer(-id, "variable", "value")
+# 
+# # update result tibbles
+# res1_long <- set_base$period1 %>% 
+#   select(id, period, chrono) %>% 
+#   inner_join(rm_dist_pca) %>%
+#   bind_rows(res1_long)
+# 
+# res2_long <- set_base$period2 %>% 
+#   select(id, period, chrono) %>% 
+#   inner_join(rm_dist_pca) %>%
+#   bind_rows(res2_long)
+# 
+# rm(list = ls(pattern = "^rm_dist_"))
+```
+-->
+
+## RM relevant to given period and/or region
+
+```{r rm-period}
+# 0 = not used
+# 0.5 = somewhat used
+# 1 = used
+rm_b <- readxl::read_xlsx(here::here("analysis/data/raw_data",
+                                     "sidliste_vs_zdroje.xlsx"), sheet = "B")
+rm_m <- readxl::read_xlsx(here::here("analysis/data/raw_data",
+                                     "sidliste_vs_zdroje.xlsx"), sheet = "M")
+
+rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
+  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
+  pivot_longer(starts_with("p"), 
+               names_to = "period", values_to = "relevance") %>% 
+  rename(rm = surovina)
+
+rm_dist_rel_p1 <- set_base$period2 %>% 
+  filter_rel_dist_rm()
+
+rm_dist_rel_p2 <- set_base$period2 %>% 
+  filter_rel_dist_rm()
+```
+
+### RM relations in *general* pottery traditions
+
+#### Morava river catchment
+
+```{r rm-dist-m-p1, fig.cap="Distance to RM sources, Morava river catchment, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p1 %>% 
+  plot_rm_reg_dist("m")
+```
+
+#### Eastern part of Bohemia
+
+```{r rm-dist-b-p1, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery traditions. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p1 %>% 
+  plot_rm_reg_dist("b")
+```
+
+### RM relations in *detailed* pottery groups
+
+#### Morava river catchment
+
+```{r rm-dist-m-p2, fig.cap="Distance to RM sources, Morava river catchment, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p2 %>% 
+  plot_rm_reg_dist("m")
+```
+
+#### Eastern part of Bohemia
+
+```{r rm-dist-b-p2, fig.cap="Distance to RM sources, Eastern part of Bohemia, pottery groups. RM sources relevant in given period in black, somewhat relevant RM sources in gray."}
+rm_dist_rel_p2 %>% 
+  plot_rm_reg_dist("b")
+```
+
+```{r rm-dist-rel}
+rm_types <- rm_dist %>% 
+  select(rm, type) %>% 
+  distinct()
+
+rm_types_tab <- rm_types %>% 
+  pull(type)
+
+names(rm_types_tab) <- rm_types %>% 
+  pull(rm)
+
+res1_long <- rm_dist_rel_p1 %>% 
+  prep_dist_rel_rm() %>% 
   bind_rows(res1_long)
 
-res2_long <- set_base$period2 %>% 
-  select(id, period, chrono) %>% 
-  inner_join(rm_dist_median) %>%
+res2_long <- rm_dist_rel_p2 %>% 
+  prep_dist_rel_rm() %>% 
   bind_rows(res2_long)
@@
 
-```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
-rm_dist %>%
-  ggplot(aes(x = min, y = rm, fill = region)) +
-  geom_violin() +
-  scale_fill_brewer(palette = "Accent") +
-  facet_wrap(~type, scales = "free_y") +
-  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
-  theme_minimal() +
-  theme(legend.position = "bottom")
+```{r rm-dist-rel1, fig.cap="Mean distance to relevant raw materials in pottery traditions."}
+rm_dist_rel_p1 %>% 
+  count_dist_rel_rm() %>% 
+  plot_dist_rel_rm()
 ```
 
-```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
-kde1 %>% select(id, period_label, kde) %>% 
-  left_join(rm_dist) %>% 
-  mutate(region = str_extract(id, "^.")) %>% 
-  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
-         region == "M") %>% 
-  ggplot(aes(median, kde)) +
-  geom_point(alpha = 0.1) + 
-  stat_smooth(method = "lm", se = FALSE) +
-  facet_grid(vars(period_label), vars(rm)) +
-  labs(x = "Median distance to raw material source (km)",
-       y = "KDE") +
-  theme_minimal() +
-  theme(strip.text.y = element_text(angle = 0))
-```
-
-```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
-kde2 %>% select(id, period_label, kde) %>% 
-  left_join(rm_dist) %>% 
-  mutate(region = str_extract(id, "^.")) %>% 
-  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
-         region == "M") %>% 
-  ggplot(aes(median, kde)) +
-  geom_point(alpha = 0.1) + 
-  stat_smooth(method = "lm", se = FALSE) +
-  facet_grid(vars(period_label), vars(rm)) +
-  labs(x = "Median distance to raw material source (km)",
-       y = "KDE") +
-  theme_minimal() +
-  theme(strip.text.y = element_text(angle = 0))
+```{r rm-dist-rel2, fig.cap="Mean distance to relevant raw materials in pottery groups."}
+rm_dist_rel_p2 %>% 
+  count_dist_rel_rm() %>% 
+  plot_dist_rel_rm()
 ```
@@
 intensity of watercourses (and landscape in general) are a major factor
-imapiring such analysis.
+impairing such analysis.
 
@@
 ```{r get-raster}
-hydro_dens <- raster::raster(
-  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
-  raster::rasterToPoints() %>%
-  as_tibble() %>%
-  rename(KDE = density_raster)
+# pkg raster
+# hydro_dens <- raster::raster(
+#   here::here(derived_data, "hydro", "density_raster.tif"))
+# 
+# hydro_dens_at_points <- hydro_dens %>% 
+#   raster::extract(set_spat) %>% 
+#   as_tibble() %>% 
+#   rename(hydro_kde = value) %>% 
+#   bind_cols(select(st_drop_geometry(set_spat), id))
+
+# pkg stars
+hydro_dens <- stars::read_stars(
+  here::here(derived_data, "hydro", "density_raster.tif")
+) %>% 
+  st_set_crs(value = st_crs(set_spat))
+
+hydro_dens_at_points <- hydro_dens %>% 
+  stars::st_extract(set_spat) %>% 
+  st_drop_geometry() %>% 
+  rename(hydro_kde = density_raster.tif) %>% 
+  bind_cols(select(st_drop_geometry(set_spat), id))
+
+# bind to results
+res1_long <- set_base$period1 %>% 
+  prep_hydro_dens_res() %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-facet) %>% 
+  prep_hydro_dens_res() %>% 
+  bind_rows(res2_long)
 ```
 
-```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
+```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites (raster clipped to research area extent)."}
+# hydro_dens %>% 
+#   raster::mask(mask) %>% 
+#   raster::rasterToPoints() %>%
+#   as_tibble() %>%
+#   rename(KDE = density_raster) %>% 
+#   ggplot() +
+#   geom_raster(aes(x, y, fill = KDE)) +
+#   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
+#   theme_void()
+
 ggplot() +
-  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
+  stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
-  theme_void()
+  theme_void() +
+  labs(fill = "KDE")
 ```
+<!--- ====================================================================== -->
+
+# Terrain surrounding the settlements
+
+We focus on slope around the settlements.
+
+```{r dem-data}
+dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))
+
+# bind to results
+res1_long <- set_base$period1 %>% 
+  prep_dem_res() %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-facet) %>% 
+  prep_dem_res() %>% 
+  bind_rows(res2_long)
+```
+
+```{r altitude-to-res}
+altitude <- set_spat %>% st_drop_geometry() %>% 
+  as_tibble() %>% 
+  select(id, altitude)
+
+# bind to results
+res1_long <- set_base$period1 %>% 
+  prep_alt_res() %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>% 
+  select(-facet) %>% 
+  prep_alt_res() %>% 
+  bind_rows(res2_long)
+
+```
+
+```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
+alt_plot <- set_spat %>% st_drop_geometry() %>% 
+  as_tibble() %>% 
+  select(id, altitude) %>% 
+  left_join(dem_data)
+
+alt_cor <- alt_plot %>% 
+  summarise(cor = cor(altitude, dem)) %>% 
+  pull(cor) %>% 
+  round(2)
+
+alt_plot %>% 
+  ggplot(aes(altitude, dem)) +
+  geom_point() +
+  theme_minimal() +
+  annotate(geom = "text", x = 175, y = 525, 
+           label = paste("Cor. ", alt_cor))
+```
+
+<!--- ====================================================================== -->
+
+# Settlement 'hierarchy'
+
+Minimal distance to fortified settlements in given regions is explored.
+
+```{r dist-fenced-input}
+dist_fenced <- read_rds(here::here(derived_data, "fortif/distance_fortif.RDS"))
+
+# results 1
+res1_long <- dist_fenced$traditions %>% 
+  prep_dist_fenced() %>% 
+  bind_rows(res1_long)
+
+# results 2
+dist_fenced_p2 <- dist_fenced$groups %>% 
+  prep_dist_fenced() %>% 
+  bind_rows(res2_long)
+```
+
+<!--- ====================================================================== -->
+
+# Settlement 'continuity'
+
+Is the same location settled in the previous period?
+
+```{r continuity-calc}
+res1_long <- set_base$period1 %>%
+  settlement_continuity() %>% 
+  bind_rows(res1_long)
+
+res2_long <- set_base$period2 %>%
+  settlement_continuity() %>% 
+  bind_rows(res2_long)
+```
+
+```{r continuity-plot1, fig.cap="Settlements continuity, pottery traditions."}
+n_set_per1 <- set_base$period1 %>% 
+  add_region() %>% 
+  group_by(period_label, reg) %>% 
+  summarise(sum = n())
+
+res1_long %>% 
+  plot_continuity(y = n_set_per1)
+```
+
+```{r continuity-plot2, fig.cap="Settlements continuity, pottery groups."}
+n_set_per2 <- set_base$period2 %>% 
+  add_region() %>% 
+  group_by(period_label, reg) %>% 
+  summarise(sum = n())
+
+res2_long %>% 
+  plot_continuity(y = n_set_per2)
+```
+
+<!--- ====================================================================== -->
+
+# Arrangement along natural lines
+
+Arrangement of settlements along natural 'terrain' lines or along waterways is explored.
+
+```{r along_lines}
+along_lines <- read_rds(here::here(derived_data, "lines/linear_arrangement.RDS"))
+
+along_lines1 <- along_lines$traditions %>% 
+  mutate(chrono = str_remove(chrono, ".$")) %>% 
+  group_by(id, chrono, period, variable) %>% 
+  summarise(value = max(value)) %>% 
+  ungroup(id, chrono, variable) %>% 
+  mutate(value = value / n()) %>% 
+  select(-chrono) %>% 
+  group_by(variable, period, id) %>% 
+  summarise(value = mean(value), .groups = "drop")
+
+along_lines2 <- along_lines$groups %>% 
+  distinct(id, chrono, period, variable, value) %>% 
+  group_by(period) %>% 
+  mutate(value = value / n()) %>% 
+  select(-chrono) %>% 
+  group_by(variable, period, id) %>% 
+  summarise(value = mean(value), .groups = "drop")
+
+# results 1
+res1_long <- res1_long %>% 
+  bind_rows(along_lines1)
+
+# results 2
+res2_long <- res2_long %>% 
+  bind_rows(along_lines2)
+```
+
+```{r clean}
+rm(list = ls(pattern = "^rm"))
+rm(list = ls(pattern = "^along"))
+rm(list = ls(pattern = "line$"))
+rm(list = ls(pattern = "violin$"))
+rm(list = ls(pattern = "^n_set"))
+rm(list = ls(pattern = "^hydro"))
+rm(list = ls(pattern = "fenced$"))
+rm(list = ls(pattern = "dem_data"))
+rm(list = ls(pattern = "^alt"))
+```
+
+<!--- ====================================================================== -->
+
+# Synthesis
+
+```{r res-cleaning}
+res1_wide <- res1_long %>% prep_wide_res()
+# res1_long <- res1_wide %>% 
+#   pivot_longer(-c(id, period), names_to = "variable", values_to = "value")
+
+res2_wide <- res2_long %>% prep_wide_res()
+# res2_long <- res2_wide %>% 
+#   pivot_longer(-c(id, period), names_to = "variable", values_to = "value")
+
+# # scale cols
+# res1_wide <- res1_wide %>% 
+#   mutate(across(where(is.numeric), scale))
+# 
+# res2_wide <- res2_wide %>% 
+#   mutate(across(where(is.numeric), scale))
+```
+
+## Violin plots
+
+```{r variable-labels}
+labs_variables <- c(
+  "settlements_kde" = "Settlement density",
+  "dist_fenced" = "Min. dist. to fenced set. (km)",
+  "line_terrain" = "Along terrain line",
+  "line_water" = "Along water line",
+  "cont" = "Settlement continuity",
+  "hydro_kde" = "Watercourse density",
+  # "dem" = "Altitude (DEM)", 
+  "altitude" = "Altitude (map)",
+  "slope" = "Slope (°)",
+  "rm_dist_chipped" = "Distance to chipped RM",
+  "rm_dist_polished" = "Distance to polished RM"
+  # "rm_pc1" = "Distance to RM sources 1",
+  # "rm_pc2" = "Distance to RM sources 2",
+  # "rm_pc3" = "Distance to RM sources 3"
+) %>% 
+  map_chr(str_wrap, 12)
+```
+
+```{r syn-p1, fig.height=12}
+res1_violin <- res1_long %>% plot_violin()
+res1_violin
+```
+
+```{r syn-p2, fig.height=12}
+res2_violin <- res2_long %>% plot_violin()
+res2_violin
+```
+
+```{r res-violin-out}
+ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
+       width = 9, height = 12)
+ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
+       width = 9, height = 12)
+```
+
+## Line plots
+
+```{r syn-p1-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
+res1_line <- res1_long %>% plot_lines()
+res1_line
+```
+
+```{r syn-p2-sum, fig.cap="Summaries for different variables across studied periods. Solid line - mean value, wide transparent line - median value.", fig.height=12}
+res2_line <- res2_long %>% plot_lines()
+res2_line
+```
+
+```{r res-line-out}
+ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
+       width = 7, height = 7)
+ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
+       width = 7, height = 7)
+```
+
+<!-- ## PCA -->
+
+<!-- ### Pottery traditions -->
+
+<!-- ```{r pca1-all} -->
+<!-- # fit PCA -->
+<!-- pca1_all_fit <- res1_wide %>%  -->
+<!--   # select(-cont) %>%  -->
+<!--   select(where(is.numeric)) %>%  -->
+<!--   scale(center = TRUE, scale = TRUE) %>%  -->
+<!--   prcomp() -->
+
+<!-- # augmented fitted values -->
+<!-- pca1_all_aug <- pca1_all_fit %>%  -->
+<!--   broom::augment(res1_wide) %>%  -->
+<!--   select(-".rownames") %>%  -->
+<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
+<!--   settlements::add_region() %>%  -->
+<!--   settlements::add_period_label() %>%  -->
+<!--   settlements::add_neo() -->
+
+<!-- # rotation matrix -->
+<!-- pca1_all_rot <- pca1_all_fit %>% -->
+<!--   broom::tidy(matrix = "rotation") %>% -->
+<!--   mutate(PC = str_c("PC", PC)) %>% -->
+<!--   pivot_wider(names_from = PC) -->
+
+<!-- # var explained -->
+<!-- pca1_all_var_exp <- pca1_all_fit %>%  -->
+<!--   broom::augment(res1_wide) %>% -->
+<!--   summarise(across(contains("PC"), var)) %>%  -->
+<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
+<!--   mutate(var_exp = variance/sum(variance), -->
+<!--          cum_var_exp = cumsum(var_exp), -->
+<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
+<!--   select(pc, var_exp, cum_var_exp) -->
+
+<!-- # PC cum var explained over 0.9 -->
+<!-- pca1_all_var_exp_over <- pca1_all_var_exp %>%  -->
+<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
+<!--   pull(n) %>%  -->
+<!--   sum() + 1 -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-var-exp-tab} -->
+<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
+<!--                                              "Prop. of variance",  -->
+<!--                                              "Cumulative prop.")) -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-loadings-plot} -->
+<!-- # PCA loadings -->
+<!-- pca1_all_loadings <- pca1_all_fit %>%  -->
+<!--   broom::tidy(matrix = "rotation") %>%  -->
+<!--   mutate(component = str_c("PC", PC), -->
+<!--          component = fct_reorder(component, PC)) %>%  -->
+<!--   filter(PC <= pca1_all_var_exp_over) %>%  -->
+<!--   group_by(PC) %>%  -->
+<!--   top_n(8, abs(value)) %>%  -->
+<!--   ungroup() %>%  -->
+<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
+<!--          column = tidytext::reorder_within(column, abs(value), component)) -->
+
+<!-- pca1_all_loadings %>%  -->
+<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
+<!--   geom_col(color = "black") + -->
+<!--   scale_fill_manual(values = c("white", "gray80")) + -->
+<!--   facet_wrap(vars(component), scales = "free_y") + -->
+<!--   tidytext::scale_y_reordered() + -->
+<!--   theme_bw() + -->
+<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
+<!-- ``` -->
+
+<!-- ```{r plot-pca1-results} -->
+<!-- plot_pca1_full <- function(x = "PC1", y = "PC2") { -->
+
+<!--   pcs <- c(x, y) -->
+
+<!--   pca1_all_plot <- pca1_all_aug %>%  -->
+<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
+<!--     geom_point(data = select(pca1_all_aug, -period_label),  -->
+<!--                alpha = 0.05, size = 0.4) + -->
+<!--     geom_point(aes(color = reg),  -->
+<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
+<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
+<!--     facet_grid(vars(period_label), vars(reg)) + -->
+<!--     theme_minimal() -->
+
+<!--   pca1_all_rot <-  pca1_all_rot %>% -->
+<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
+<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
+<!--     geom_text( -->
+<!--       aes(label = column), -->
+<!--       hjust = 1, nudge_x = -0.02, -->
+<!--       color = "#904C2F", -->
+<!--       size = 2.6 -->
+<!--     ) + -->
+<!--     theme_minimal() -->
+
+<!--   pca1_all_load <- pca1_all_loadings %>%  -->
+<!--     filter(component %in% pcs) %>%  -->
+<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
+<!--     geom_col(color = "black", show.legend = FALSE) + -->
+<!--     scale_fill_manual(values = c("white", "gray80")) + -->
+<!--     facet_wrap(vars(component), scales = "free_y") + -->
+<!--     tidytext::scale_y_reordered() + -->
+<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
+<!--     theme_minimal() -->
+
+<!--   pca1_all_plot / (pca1_all_rot | pca1_all_load) +  -->
+<!--     plot_layout(heights = c(6, 1)) -->
+<!-- } -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-plots-pc1-pc2} -->
+<!-- plot_pca1_full() -->
+<!-- ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-plots-pc3-pc4} -->
+<!-- plot_pca1_full("PC3", "PC4") -->
+<!-- ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-plots-pc5-pc6} -->
+<!-- plot_pca1_full("PC5", "PC6") -->
+<!-- ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca1-all-plots-pc7-pc8} -->
+<!-- plot_pca1_full("PC7", "PC8") -->
+<!-- ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ### Pottery groups -->
+
+<!-- ```{r pca2-all} -->
+<!-- # fit PCA -->
+<!-- pca2_all_fit <- res2_wide %>%  -->
+<!--   # select(-cont) %>%  -->
+<!--   select(where(is.numeric)) %>%  -->
+<!--   scale(center = TRUE, scale = TRUE) %>%  -->
+<!--   prcomp() -->
+
+<!-- # augmented fitted values -->
+<!-- pca2_all_aug <- pca2_all_fit %>%  -->
+<!--   broom::augment(res2_wide) %>%  -->
+<!--   select(-".rownames") %>%  -->
+<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
+<!--   settlements::add_region() %>%  -->
+<!--   settlements::add_period_label() %>%  -->
+<!--   settlements::add_neo() -->
+
+<!-- # rotation matrix -->
+<!-- pca2_all_rot <- pca2_all_fit %>% -->
+<!--   broom::tidy(matrix = "rotation") %>% -->
+<!--   mutate(PC = str_c("PC", PC)) %>% -->
+<!--   pivot_wider(names_from = PC) -->
+
+<!-- # var explained -->
+<!-- pca2_all_var_exp <- pca2_all_fit %>%  -->
+<!--   broom::augment(res2_wide) %>% -->
+<!--   summarise(across(contains("PC"), var)) %>%  -->
+<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
+<!--   mutate(var_exp = variance/sum(variance), -->
+<!--          cum_var_exp = cumsum(var_exp), -->
+<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
+<!--   select(pc, var_exp, cum_var_exp) -->
+
+<!-- # PC cum var explained over 0.9 -->
+<!-- pca2_all_var_exp_over <- pca2_all_var_exp %>%  -->
+<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
+<!--   pull(n) %>%  -->
+<!--   sum() + 1 -->
+<!-- ``` -->
+
+<!-- <!-- -->
+<!-- ```{r} -->
+<!-- fuu <- res2_wide %>% mutate(across(where(is.numeric), \(x) x == 0)) %>%  -->
+<!-- select(where(is.logical)) %>%  -->
+<!-- rowSums() %>%  -->
+<!-- as_tibble() -->
+
+<!-- shit <- res2_wide %>% select(id, period) %>% bind_cols(fuu) %>% arrange(desc(value), id, period) %>%  -->
+<!-- filter(value == 8) %>% pull(id) %>% factor() %>% levels() -->
+
+<!-- settlements2 %>%  -->
+<!-- filter(id %in% shit) %>%  -->
+<!-- sf::st_transform(4326) %>%  -->
+<!-- leaflet::leaflet() %>%  -->
+<!-- leaflet::addTiles() %>%  -->
+<!-- leaflet::addMarkers() -->
+
+<!-- res2_wide %>% filter(id %in% shit) %>% View() -->
+<!-- ``` -->
+<!-- -->
+
+<!-- ```{r pca2-all-var-exp-tab} -->
+<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
+<!--                                              "Prop. of variance",  -->
+<!--                                              "Cumulative prop.")) -->
+<!-- ``` -->
+
+<!-- ```{r pca2-all-loadings-plot} -->
+<!-- # PCA loadings -->
+<!-- pca2_all_loadings <- pca2_all_fit %>%  -->
+<!--   broom::tidy(matrix = "rotation") %>%  -->
+<!--   mutate(component = str_c("PC", PC), -->
+<!--          component = fct_reorder(component, PC)) %>%  -->
+<!--   filter(PC <= pca2_all_var_exp_over) %>%  -->
+<!--   group_by(PC) %>%  -->
+<!--   top_n(8, abs(value)) %>%  -->
+<!--   ungroup() %>%  -->
+<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
+<!--          column = tidytext::reorder_within(column, abs(value), component)) -->
+
+<!-- pca2_all_loadings %>%  -->
+<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
+<!--   geom_col(color = "black") + -->
+<!--   scale_fill_manual(values = c("white", "gray80")) + -->
+<!--   facet_wrap(vars(component), scales = "free_y") + -->
+<!--   tidytext::scale_y_reordered() + -->
+<!--   theme_bw() + -->
+<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
+<!-- ``` -->
+
+<!-- ```{r plot-pca2-results} -->
+<!-- plot_pca2_full <- function(x = "PC1", y = "PC2") { -->
+
+<!--   pcs <- c(x, y) -->
+
+<!--   pca2_all_plot <- pca2_all_aug %>%  -->
+<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
+<!--     geom_point(data = select(pca2_all_aug, -period_label),  -->
+<!--                alpha = 0.05, size = 0.4) + -->
+<!--     geom_point(aes(color = reg),  -->
+<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
+<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
+<!--     facet_grid(vars(period_label), vars(reg)) + -->
+<!--     theme_minimal() -->
+
+<!--   pca2_all_rot <-  pca2_all_rot %>% -->
+<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
+<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
+<!--     geom_text( -->
+<!--       aes(label = column), -->
+<!--       hjust = 1, nudge_x = -0.02, -->
+<!--       color = "#904C2F", -->
+<!--       size = 2.6 -->
+<!--     ) + -->
+<!--     theme_minimal() -->
+
+<!--   pca2_all_load <- pca2_all_loadings %>%  -->
+<!--     filter(component %in% pcs) %>%  -->
+<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
+<!--     geom_col(color = "black", show.legend = FALSE) + -->
+<!--     scale_fill_manual(values = c("white", "gray80")) + -->
+<!--     facet_wrap(vars(component), scales = "free_y") + -->
+<!--     tidytext::scale_y_reordered() + -->
+<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
+<!--     theme_minimal() -->
+
+<!--   pca2_all_plot / (pca2_all_rot | pca2_all_load) +  -->
+<!--     plot_layout(heights = c(6, 1)) -->
+<!-- } -->
+<!-- ``` -->
+
+<!-- ```{r pca2-all-plots-pc1-pc2} -->
+<!-- plot_pca2_full() -->
+<!-- ggsave(here::here("analysis/figures/pca2_full_pc12.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca2-all-plots-pc3-pc4} -->
+<!-- plot_pca2_full("PC3", "PC4") -->
+<!-- ggsave(here::here("analysis/figures/pca2_full_pc34.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca2-all-plots-pc5-pc6} -->
+<!-- plot_pca2_full("PC5", "PC6") -->
+<!-- ggsave(here::here("analysis/figures/pca2_full_pc56.png"), height = 14) -->
+<!-- ``` -->
+
+<!-- ```{r pca2-all-plots-pc7-pc8} -->
+<!-- plot_pca2_full("PC7", "PC8") -->
+<!-- ggsave(here::here("analysis/figures/pca2_full_pc78.png"), height = 14) -->
+<!-- ``` -->
+
+## Principal components analysis
+
+```{r nested-pca}
+# traditions
+# traditions, by periods
+pca1_per <- res1_wide %>%
+  add_region() %>% 
+  group_by(reg, period) %>% 
+  nest() %>% 
+  nested_pca()
+
+# traditions, by neo BC
+pca1_neo <- res1_wide %>%
+  add_region() %>% 
+  add_neo() %>% 
+  select(-period) %>% 
+  group_by(reg, neo) %>% 
+  nest() %>% 
+  nested_pca()
+
+# groups
+# groups, by periods
+pca2_per <- res2_wide %>%
+  add_region() %>% 
+  group_by(reg, period) %>% 
+  nest() %>% 
+  nested_pca()
+
+# groups, by neo BC
+pca2_neo <- res2_wide %>%
+  add_region() %>% 
+  add_neo() %>% 
+  select(-period) %>% 
+  group_by(reg, neo) %>% 
+  nest() %>% 
+  nested_pca()
+
+# pca1_neo %>%
+#   select(reg, neo, sdev) %>%
+#   unnest(sdev) %>%
+#   ungroup(neo, reg) %>%
+#   ggplot(aes(factor(pc), cum_prop)) +
+#   geom_point() +
+#   geom_path(aes(group = neo)) +
+#   geom_hline(yintercept = 0.9) +
+#   facet_grid(vars(neo), vars(reg)) +
+#   scale_y_continuous(limits = c(0, 1)) +
+#   theme_minimal()
+```
+
+Variable labels in PCA plots:
+
+altitude - alt.,
+cont - cont.,
+dist_fenced - d. fenced,
+hydro_kde - hydro.,
+line_terrain -  t. line,
+settlements_kde - dens.,
+slope ~ slope,
+rm_dist_chipped - d. ch.,
+line_water - w. line,
+rm_dist_polished - d. pol.
+
+### Pottery traditions in Neolithic B and C
+
+```{r plot-pca1-neo1, fig.cap="PCA for pottery traditions data in Neolithic B and C, PC 1-2 space."}
+pca1_neo %>% plot_nested_pca("neo")
+```
+
+```{r plot-pca1-neo2, fig.cap="PCA for pottery traditions data in Neolithic B and C, PC 3-4 space."}
+pca1_neo %>% plot_nested_pca("neo", c("PC3", "PC4"))
+```
+
+### Pottery groups in Neolithic B vs Neolithic C
+
+```{r plot-pca2-neo1, fig.cap="PCA for pottery groups data in Neolithic B and C, PC 1-2 space."}
+pca2_neo %>% plot_nested_pca("neo")
+```
+
+```{r plot-pca2-neo2, fig.cap="PCA for pottery groups data in Neolithic B and C, PC 3-4 space."}
+pca2_neo %>% plot_nested_pca("neo", c("PC3", "PC4"))
+```
+
+### Pottery traditions across periods
+
+```{r plot-pca1-period1, fig.cap="PCA for pottery traditions data across periods, PC 1-2 space."}
+pca1_per %>% 
+  add_period_label() %>% 
+  plot_nested_pca("period_label")
+```
+
+```{r plot-pca1-period2, fig.cap="PCA for pottery traditions data across periods, PC 3-4 space."}
+pca1_per %>% 
+  add_period_label() %>% 
+  plot_nested_pca("period_label", c("PC3", "PC4"))
+```
+
+### Pottery groups across periods
+
+```{r plot-pca2-period1, fig.cap="PCA for pottery groups data across periods, PC 1-2 space."}
+pca2_per %>% 
+  add_period_label() %>% 
+  plot_nested_pca("period_label")
+```
+
+```{r plot-pca2-period2, fig.cap="PCA for pottery groups data across periods, PC 3-4 space."}
+pca2_per %>% 
+  add_period_label() %>% 
+  plot_nested_pca("period_label", c("PC3", "PC4"))
+```
+
+## Variables clustering
+
+### Pottery traditions
+
+```{r pca1-neo-hclust, fig.cap="Variables clustering dendrogram for pottery traditions."}
+hclust1_neo <- pca1_neo %>% 
+  hclust_variables()
+
+hclust_plot_dendro(hclust1_neo, c("reg", "neo"))
+```
+
+### Pottery groups
+
+```{r pca2-neo-hclust, fig.cap="Variables clustering dendrogram for pottery groups."}
+hclust2_neo <- pca2_neo %>% 
+  hclust_variables()
+
+hclust_plot_dendro(hclust2_neo, c("reg", "neo"))
+```
+
+<!--- ====================================================================== -->
 
@@
 res1_long %>%
+  distinct(id, period, variable, .keep_all = TRUE) %>% 
   pivot_wider(names_from = variable, values_from = value) %>% 
-  arrange(id, period, chrono) %>% 
+  arrange(id, period) %>% 
   write_csv(here::here(derived_data, "results", "results1.csv"))
@@
 res2_long %>% 
+  distinct(id, period, variable, .keep_all = TRUE) %>% 
   pivot_wider(names_from = variable, values_from = value) %>% 
-  arrange(id, period, chrono) %>% 
+  arrange(id, period) %>% 
   write_csv(here::here(derived_data, "results", "results2.csv"))
@@
 
-<!-- The following line inserts a page break -->
+<!--- ====================================================================== -->
 

### 2021-09-16 analysis/paper/paper.Rmd
model based clustering

@@
 
+## Groups in the data
+
+This is not done for pottery traditions due to an error in one of the periods/
+regions that is caused by not enough variance in the PCA data. Weird...
+
+### Pottery groups
+
+```{r mclust2-neo}
+mb_clust2 <- pca2_neo %>% mbc_cluster()
+
+mb_clust2 %>% pull(classif)
+
+mb_clust2 %>%
+    select(classif) %>% 
+    ungroup(c(reg, neo)) %>%
+  unnest() %>% View()
+    mutate(g = str_c(reg, g))
+
+mb_classif2_neoB <- mb_clust2 %>% mbc_classify("NeoB")
+mb_classif2_neoC <- mb_clust2 %>% mbc_classify("NeoC")
+
+(mb_classif2_neoB == mb_classif2_neoC) %>% all()
+
+# bind_cols(mb_clustering_neo2$x[[2]][, 1:2], mb_clustering_neo2$classif[[2]]) %>% 
+#   ggplot(aes(PC1, PC2, color = factor(g))) +
+#   stat_ellipse() +
+#   geom_point() +
+#   theme_bw() +
+#   facet_wrap(vars(g))
+
+mb_map2_neoB <- mb_classif2 %>% mbc_map(sf = settlements2, neo = "NeoB")
+mb_map2_neoC <- mb_classif2 %>% mbc_map(sf = settlements2, neo = "NeoC")
+
+# boxploty pro proměnné ve skupinách...
+```
+
+```{r mclust2-map-neoB, fig.cap="Clusters in pottery groups in Neolithic B."}
+mb_map2_neoB
+```
+
+```{r mclust2-map-neoC, fig.cap="Clusters in pottery groups in Neolithic C."}
+mb_map2_neoC
+```
+
+```{r}
+res2_long %>% 
+  filter(variable != "cont") %>% 
+  add_neo() %>% 
+  right_join(mb_classif2, by = c("id", "neo")) %>% 
+  filter(neo == "NeoC") %>% 
+  ggplot(aes(g, value, color = g))+
+  geom_boxplot(show.legend = FALSE) +
+  facet_wrap(vars(variable), scales = "free_y", ncol = 2) +
+  theme_minimal()
+```
+
 <!--- ====================================================================== -->

### 2021-09-16 R/clustering.R

@@ -1,0 +1,80 @@
+# model-based clustering
+
+#' Model-based clustering
+#'
+#' @param x A nested PCA result.
+#' @param prob A threshold of cumulative variability explained.
+#'     Only PCs that cumulatively explain at least the given value are kept.
+#'     Defaults to 0.75.
+#'
+#' @return A nested \code{tibble} with \code{mclust} objects and classification
+#'     for individual points.
+#' @export
+#'
+#' @examples
+mbc_cluster <- function(x, prob = 0.75) {
+  mb_clustering_neo2 <- x %>%
+    mutate(
+      # select only PCs that cumulatively explain given variability
+      overProb = purrr::map(sdev, \(x) x[(cumsum(x[, "var_prop"]) >= prob), ]$pc[1]),
+      overProb = purrr::map(overProb, \(x) 1:x),
+      # select PCs
+      x2 = purrr::map2(x, overProb, \(x, y) x[, c(y)]),
+      # model-based clustering
+      bic = purrr::map(x2, \(x) mclust::mclustBIC(x)),
+      bic_sum = purrr::map(bic, \(x) mclust::summary.mclustBIC(x)),
+      # select best model based on BIC
+      bic_best = purrr::map(bic_sum, \(x) names(x[1])),
+      bic_best = purrr::map(bic_best, \(x) unlist(strsplit(x, ","))),
+      mclust = purrr::pmap(list(x2, bic_best, bic),
+                           \(data, model, bic)
+                           mclust::Mclust(data = data,
+                                          G = as.numeric(model[[2]]),
+                                          modelNames = model[[1]],
+                                          x = bic)),
+      classif = purrr::map(mclust, \(x) x$classification),
+      classif = purrr::map2(id, classif, \(x, y) bind_cols(id = x, g = y))
+    )
+}
+
+#' Classification based on model-based clustering
+#'
+#' @param x Nested object, result of \code{mbc_cluster}.
+#'
+#' @return
+#' @export
+#'
+#' @examples
+mbc_classify <- function(x, neo) {
+  x %>%
+    select(classif) %>%
+    filter(neo == neo) %>%
+    unnest() %>%
+    ungroup(c(reg, neo)) %>%
+    mutate(g = str_c(reg, g))
+}
+
+
+#' Plot a map of groups resulting from model based clustering
+#'
[…]

### 2022-03-23 R/clustering.R
clean functions

@@ -47,11 +47,11 @@ mbc_cluster <- function(x, prob = 0.75) {
 #' @examples
 mbc_classify <- function(x, neo) {
   x %>%
-    select(classif) %>%
-    filter(neo == neo) %>%
-    unnest() %>%
-    ungroup(c(reg, neo)) %>%
-    mutate(g = str_c(reg, g))
+    dplyr::select(classif) %>%
+    dplyr::filter(neo == neo) %>%
+    tidyr::unnest() %>%
+    dplyr::ungroup(c(reg, neo)) %>%
+    dplyr::mutate(g = str_c(reg, g))
 }
 
 

### 2022-03-23 R/dem.R

@@ -1,0 +1,49 @@
+#' Extract values from terrain model derived from DEM.
+#'
+#' @param x Object of class \code{sf} or similar.
+#' @param dem Digital elevation model.
+#' @param model What model to use: slope, aspect, TPI, TRI...
+#' @param fun Summary function.
+#'
+#' @return Object x with result in column \code{value}.
+#' @export
+#'
+#' @examples
+extract_from_dem <- function(x, dem, model = "slope", fun = "mean") {
+
+  # create model from dem
+  message("Creating terrain model.")
+  terrain_model <- terra::terrain(x = dem, v = model)
+
+  # check type of geom
+  geom_type <- terra::geomtype(terra::vect(x))
+  if (!(geom_type %in% c("points", "polygons"))) {
+    stop("x must contain points or polygons.")
+  }
+
+  # transform x
+  if (terra::crs(terra::vect(x)) != terra::crs(dem)) {
+    x <- sf::st_transform(x, terra::crs(dem))
+    message("Transforming x to CRS of dem.")
+  }
+
+  # slope in the vicinity of settlements
+  message("Extracting values.")
+  model2values <- terra::extract(terrain_model,
+                                 terra::vect(x))
+  cols <- colnames(model2values)
+
+  # return values
+  if (geom_type == "points") {
+    res <- model2values |>
+      dplyr::bind_cols(x) |>
+      dplyr::select(-!!dplyr::sym(cols[1]))
+  } else if (geom_type == "polygons") {
+    res <- model2values |>
+      dplyr::group_by(!!dplyr::sym(cols[1])) |>
+      dplyr::summarise(value = match.fun(fun)(!!dplyr::sym(cols[2]))) |>
+      dplyr::bind_cols(x) |>
+      dplyr::select(-!!dplyr::sym(cols[1]))
+  }
+  return(res)
+}

### 2022-03-23 R/pca.R

@@ -9,26 +9,26 @@
 nested_pca <- function(x) {
   x %>%
     mutate(
-      id = map(data, \(x) select(x, id)),
-      data = map(data, \(x) select(x, -id)),
-      mx = map(data, \(x) complete(x)),
-      mx = map(mx, \(x) select(x, where(~ sum(.x) != 0))),
-      mx = map(mx, \(x) select(x, where(~ sum(.x)/length(.x) != 1))),
-      mx = map(mx, \(x) mutate(x, across(where(is.numeric), scale))),
-      mx = map(mx, as.matrix),
-      pca = map(mx, prcomp),
-      sdev = map(pca, \(x) tibble(sdev = x$sdev)),
-      sdev = map(sdev, \(x) mutate(x,
-                                   var_prop = sdev^2 / sum(sdev^2),
-                                   cum_prop = cumsum(var_prop),
-                                   pc = row_number())),
-      x = map(pca, \(x) as_tibble(x$x)),
-      rotation = map(pca, broom::tidy, matrix = "rotation"),
-      rotation = map(rotation, pivot_wider,
-                     names_from = "PC", names_prefix = "PC",
-                     values_from = "value")
+      id = purrr::map(data, \(x) dplyr::select(x, id)),
+      data = purrr::map(data, \(x) dplyr::select(x, -id)),
+      mx = purrr::map(data, \(x) complete(x)),
+      mx = purrr::map(mx, \(x) dplyr::select(x, where(~ sum(.x) != 0))),
+      mx = purrr::map(mx, \(x) dplyr::select(x, where(~ sum(.x)/length(.x) != 1))),
+      mx = purrr::map(mx, \(x) dplyr::mutate(x, across(where(is.numeric), scale))),
+      mx = purrr::map(mx, as.matrix),
+      pca = purrr::map(mx, prcomp),
+      sdev = purrr::map(pca, \(x) tibble(sdev = x$sdev)),
+      sdev = purrr::map(sdev, \(x) dplyr::mutate(x,
+                                                 var_prop = sdev^2 / sum(sdev^2),
+                                                 cum_prop = cumsum(var_prop),
+                                                 pc = row_number())),
+      x = purrr::map(pca, \(x) as_tibble(x$x)),
+      rotation = purrr::map(pca, broom::tidy, matrix = "rotation"),
[…]

### 2022-03-23 analysis/code/analysis_dem.R
edit code

@@ -1,218 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# ANALYSIS DEM
-# author: Petr Pajdla
-# Getting values from DEM
-
-library(tidyverse)
-library(sf)
-library(stars)
-library(starsExtra)
-
-
-# data --------------------------------------------------------------------
-
-derived_data <- "analysis/data/derived_data"
-
-path_der <- here::here(derived_data, "dem/")
-path_maps <- here::here(derived_data, "maps/")
-
-# mask
-mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
-bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)
-# bboxes_m <- st_read(paste0(path_der, "bboxes_morava.geojson"))
-
-# dem
-# dem <- read_stars(paste0(path_der, "dem_aster_sjtsk.tif")) %>%
-#   st_set_crs(st_crs(bbox))
-# write_stars(dem, paste0(path_der, "dem_aster_sjtsk_stars.tif"))
-dem <- read_stars(paste0(path_der, "dem_aster_sjtsk_stars.tif"), proxy = FALSE)
-
-# settlements
-set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
-                    quiet = TRUE)
-
-# reduce dem ---------------------------------------------------------------
-
-# bbox_grid <- st_make_grid(bbox, cellsize = 5000)
-#
-# plot(bbox_grid)
-# plot(st_geometry(mask), add = TRUE)
-#
-# dem_grid <- aggregate(dem, bbox_grid, mean, na.rm = TRUE)
-#
-# dem_grid_sf <- st_as_sf(dem_grid)
-
-# st_write(dem_grid_sf, paste0(path_der, "dem_aster_sjtsk_sf.geojson"))
-dem_sf <- st_read(paste0(path_der, "dem_aster_sjtsk_sf.geojson")) %>%
-  rename(value = dem_aster_sjtsk.tif)
-
-
-# B -----------------------------------------------------------------------
-
-# bbox
-b_bbox <- set_spat %>%
-  filter(str_detect(id, "^B")) %>%
-  st_buffer(dist = 5e3) %>%
-  st_union() %>%
-  st_bbox()
-
-# crop
-b_dem <- dem[b_bbox]
-
-# slope
-b_slope <- slope(b_dem)
-
[…]

### 2022-03-23 analysis/code/chrono_plot.R

@@ -1,49 +1,0 @@
-# Project "Settlements"
-# Script nr. 0
-# PLOT SETTLEMENTS
-# author: Petr Pajdla
-# Plot settlement distribution in different periods
-
-library(tidyverse)
-library(here)
-
-dt_raw <- "analysis/data/raw_data"
-dt_der <- "analysis/data/derived_data"
-
-chrono <- read_csv(here(dt_raw, "chronotable.csv"))
-set <- read_rds(here(dt_der, "settlements.RDS"))
-chrono_lvls <- read_rds(here(dt_der, "chrono_labels.RDS"))
-
-chrono %>%
-  filter(chrono_lvl == 2) %>%
-  mutate(chrono = factor(chrono, levels = chrono_lvls$chrono2)) %>%
-  ggplot(aes(y = chrono)) +
-  geom_point(aes(x = from, color = chrono)) +
-  geom_point(aes(x = to, color = chrono)) +
-  geom_linerange(aes(xmin = from, xmax = to, color = chrono)) +
-  facet_wrap(~region, nrow = 2, scales = "free_y") +
-  theme_minimal()
-
-chrono_yvals <- tribble(
-  ~region, ~chrono, ~ymin, ~ymax,
-  "nw", "SBK", "SBK Early", "SBK Late",
-  "nw", "LgK", "LgK I", "Epi LgK",
-  "nw", "Early Ene.", "Jordanow", "Michelsberg",
-  "nw", "TRB", "TRB Early", "TRB Boleraz/Saalzmund",
-)
-
-chrono_rects <- chrono %>% filter(chrono_lvl == 1) %>%
-  left_join(chrono_yvals) %>%
-  mutate(
-    # chrono = factor(chrono, levels = chrono_lvls$chrono1),
-    ymin = factor(ymin, levels = chrono_lvls$chrono2),
-    ymax = factor(ymax, levels = chrono_lvls$chrono2))
-
-
-
-geom_point(data = filter(chrono, chrono_lvl == 2),
-           aes(x = from, y = chrono, color = chrono)) +
-  geom_point(data = filter(chrono, chrono_lvl == 2),
-             aes(x = to, y = chrono, color = chrono)) +
-  geom_linerange(data = filter(chrono, chrono_lvl == 2),
-                 aes(xmin = from, xmax = to, y = chrono, color = chrono))

### 2022-03-23 analysis/code/data_load.R

@@ -1,0 +1,49 @@
+# Project "Settlements"
+# LOADS DATA
+# author: Petr Pajdla
+# This script loads basic data when sourced
+
+# input data path
+input_data <- "analysis/data/input_data"
+
+# derived data paths
+derived_data <- "analysis/data/derived_data"
+path_data <- here::here(derived_data)
+path_maps <- here::here(derived_data, "maps/")
+path_dem <- here::here(derived_data, "dem/")
+path_temp <- here::here(derived_data, "temp/")
+
+# mask
+mask <- sf::st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+bbox <- sf::st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)
+
+# rivers + admin for plots
+admin0 <- sf::st_read(paste0(path_maps, "admin0.geojson"), quiet = TRUE)
+admin1 <- sf::st_read(paste0(path_maps, "admin1.geojson"), quiet = TRUE)
+rivers0 <- sf::st_read(paste0(path_maps, "rivers0.geojson"), quiet = TRUE)
+rivers1 <- sf::st_read(paste0(path_maps, "rivers1.geojson"), quiet = TRUE)
+
+# labels
+labs_chrono <- readr::read_rds(here::here(input_data, "chrono_labels.RDS"))
+
+# database
+set_base <- readr::read_rds(here::here(input_data, "settlements.RDS"))
+
+# spatial data
+set_spat <- sf::st_read(here::here(input_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+# neolithic B/C (period0)
+
+# pottery traditions (period1)
+settlements1 <- dplyr::right_join(set_spat, set_base$period1, by = "id")
+settlements1 <- dplyr::mutate(settlements1, label = labs_chrono$chrono1[chrono])
+
+# pottery groups (period2)
+settlements2 <- dplyr::right_join(set_spat, set_base$period2, by = "id")
+settlements2 <- dplyr::mutate(settlements2, label = labs_chrono$chrono2[chrono])
+
+# colors
+mycol <- c("B" = "#7FC97F", "M" = "#BEAED4", "b" = "#7FC97F", "m" = "#BEAED4")
+
+rm(list = ls(pattern = "path"))

### 2022-03-23 analysis/code/data_prep.R

@@ -1,8 +1,8 @@
 # Project "Settlements"
 # Script nr. 0
-# DATA PREPARATION II
+# DATA PREPARATION
 # author: Petr Pajdla
-# Downloads updated data from GD
+# Downloads updated data from GD and prepares input files
 
 library(here)
 library(tidyverse)
@@ -10,31 +10,28 @@ library(sf)
 library(googledrive)
 
 # paths
+dt_in <- "analysis/data/input_data"
 dt_der <- "analysis/data/derived_data"
 gd_path <- "~/settlements/data/database/"
 
-# dir.create(here(dt_der, "input_gd"))
+# dir.create(here(dt_in))
+# dir.create(here(dt_in, "gd"))
 
 # get updated data from drive
 files_gd <- drive_ls(gd_path)
 
 for (i in seq_along(files_gd$id)) {
   drive_download(as_id(files_gd[i, ]),
-                 here(dt_der, "input_gd", files_gd$name)[i],
+                 here(dt_in, "gd", files_gd$name)[i],
                  type = "csv", overwrite = TRUE)
 }
 
 
 # prepare derived data ----------------------------------------------------
-labs_chrono <- read_rds(here::here(dt_der, "chrono_labels.RDS"))
-
-# labs_chrono$periods["p5.0"] <- "4.9 – 4.8 k"
-# labs_chrono$periods["p3.4"] <- "3.4 – 3.3 k"
-#
-# write_rds(labs_chrono, here::here(dt_der, "chrono_labels.RDS"))
+labs_chrono <- read_rds(here::here(dt_in, "chrono_labels.RDS"))
 
 # list downloaded files
-files_disk <- list.files(here(dt_der, "input_gd"), full.names = TRUE)
+files_disk <- list.files(here(dt_in, "gd"), full.names = TRUE)
 
 # prepare input files for analysis
 set_base <- vector("list")
@@ -84,9 +81,9 @@ set_base$references <- read_csv(str_subset(files_disk, "references")) %>%
 read_csv(str_subset(files_disk, "spatial")) %>%
   st_as_sf(coords = c("X", "Y")) %>%
   st_set_crs(5514) %>%
-  st_write(here(dt_der, "settlements_sf.geojson"), delete_dsn = TRUE)
+  st_write(here(dt_in, "settlements_sf.geojson"), delete_dsn = TRUE)
 
-write_rds(set_base, here(dt_der, "settlements.RDS"))
[…]

### 2022-03-23 analysis/code/dataprep_dem.R

@@ -1,58 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# DEM PREP
-# author: Petr Pajdla
-# Preparation of rasters for slope calculation
-
-library(sf)
-library(raster)
-
-
-# data --------------------------------------------------------------------
-
-raw_data <- "analysis/data/raw_data"
-derived_data <- "analysis/data/derived_data"
-
-path_raw <- here::here(raw_data, "dem/")
-path_der <- here::here(derived_data, "dem/")
-path_maps <- here::here(derived_data, "maps/")
-
-if (!dir.exists(path_der)) {
-  dir.create(path_der)
-}
-
-# mask
-mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
-bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)
-
-# dem
-dem_aster <- raster(paste0(path_raw, "DEM_digital_elevation.tif"))
-
-# crop
-dem_aster_cropped <- dem_aster %>%
-  crop(st_transform(bbox, st_crs(dem_aster)))
-
-# reproject to sjtsk
-dem_aster_sjtsk <- dem_aster_cropped %>%
-  projectRaster(crs = crs(bbox))
-
-dem_aster_sjtsk_cropped <- dem_aster_sjtsk %>%
-  crop(bbox)
-
-# write raster for analysis
-writeRaster(dem_aster_sjtsk_cropped,
-            filename = paste0(path_der, "dem_aster_sjtsk.tif"),
-            overwrite = TRUE)
-
-
-# exploratory plots -------------------------------------------------------
-
-# plot(dem_aster_sjtsk_cropped)
-# plot(mask, col = NA, add = TRUE)
-#
-# res(dem_aster_sjtsk_cropped)
-#
-# plot(dem_aster_cropped)
-# plot(dem_aster_sjtsk)
-# plot(dem_aster_sjtsk_cropped)
-

### 2022-03-23 analysis/code/dataprep_hydro.R

@@ -1,105 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# WATERCOURSE DENSITY
-# author: Petr Pajdla
-# Joins neccessary layers and calculates watercourse intensity (KDE)
-
-library(sf)
-library(here)
-library(tidyverse)
-library(spatstat)
-library(raster)
-
-# read data
-dt_hydro <- "analysis/data/raw_data/hydro/"
-
-# hydrology
-cz <- st_read(here(dt_hydro, "hydro_czechia"))
-
-au <- st_read(here(dt_hydro, "hydro_austria"))
-
-sk <- st_read(here(dt_hydro, "hydro_slovakia"))
-
-# project mask
-mask_orig <- st_read(here("analysis/data/derived_data/maps/", "mask.geojson"))
-
-mask <- st_buffer(mask_orig, 20000) %>% st_simplify()
-
-plot(st_geometry(mask))
-plot(st_geometry(mask_orig), add = TRUE)
-
-# transform to S-JTSK and crop to mask
-au_crop <- au %>%
-  st_zm() %>%
-  st_transform(st_crs(mask)) %>%
-  st_intersection(mask)
-
-sk_crop <- sk %>%
-  st_transform(st_crs(mask)) %>%
-  st_intersection(mask)
-
-cz_crop <- cz %>%
-  st_intersection(mask)
-
-rm(list = c("cz", "au", "sk"))
-
-# control plot
-ggplot() +
-  geom_sf(data = au_crop, color = "green") +
-  geom_sf(data = cz_crop, color = "blue") +
-  geom_sf(data = sk_crop, color = "orange") +
-  theme_void()
-
-# join layers?
-sk_geom <- st_combine(st_geometry(sk_crop))
-au_geom <- st_combine(st_geometry(au_crop))
-cz_geom <- st_combine(st_geometry(cz_crop))
-
-rm(list = c("cz_crop", "au_crop", "sk_crop"))
-
-# save temp files...
-dt_hydro_derived <- "analysis/data/derived_data/hydro/"
-# dir.create(here(dt_hydro_derived))
-# st_write(sk_geom, here(dt_hydro_derived, "sk_prep.geojson"))
-# st_write(au_geom, here(dt_hydro_derived, "au_prep.geojson"))
-# st_write(cz_geom, here(dt_hydro_derived, "cz_prep.geojson"))
-#
-# rstudioapi::restartSession()
-#
-# sk_geom <- st_read(here(dt_hydro_derived, "sk_prep.geojson"))
[…]

### 2022-03-23 analysis/code/dataprep_init.R

@@ -1,349 +1,0 @@
-# # Project "Settlements"
-# # Script nr. 0
-# # DATA PREPARATION
-# # author: Petr Pajdla
-# # Prepares data for the analysis from raw spreadsheet formats
-#
-# library(here)
-# library(tidyverse)
-# library(sf)
-# library(googledrive)
-#
-# dt_raw <- "analysis/data/raw_data"
-# dt_der <- "analysis/data/derived_data"
-#
-# # funs --------------------------------------------------------------------
-#
-# # 1 / 0 indicators to logical TRUE / FALSE
-# indicator_to_logical <- function(x) {
-#   x <- as.character(x)
-#   x <- if_else(str_detect(x, "1"), "TRUE", x)
-#   x <- if_else(str_detect(x, "0"), "FALSE", x)
-#   x <- if_else(str_detect(x, "\\?"), "TRUE", x)
-#   return(as.logical(x))
-# }
-#
-# # creates chronologies from indicators
-# create_chrono <- function(x, chrono) {
-#   stopifnot(chrono %in% c(1, 2, 3))
-#
-#   if (chrono == 1) {
-#     chrono_cols <- c("SBK", "LgK", "ENE", "TRB")
-#   } else if (chrono == 2) {
-#     chrono_cols <- c("SBK1", "SBK2", "LgK1", "LgK2", "ENE0",
-#                      "TRB1", "TRB2")
-#   } else if (chrono == 3) {
-#     chrono_cols = c("ENE0_Epilengyel", "ENE0_Ludanice",
-#                     "ENE0_Schussenried", "ENE0_Michelsberg", "ENE0_Jordanow")
-#   }
-#
-#   x %>%
-#     select(id, any_of(chrono_cols)) %>%
-#     pivot_longer(-id,
-#                  names_to = "chrono",
-#                  values_to = "chrono_dubious", values_drop_na = TRUE) %>%
-#     # filter only certain sites
-#     filter(!str_detect(chrono_dubious, "\\?")) %>%
-#     select(-chrono_dubious)
-# }
-#
-# # create dataset
-# create_data_set <- function(x, pad) {
-#   stopifnot(pad %in% c("M", "B"))
-#
-#   res <- list(
-#     data = NA,
-#     chrono1 = NA,
-#     chrono2 = NA
-#   )
-#
-#   # unique ids
-#   dt <- x %>% mutate(id = str_pad(rowid, width = 4, side = "left", pad = "0"),
[…]

### 2022-03-23 analysis/code/dataprep_map.R

@@ -1,122 +1,0 @@
-# Project "Settlements"
-# Script nr. 0
-# PREPARE BACKGROUND MAP DATA
-# author: Petr Pajdla
-# Prepare background data for maps (crop, transform etc.)
-# Data from Natural Earth Project using rnaturalearth package
-# Own data for raw materials
-
-library(here)
-library(tidyverse)
-library(sf)
-library(rnaturalearth)
-
-# paths -------------------------------------------------------------------
-dt_raw <- "analysis/data/raw_data"
-dt_der <- "analysis/data/derived_data"
-
-dt_ne <- here(paste0(dt_raw, "/naturalearth"))
-dt_area <- here(paste0(dt_raw, "/project_area"))
-
-dt_map <- here(paste0(dt_der, "/maps"))
-
-# bbox & mask -------------------------------------------------------------
-bbox_large <- st_read(dt_area, "bbox_wgs84_large")
-mask <- st_read(dt_area, "maska_poly_wgs84") %>%
-  st_transform(5514)
-bbox <- st_bbox(st_buffer(mask, dist = 20000)) %>% st_as_sfc()
-
-# natural earth data ------------------------------------------------------
-crop_and_transform <- function(x) {
-  x %>%
-    st_crop(bbox_large) %>%
-    st_transform(5514) %>%
-    st_crop(bbox) %>%
-    st_geometry()
-}
-
-# rivers
-rivers0 <- ne_load(destdir = dt_ne,
-                   scale = 10,
-                   category = "physical",
-                   type = "rivers_lake_centerlines",
-                   returnclass = "sf") %>%
-  crop_and_transform()
-
-rivers1 <- ne_load(destdir = dt_ne,
-                   scale = 10,
-                   category = "physical",
-                   type = "rivers_europe",
-                   returnclass = "sf") %>%
-  crop_and_transform()
-
-# state + admin boundaries
-admin0 <- ne_load(scale = 10,
-                  category = "cultural",
-                  type = "boundary_lines_land",
-                  destdir = dt_ne,
-                  returnclass = "sf") %>%
-  crop_and_transform()
-
-admin1 <- ne_load(scale = 10,
-                  category = "cultural",
-                  type = "admin_1_states_provinces",
[…]

### 2022-03-23 analysis/code/dataprep_references.R

@@ -1,117 +1,0 @@
-# Project "Settlements"
-# Script nr. ?
-# PREPARE REFERENCES
-# author: Petr Pajdla
-# Corrected references are added to base database
-
-library(tidyverse)
-library(here)
-library(httr)
-
-# References
-refs <- readxl::read_excel(here("analysis/data/raw_data", "references.xlsx")) %>%
-  select(id, references) %>%
-  separate(references, into = str_c("r", 1:15), sep = ";") %>%
-  pivot_longer(-id) %>%
-  filter(!is.na(value)) %>%
-  mutate(value = str_trim(value)) %>%
-  select(id, reference = value)
-
-# PIAN to Akce
-pian <- refs %>%
-  mutate(pian = if_else(str_detect(reference, "^P-\\d+"), TRUE, FALSE)) %>%
-  filter(pian) %>%
-  select(-pian) %>%
-  mutate(pian1 = str_extract(reference, "^P-\\d+"),
-         pian2 = str_extract(reference, "\\d{5}$"),
-         pian = str_c(pian1, "-0", pian2)) %>%
-  select(-pian1, -pian2)
-
-# ask for password somwhere
-password <- rstudioapi::askForPassword()
-
-
-# query API ---------------------------------------------------------------
-
-#'@param child Child to select from the XML tree
-query_amcr_api <- function(amcr_ident, username, child) {
-
-  api_url <- "https://api.aiscr.cz/dapro/"
-  verb <- "oai?verb=GetRecord"
-  ident <- paste0("&identifier=https://api.aiscr.cz/id/", amcr_ident)
-  prefix <- "&metadataPrefix=oai_amcr"
-
-  resp <- GET(paste0(api_url, verb, ident, prefix),
-              authenticate(username, password))
-
-  res <- content(resp, "parsed", encoding = "UTF-8") %>%
-    rvest::html_elements(xpath = child) %>%
-    rvest::html_text()
-
-  return(res)
-}
-
-#'@param x vector of AMCR identifiers
-query_amcr_api_loop <- function(x, username, child) {
-
-  res <- vector("list", length(x))
-  names(res) <- x
-
-  for (i in seq_along(x)) {
-    Sys.sleep(0.1)
-    res[[i]] <- query_amcr_api(x[i], username, child)
-    print(i)
-  }
-
-  return(res)
-
-}
-
-
-# Queries -----------------------------------------------------------------
-
-# get dokumentační jednotky
[…]

### 2022-03-23 analysis/code/dem_aspect.R

@@ -1,0 +1,71 @@
+# Project "Settlements"
+# Script nr. -
+# DEM: ASPECT
+# author: Petr Pajdla
+# Extract values from DEM: aspect, not included further
+
+library(dplyr)
+library(sf)
+devtools::load_all()
+
+# size of a buffer zone around the settlement (meters)
+buffer_dist <- units::set_units(0.2, "km")
+
+# data --------------------------------------------------------------------
+
+source(here::here("analysis/code/data_load.R"))
+
+# dem
+r <- terra::rast(here::here("analysis/data/raw_data/dem/DEM_digital_elevation.tif"))
+
+# vicinity of settlemetns
+set_buffer <- set_spat %>%
+  st_transform(terra::crs(r)) %>%
+  st_buffer(dist = buffer_dist)
+
+# slope -------------------------------------------
+
+median_asp <- extract_from_dem(set_buffer, r, model = "aspect", fun = "median")
+
+# median_asp %>%
+#   select(id, value) %>%
+#   readr::write_csv(here::here(derived_data, "from_dem/tpi.csv"))
+
+# explore -----------------------------------------
+
+shapiro.test(filter(median_asp, reg == "B")$value)
+shapiro.test(filter(median_asp, reg == "M")$med_asp)
+
+library(ggplot2)
+
+median_asp %>%
+  mutate(reg = stringr::str_extract(id, ".")) %>%
+  ggplot(aes(value)) +
+  geom_histogram(color = "black", fill = "white") +
+  scale_x_continuous(limits = c(0, 360),
+                     breaks = c(0, 90, 180, 270)) +
+  coord_polar(start = 0) +
+  facet_wrap(vars(reg)) +
+  theme_bw() +
+  labs(x = "median aspect (°)")
+
+median_asp %>%
+  mutate(reg = stringr::str_extract(id, ".")) %>%
+  ggplot(aes(value)) +
+  geom_histogram(color = "black", fill = "white") +
+  scale_x_continuous(limits = c(0, 360),
+                     breaks = c(0, 90, 180, 270)) +
+  facet_wrap(vars(reg)) +
+  theme_bw() +
+  labs(x = "median aspect (°)")
+
+median_asp %>%
+  mutate(reg = stringr::str_extract(id, ".")) %>%
+  inner_join(settlements2, by = "id") %>%
+  ggplot(aes(value)) +
+  geom_histogram(color = "black", fill = "white") +
[…]

### 2022-03-23 analysis/code/dem_slope.R

@@ -1,0 +1,76 @@
+# Project "Settlements"
+# Script nr. 1
+# DEM: SLOPE
+# author: Petr Pajdla
+# Extract values from DEM: slope
+
+library(dplyr)
+library(sf)
+devtools::load_all()
+
+# size of a buffer zone around the settlement (meters)
+buffer_dist <- units::set_units(0.5, "km")
+
+# data --------------------------------------------------------------------
+
+source(here::here("analysis/code/data_load.R"))
+
+# dem
+r <- terra::rast(here::here("analysis/data/raw_data/dem/DEM_digital_elevation.tif"))
+
+# vicinity of settlemetns
+set_buffer <- set_spat %>%
+  st_transform(terra::crs(r)) %>%
+  st_buffer(dist = buffer_dist)
+
+
+# # altitude from dem -------------------------------
+#
+# nmv <- extract(r, vect(set_buffer))
+#
+# nmv %>% tibble::tibble() %>%
+#   group_by(ID) %>%
+#   summarise(mean = mean(DEM_digital_elevation)) %>%
+#   bind_cols(id = set_spat$id, alt = set_spat$altitude) %>%
+#   summarise(cor = cor(mean, alt))
+
+
+# slope -------------------------------------------
+
+mean_slope <- extract_from_dem(set_buffer, r, model = "slope", fun = "mean")
+
+mean_slope %>%
+  select(id, value) %>%
+  readr::write_csv(here::here(derived_data, "slope_at_points.csv"))
+
+# explore -----------------------------------------
+
+library(ggplot2)
+
+mean_slope %>%
+  mutate(reg = stringr::str_extract(id, ".")) %>%
+  ggplot(aes(value, reg)) +
+  geom_violin() +
+  geom_boxplot(width = 0.2) +
+  geom_jitter(alpha = 0.1, height = 0.2) +
+  theme_bw() +
+  coord_flip() +
+  labs(x = "slope (°)")
+
+# mean_slope %>%
+#   mutate(reg = stringr::str_extract(id, ".")) %>%
+#   inner_join(settlements2, by = "id") %>%
+#   ggplot(aes(slope)) +
+#   geom_histogram(color = "black", fill = "white") +
+#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+#
+# mean_slope %>%
[…]

### 2022-03-23 analysis/code/dem_tpi.R

@@ -1,0 +1,68 @@
+# Project "Settlements"
+# Script nr. 2
+# DEM: TPI
+# author: Petr Pajdla
+# Extract values from DEM: Terrain/Topography position index
+
+library(dplyr)
+library(sf)
+devtools::load_all()
+
+# size of a buffer zone around the settlement (meters)
+buffer_dist <- units::set_units(0.2, "km")
+
+# data --------------------------------------------------------------------
+
+source(here::here("analysis/code/data_load.R"))
+
+# dem
+r <- terra::rast(here::here("analysis/data/raw_data/dem/DEM_digital_elevation.tif"))
+
+# vicinity of settlemetns
+set_buffer <- set_spat %>%
+  st_transform(terra::crs(r)) %>%
+  st_buffer(dist = buffer_dist)
+
+
+# tpi ---------------------------------------------
+
+mean_tpi <- extract_from_dem(set_buffer, r, model = "TPI", fun = "mean")
+
+mean_tpi %>%
+  select(id, value) %>%
+  readr::write_csv(here::here(derived_data, "tpi_at_points.csv"))
+
+# explore -----------------------------------------
+
+library(ggplot2)
+
+mean_tpi %>%
+  mutate(reg = stringr::str_extract(id, ".")) %>%
+  ggplot(aes(value, reg)) +
+  geom_violin() +
+  geom_boxplot(width = 0.2) +
+  geom_jitter(alpha = 0.1, height = 0.2) +
+  theme_bw() +
+  coord_flip() +
+  labs(x = "TPI (terrain position index)")
+
+# mean_slope %>%
+#   mutate(reg = stringr::str_extract(id, ".")) %>%
+#   inner_join(settlements2, by = "id") %>%
+#   ggplot(aes(slope)) +
+#   geom_histogram(color = "black", fill = "white") +
+#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+#
+# mean_slope %>%
+#   mutate(reg = stringr::str_extract(id, ".")) %>%
+#   inner_join(settlements2, by = "id") %>%
+#   ggplot(aes(slope, period_label)) +
+#   geom_boxplot(color = "black", fill = "white") +
+#   facet_wrap(vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+#
[…]

### 2022-03-23 analysis/code/fenced_dist.R

@@ -1,193 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# DISTANCE TO FENCED SETTLEMENT
-# author: Petr Pajdla
-# Calculated separately for the speed
-
-# devtools::load_all(".")
-library(tidyverse)
-library(sf)
-
-derived_data <- "analysis/data/derived_data"
-path_maps <- here::here(derived_data, "maps/")
-
-mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
-
-# database
-set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
-
-set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
-                    quiet = TRUE)
-
-chrono_labels <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
-
-# settlements spatial by periods
-set_spat_1 <- set_spat %>%
-  right_join(set_base$period1) %>%
-  mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
-
-set_spat_2 <- set_spat %>%
-  right_join(set_base$period2) %>%
-  mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
-
-# fenced settlements
-b <- readxl::read_excel(
-  here::here("analysis/data/raw_data/ohrazena_sidliste.xlsx"),
-  sheet = "b")
-
-m <- readxl::read_excel(
-  here::here("analysis/data/raw_data/ohrazena_sidliste.xlsx"),
-  sheet = "m")
-
-# point layer
-# period I - pottery tradtions
-fortif_p1 <- bind_rows(b = b, m = m, .id = "reg") %>%
-  separate(abbrv_p2, into = c("p2_1", "p2_2"), sep = ";") %>%
-  pivot_longer(c(p2_1, p2_2), values_to = "abbrv_p2") %>%
-  select(-name) %>%
-  filter(!is.na(abbrv_p2)) %>%
-  mutate(abbrv_p2 = str_trim(abbrv_p2, "both"),
-         "p5.0" = if_else(abbrv_p1  == "SBK", TRUE, FALSE),
-         "p4.8" = if_else(abbrv_p1  %in% c("SBK", "LgK"), TRUE, FALSE),
-         "p4.6" = if_else(abbrv_p1  == "LgK", TRUE, FALSE),
[…]

### 2022-03-23 analysis/code/hydro.R

@@ -1,0 +1,47 @@
+# Project "Settlements"
+# Script nr. 3
+# DENSITY OF WATERCOURSE
+# author: Petr Pajdla
+#
+
+library(dplyr)
+library(sf)
+library(terra)
+
+# size of a buffer zone around the settlement (meters)
+buffer_dist <- units::set_units(0.2, "km")
+
+# data --------------------------------------------------------------------
+
+source(here::here("analysis/code/data_load.R"))
+
+# watercourse raster
+hydro_dens <- terra::rast(
+  here::here(derived_data, "hydro", "density_raster.tif")
+)
+
+# vicinity of settlemetns
+set_buffer <- set_spat %>%
+  st_transform(terra::crs(hydro_dens)) %>%
+  st_buffer(dist = buffer_dist)
+
+
+# extract vals ------------------------------------
+
+hydro_dens_at_points <- terra::extract(
+  hydro_dens,
+  terra::vect(set_buffer)
+) %>%
+  group_by(ID) %>%
+  summarise(hydro_kde = mean(density_raster)) %>%
+  bind_cols(id = set_spat$id) %>%
+  select(-ID)
+
+hydro_dens_at_points %>%
+  readr::write_csv(here::here(derived_data, "hydro_at_points.csv"))
+
+# explore -----------------------------------------
+
+hydro_dens_at_points %>%
+  pull(hydro_kde) %>%
+  hist()

### 2022-03-23 analysis/code/joad/joad_data.R

@@ -1,0 +1,228 @@
+# prepare data for JOAD submission / Zotero
+
+library(sf)
+library(tidyverse)
+
+
+# data --------------------------------------------------------------------
+
+derived_data <- "analysis/data/derived_data"
+path_maps <- here::here(derived_data, "maps/")
+path_data <- here::here(derived_data)
+path_out <- here::here("analysis/data/joad_data/")
+
+mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
+
+labs_chrono <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
+
+set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
+
+set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+                    quiet = TRUE)
+
+source("./R/rm_distance.R")
+rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>%
+  add_rm_type()
+
+rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>%
+  add_rm_type()
+
+
+# filtering ---------------------------------------------------------------
+
+in_analysis <- set_base$base %>%
+  filter(in_analysis) %>%
+  pull(id)
+
+# maska -------------------------------------------------------------------
+# mask %>%
+#   st_geometry() %>%
+#   st_write(paste0(path_out, "geodata/", "study_area.gml"), append = FALSE)
+
+# plot(st_read(paste0(path_out, "study_area.gml"), options="FORCE_SRS_DETECTION=YES"))
+
+
+# site locations ----------------------------------------------------------
+
+# set_spat %>%
+#   select(-site) %>%
+#   filter(id %in% in_analysis) %>%
+#   mutate(accuracy = as.integer(accuracy)) %>%
+#   arrange(id) %>%
+#   st_write(paste0(path_out, "geodata/", "site_locations.gml"), append = FALSE)
+
+# fuu <- st_read(paste0(path_out, "site_locations.gml"), options="FORCE_SRS_DETECTION=YES")
+
+
+# regions ----------------------------------------
+
+buffer <- set_spat %>%
+  filter(id %in% in_analysis) %>%
[…]

### 2022-03-23 analysis/code/joad/joad_map.R

@@ -1,0 +1,153 @@
+# fancy map for doaj article
+
+library(tidyverse)
+library(sf)
+library(stars)
+
+
+# data --------------------------------------------------------------------
+
+set_spat <- st_read(here::here("analysis/data/derived_data/settlements_sf.geojson"))
+
+mask <- st_read(here::here("analysis/data/raw_data/project_area/maska_poly_wgs84.dbf")) %>%
+  st_transform(32633)
+
+reg <- st_read(here::here("analysis/data/joad_data/geodata/regions.gml"),
+               options="FORCE_SRS_DETECTION=YES") %>%
+  st_transform(32633)
+
+# ne data
+# https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/raster/NE2_LR_LC_SR_W_DR.zip
+# https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/raster/GRAY_HR_SR_W.zip
+# ne_raster <- rnaturalearth::ne_download(scale = 10, category = "raster",
+#                                         type = "GRAY_HR_SR_W",
+#                                         destdir = here::here("analysis/data/temp/"),
+#                                         returnclass = "sf")
+
+# ne_raster <- read_stars(here::here("analysis/data/temp/GRAY_HR_SR_W.tif"))
+
+raster <- read_stars(here::here("analysis/data/raw_data/dem/DEM_digital_elevation.tif"))
+
+riv <- rnaturalearth::ne_download(scale = 10, category = "physical",
+                                  type = "rivers_lake_centerlines_scale_rank",
+                                  returnclass = "sf")
+
+# https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip
+bord <- rnaturalearth::ne_download(scale = 10, category = "cultural",
+                                   type = "admin_0_countries",
+                                   returnclass = "sf")
+
+# prep --------------------------------------------------------------------
+
+bbox <- set_spat %>%
+  st_transform(32633) %>%
+  st_buffer(1.5e4) %>%
+  st_bbox() %>%
+  st_as_sfc()
+
+# crop to ce
+ce <- st_crop(raster, bbox)
+# plot(ce)
+
+# rivers
+riv2 <- riv %>%
+  filter(st_is_valid(riv)) %>%
+  st_crop(st_transform(bbox, 4326)) %>%
[…]

### 2022-03-23 analysis/code/percolation.R

@@ -1,344 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# PERCOLATION ANALYSIS
-# author: Petr Pajdla
-# Find percolation distances
-
-library(tidyverse)
-library(igraph)
-library(sf)
-library(here)
-
-
-# functions ---------------------------------------------------------------
-
-# percolation function
-percolation <- function(x, lower, upper, step) {
-  stopifnot(any(class(x) == "sf"))
-  m <- st_coordinates(x)
-  rownames(m) <- x$id
-
-  dist <- as.matrix(stats::dist(m, method = "euclidean"))
-
-  # radius
-  # triangle <- dist[lower.tri(dist)]
-  # if (is.null(lower)) {
-  #   lower <- min(triangle)
-  # }
-  # if (is.null(upper)) {
-  #   upper <- max(triangle) / 4
-  # }
-  # if (is.null(step)) {
-  #   step <- (upper - lower) / 10
-  # }
-  radii <- seq(lower, upper, by = step)
-
-  res_lst <- vector("list", length(radii))
-  names(res_lst) <- radii
-
-  # distances
-  dist_df <- reshape2::melt(dist)
-  colnames(dist_df) <- c("from", "to", "value")
-  dist_df <- dist_df[!(dist_df$from == dist_df$to), ]
-
-  for (i in seq_along(radii)) {
-    index <- dist_df$value <= radii[i]
-    if (!any(index)) {
-      res_lst[[as.character(radii[i])]] <- matrix(
-        ncol = 2,
-        dimnames = list(NULL, c("id", "clust")))
-    } else {
-      nodes <- dist_df[index, c("from", "to")]
-      nodes <- sapply(nodes, as.character)
-      g <- graph_from_edgelist(nodes)
-
-      # get clusters
-      clust_memb <- clusters(g)$membership
-      # clust_memb <- clusters(g, mode = "weak")$membership
-      memb <- matrix(ncol = 2, nrow = length(clust_memb))
-      colnames(memb) <- c("id", "clust")
-      memb[, "clust"]  <- unname(clust_memb)
-      memb[, "id"] <- names(clust_memb)
-
-      # result - list
-      res_lst[[as.character(radii[i])]] <- memb
-    }
-  }
-
-  res_tbl <- lapply(res_lst, as_tibble) %>%
-    bind_rows(.id = "radius") %>%
-    filter(!is.na(id)) %>%
-    mutate(radius = as.numeric(radius))
-
-  # membership for different radii
-  res <- vector("list", 3)
[…]

### 2022-03-23 analysis/code/rm_distance.R

@@ -9,20 +9,20 @@ library(tidyverse)
 library(sf)
 
 
-derived_data <- "analysis/data/derived_data"
-temp_data <- here::here(derived_data, "temp")
+input_path <- "analysis/data/input_data"
+input_data <- here::here(derived_data)
 
 # database
-set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
+set_base <- read_rds(here::here(input_data, "settlements.RDS"))
 
-set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
+set_spat <- st_read(here::here(input_data, "settlements_sf.geojson"),
                     quiet = TRUE)
 
 # raw materials
-rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>%
+rm_pts <- st_read(here::here("analysis/data/derived_data/rm_points.geojson"), quiet = TRUE) %>%
   mutate(label = if_else(orig == "l", "Chipped", "Polished"))
 
-rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>%
+rm_lns <- st_read(here::here("analysis/data/derived_data/rm_lines.geojson"), quiet = TRUE) %>%
   mutate(label = if_else(orig == "l", "Chipped", "Polished"))
 
 # distance
@@ -39,12 +39,11 @@ dist <- bind_rows(p = dist_pts, l = dist_lns, .id = "shp") %>%
          type = if_else(rm %in% pstrm, "Polished", "Chipped"),
          region = if_else(str_detect(id, "^B"), "Boh.", "Mor."))
 
-
 # save result
 
-if (!dir.exists(temp_data)) {
-  dir.create(temp_data)
-}
+# if (!dir.exists(temp_data)) {
+#   dir.create(temp_data)
+# }
 
-write_csv(dist, paste0(temp_data, "/rm_dist.csv"))
+write_csv(dist, here::here("analysis/data/input_data/rm_dist.csv"))
 

### 2022-03-23 analysis/groups/groups.Rmd
add new docs edit orig paper

---
title: "SI: Settlements / Pottery groups"
author: Petr Pajdla
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      toc: true
      toc_depth: 3
bibliography: ../paper/references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
# keywords: |
#   keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

<!-- Keywords: `r rmarkdown::metadata$keywords` -->

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = FALSE,
  comment = "#>",
  # fig.path = "../figures/",
  dpi = 300
)
```

```{r packages, cache=FALSE}
library(sf)
library(tidyverse)
library(spatstat)
library(patchwork)
devtools::load_all(".")
```

# Introduction

The analysis is divided into three scales:

-   Neolithic B/C
-   Pottery traditions
-   Pottery groups

**This document describes results for Pottery groups.**

The study area is broadly defined as *Easter part of Bohemia* (abbreviated *B*)
and *Morava river catchment* (*M*). It spans across the area of the Czech
Republic and bordering regions of Austria and Slovakia.

<!-- ======================================================================= -->

## Data

Data is published in the Journal of Open Archaeology Data:

> Pajdla, P. and Trampota, F., 2021. Neolithic Settlements in Central Europe: Data from the Project ‘Lifestyle as an Unintentional Identity in the Neolithic’. Journal of Open Archaeology Data, 9, p.13. DOI: <http://doi.org/10.5334/joad.88>

```{r load-data, cache = FALSE}
source(here::here("analysis/code/data_load.R"))
```

```{r tab-set-nr}
set_base$period2 %>%
  table_nr_set()
```

```{r map-set-nr, fig.cap = 'Settlements across time periods.'}
settlements2 %>%
  plot_nr_set()
```

<!-- ======================================================================= -->

# Analysis

## Settlement density

Settlement density is estimated using KDE. Kernel size of 4 km is used. This
value is completely arbitrary but is selected because settlement clusters are
nicely highlighted at this scale.

```{r map-kde, fig.cap = 'Kernel density estimation for different periods.'}
settlements2 %>% 
  plot_kde()
```

```{r set-kde}
kde <- settlements2 %>% 
  est_kde() %>% 
  mutate(reg = str_extract(id, "^."))
```

```{r plot-kde}
kde %>% ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "KDE")
```

<!--- ====================================================================== -->

## Terrain surrounding the settlements

### Altitude

```{r alt-data}
alt <- settlements2 %>% 
  select(id, altitude) %>% 
  st_drop_geometry() %>% 
  distinct(id, altitude) %>% 
  transmute(id, value = altitude,
            variable = "altitude")
```

```{r alt}
settlements2 %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, altitude)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Altitude (m)")
```

### Slope

```{r slope}
slope <- read_csv(here::here("analysis/data/derived_data", 
                             "slope_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "slope")
```

```{r plot-slope}
slope %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Slope (°)")
```

### TPI

> Terrain / Topographic position index

```{r tpi}
tpi <- read_csv(here::here("analysis/data/derived_data", 
                           "tpi_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "tpi")
```

```{r plot-tpi}
tpi %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "TPI")
```

### Density of water courses

```{r hydro}
hydro <- read_csv(here::here("analysis/data/derived_data", 
                             "hydro_at_points.csv")) %>% 
  filter(id %in% settlements2$id) %>% 
  mutate(variable = "hydro") %>% 
  rename(value = hydro_kde)
```

```{r plot-hydro}
hydro %>% 
  left_join(settlements2, by = "id") %>% 
  mutate(reg = str_extract(id, "^.")) %>% 
  ggplot(aes(period, value)) +
  geom_violin() +
  geom_boxplot(alpha = 0.4, width = 0.2) +
  geom_jitter(alpha = 0.05, width = 0.4) +
  facet_wrap(vars(reg)) +
  theme_bw() +
  labs(y = "Hydrology KDE")
```

<!--- ====================================================================== -->

<!-- # Settlement 'hierarchy' -->

<!-- Minimal distance to fortified settlements in given regions is explored. -->

<!--- ====================================================================== -->

# Arrangement along natural lines

```{r along-lines}
along_lines <- read_rds(here::here(derived_data, "lines/linear_arrangement.RDS"))

along_lines <- along_lines$groups %>% 
  mutate(chrono = str_remove(chrono, ".$")) %>% 
  group_by(id, chrono, period, variable) %>% 
  summarise(value = max(value)) %>% 
  ungroup(id, chrono, variable) %>% 
  mutate(value = value / n()) %>% 
  select(-chrono) %>% 
  group_by(variable, period, id) %>% 
  summarise(value = mean(value), .groups = "drop")
```

<!--- ====================================================================== -->

# Distance to raw material sources

```{r rm}
# calculated separately in a script `rm_distance.R`
# dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
# dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")
rm_dist <- read_csv(here::here("analysis/data/input_data", "rm_dist.csv")) %>% 
  mutate(rm = str_to_lower(rm)) %>% 
  filter(id %in% settlements2$id)

# 0 = not used
# 0.5 = somewhat used
# 1 = used
rm_b <- readxl::read_xlsx(here::here("analysis/data/raw_data",
                                     "sidliste_vs_zdroje.xlsx"), sheet = "B")
rm_m <- readxl::read_xlsx(here::here("analysis/data/raw_data",
                                     "sidliste_vs_zdroje.xlsx"), sheet = "M")

rm_relevance <- bind_rows(b = rm_b, m = rm_m, .id = "reg") %>% 
  mutate(across(starts_with("p"), function(x) x * 0.5)) %>% 
  pivot_longer(starts_with("p"), 
               names_to = "period", values_to = "relevance") %>% 
  rename(rm = surovina)

rm_dist_rel <- set_base$period2 %>% 
  filter_rel_dist_rm()

# ---
rm_types <- rm_dist %>% 
  select(rm, type) %>% 
  distinct()
rm_types_tab <- rm_types %>% 
  pull(type)
names(rm_types_tab) <- rm_types %>% 
  pull(rm)
# ---

rm_dist_fin <- rm_dist_rel %>% 
  prep_dist_rel_rm()
```

<!--- ====================================================================== -->

# Synthesis

```{r variable-labels}
labs_variables <- c(
  "settlements_kde" = "Settlement density",
  # "dist_fenced" = "Min. dist. to fenced set. (km)",
  "line_terrain" = "Along terrain line",
  "line_water" = "Along water line",
  # "cont" = "Settlement continuity",
  "hydro" = "Watercourse density",
  # "dem" = "Altitude (DEM)", 
  "altitude" = "Altitude (map)",
  "slope" = "Slope (°)",
  "tpi" = "TPI",
  "rm_dist_chipped" = "Distance to chipped RM",
  "rm_dist_polished" = "Distance to polished RM"
  # "rm_pc1" = "Distance to RM sources 1",
  # "rm_pc2" = "Distance to RM sources 2",
  # "rm_pc3" = "Distance to RM sources 3"
) %>% 
  map_chr(str_wrap, 12)
```

```{r variables}
res_long <- alt %>% 
  bind_rows(slope) %>% 
  bind_rows(tpi) %>% 
  bind_rows(hydro) %>% 
  left_join(set_base$period2) %>% 
  select(id, variable, value, period) %>% 
  bind_rows(along_lines) %>% 
  bind_rows(rm_dist_fin) %>% 
  bind_rows(select(kde, -reg)) %>% 
  filter(id %in% set_base$period2$id) %>% 
  group_by(variable, id, period) %>% 
  summarise(value = mean(value)) %>% 
  ungroup()

res_wide <- res_long %>% 
  pivot_wider(id_cols = c(id, period), 
              names_from = variable, 
              values_from = value, values_fill = 0)

```

## Principal components analysis

```{r nested-pca}
# traditions
pca_per <- res_wide %>%
  add_region() %>% 
  group_by(reg, period) %>% 
  nest() %>% 
  nested_pca()
```

```{r pca-plot}
pca_per %>% plot_nested_pca(var = "period")
```

## Groups in the data

```{r package-mclust}
library(mclust)
```

```{r mclust2-neo}
mb_clust <- pca_per %>% mbc_cluster()

classif <- mb_clust %>%
  dplyr::select(reg, period, classif) %>% 
  tidyr::unnest(c(classif)) %>% 
  dplyr::ungroup(c(reg, period)) %>% 
  dplyr::mutate(g = str_c(reg, g))

set_spat %>% 
  select(id) %>% 
  inner_join(classif, by = "id") %>% 
  ggplot() +
  geom_sf(aes(color = g)) +
  facet_grid(vars(period), vars(g)) +
  theme_void()
```

```{r spat-out}
# set_spat %>% 
#   select(id) %>% 
#   inner_join(classif, by = "id") %>% 
#   write_sf(here::here("analysis/groups/groups.geojson"), delete_layer = TRUE)
```

```{r plot-vars-g}
plot_g_distribution <- function(var) {
  res_long %>% 
    filter(variable == var) %>% 
    inner_join(classif, by = c("id", "period")) %>% 
    add_period_label() %>% 
    ggplot(aes(g, value)) +
    geom_boxplot() +
    facet_wrap(vars(period_label), scales = "free_x") +
    theme_bw() +
    labs(title = var)
}

plot_g_distribution("altitude")
plot_g_distribution("slope")
plot_g_distribution("tpi")

plot_g_distribution("hydro")

plot_g_distribution("line_terrain")
plot_g_distribution("line_water")

plot_g_distribution("rm_dist_chipped")
plot_g_distribution("rm_dist_polished")

plot_g_distribution("settlements_kde")
```

<!--- ====================================================================== -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational
environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

### 2022-03-23 analysis/neo/neo.Rmd
add new docs edit orig paper

--- 2022-03-23 analysis/groups/groups.Rmd
@@
 ---
-title: "SI: Settlements / Pottery groups"
+title: "SI: Settlements / Neo"
 author: Petr Pajdla
@@
 
-**This document describes results for Pottery groups.**
+**This document describes results for Neolithic (A)/B/C.**
 
@@
 
+```{r}
+set_base$period2$id %in% set_base$period1$id %>% table()
+
+neoB <- c("SBK", "LgK")
+neoC <- c("ENE", "TRB")
+
+neo <- set_base$period2 %>% 
+  mutate(chrono = str_remove(chrono, ".$")) %>% 
+  select(-facet) %>% 
+  bind_rows(set_base$period1) %>% 
+  select(-starts_with("period")) %>% 
+  arrange(id) %>% 
+  distinct(id, chrono) %>% 
+  mutate(period = if_else(chrono %in% neoB, "NeoB", NA_character_),
+         period = if_else(chrono %in% neoC, "NeoC", period)) %>% 
+  select(-chrono) %>% 
+  distinct(id, period) %>% 
+  mutate(period_label = if_else(period == "NeoB", "Neolithic B", "Neolithic C"),
+         label = period_label)
+
+neo_spat <- neo %>%
+  left_join(set_spat, by = "id") %>% 
+  st_as_sf()
+
+# edit labs for chrono
+labs_chrono$periods <- c("Neolithic B", "Neolithic C") %>% 
+  setNames(c("NeoB", "NeoC"))
+```
+
 ```{r tab-set-nr}
-set_base$period2 %>%
+neo %>%
   table_nr_set()
@@
 ```{r map-set-nr, fig.cap = 'Settlements across time periods.'}
-settlements2 %>%
+neo_spat %>% 
   plot_nr_set()
@@
 ```{r map-kde, fig.cap = 'Kernel density estimation for different periods.'}
-settlements2 %>% 
+neo_spat %>% 
   plot_kde()
@@
 ```{r set-kde}
-kde <- settlements2 %>% 
+kde <- neo_spat %>% 
+  mutate(chrono = period) %>% 
   est_kde() %>% 
@@
 ```{r alt-data}
-alt <- settlements2 %>% 
+alt <- neo_spat %>% 
   select(id, altitude) %>% 
@@
 ```{r alt}
-settlements2 %>% 
+neo_spat %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                              "slope_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% neo$id) %>% 
   mutate(variable = "slope")
@@
 slope %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(neo, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                            "tpi_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% neo$id) %>% 
   mutate(variable = "tpi")
@@
 tpi %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(neo, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                              "hydro_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% neo$id) %>% 
   mutate(variable = "hydro") %>% 
@@
 hydro %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(neo, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
 along_lines <- along_lines$groups %>% 
+  bind_rows(along_lines$traditions) %>% 
   mutate(chrono = str_remove(chrono, ".$")) %>% 
-  group_by(id, chrono, period, variable) %>% 
-  summarise(value = max(value)) %>% 
-  ungroup(id, chrono, variable) %>% 
-  mutate(value = value / n()) %>% 
+  select(-starts_with("period")) %>% 
+  arrange(id) %>% 
+  mutate(period = if_else(chrono %in% neoB, "NeoB", NA_character_),
+         period = if_else(chrono %in% neoC, "NeoC", period)) %>% 
   select(-chrono) %>% 
   group_by(variable, period, id) %>% 
-  summarise(value = mean(value), .groups = "drop")
+  summarise(value = mean(value)) %>% 
+  ungroup(c(variable, period, id)) %>% 
+  mutate(period_label = if_else(period == "NeoB", "Neolithic B", "Neolithic C"),
+         label = period_label)
+```
+
+```{r along-lines-plot}
+along_lines %>% 
+  mutate(reg = str_extract(id, "^.")) %>% 
+  ggplot(aes(period, value)) +
+  geom_violin() +
+  geom_boxplot(alpha = 0.4, width = 0.2) +
+  geom_jitter(alpha = 0.05, width = 0.4) +
+  facet_grid(vars(variable), vars(reg)) +
+  theme_bw() +
+  labs(y = NULL)
 ```
@@
                names_to = "period", values_to = "relevance") %>% 
-  rename(rm = surovina)
+  rename(rm = surovina) %>% 
+  mutate(period2 = if_else(period %in% c("p5.0", "p4.8", "p4.6", "p4.4"), "NeoB", NA_character_),
+         period2 = if_else(period %in% c("p4.2", "p4.0", "p3.8", "p3.6", "p3.4"), "NeoC", period2)) %>% 
+  select(rm, period = period2, relevance, reg)
 
-rm_dist_rel <- set_base$period2 %>% 
+rm_dist_rel <- neo %>% 
   filter_rel_dist_rm()
@@
   bind_rows(hydro) %>% 
-  left_join(set_base$period2) %>% 
+  left_join(neo) %>% 
   select(id, variable, value, period) %>% 
-  bind_rows(along_lines) %>% 
+  bind_rows(select(along_lines, -ends_with("label"))) %>% 
   bind_rows(rm_dist_fin) %>% 
   bind_rows(select(kde, -reg)) %>% 
-  filter(id %in% set_base$period2$id) %>% 
+  filter(id %in% neo$id) %>% 
   group_by(variable, id, period) %>% 
@@
               values_from = value, values_fill = 0)
-
 ```
@@
   geom_sf(aes(color = g)) +
-  facet_grid(vars(period), vars(g)) +
+  facet_grid(vars(g), vars(period)) +
   theme_void()
@@
 #   inner_join(classif, by = "id") %>% 
-#   write_sf(here::here("analysis/groups/groups.geojson"), delete_layer = TRUE)
+#   write_sf(here::here("analysis/neo/neo.geojson"), delete_layer = TRUE)
 ```

### 2022-03-23 analysis/paper/paper.Rmd
add new docs edit orig paper

@@
 # abstract: |
-keywords: |
-  keyword 1; keyword 2; keyword 3
+# keywords: |
+#   keyword 1; keyword 2; keyword 3
 editor_options: 
@@
 
-Keywords: `r rmarkdown::metadata$keywords`
+<!-- Keywords: `r rmarkdown::metadata$keywords` -->
 
@@
   mutate(rm = str_to_lower(rm))
-
-# rm_dist_mean <- rm_dist %>%
-#   select(-region, -type, -shp, -min, -median, -max) %>%
-#   group_by(id, rm) %>%
-#   summarise(dist_rm = mean(mean)) %>%
-#   pivot_longer(cols = dist_rm, names_to = "variable") %>%
-#   mutate(variable = str_c(variable, rm, sep = "_")) %>%
-#   select(-rm)
 ```
@@
 
+```{r rm-clean}
+rm(list = ls(pattern = "^rm_"))
+```
+
 <!-- ======================================================================= -->
@@
 
-```{r get-raster}
+```{r count-water-dens}
 # pkg raster
@@
 # pkg stars
-hydro_dens <- stars::read_stars(
+# hydro_dens <- stars::read_stars(
+#   here::here(derived_data, "hydro", "density_raster.tif")
+# ) %>% 
+#   st_set_crs(value = st_crs(set_spat))
+# 
+# hydro_dens_at_points <- hydro_dens %>% 
+#   stars::st_extract(set_spat) %>% 
+#   st_drop_geometry() %>% 
+#   rename(hydro_kde = density_raster.tif) %>% 
+#   bind_cols(select(st_drop_geometry(set_spat), id))
+
+# pkg terra, implementing buffer around settlements
+# settlements represented as points
+
+hydro_dens <- terra::rast(
   here::here(derived_data, "hydro", "density_raster.tif")
+)
+
+hydro_dens_at_points <- terra::extract(
+  hydro_dens, 
+  terra::vect(st_buffer(set_spat, dist = 2e3))
 ) %>% 
-  st_set_crs(value = st_crs(set_spat))
-
-hydro_dens_at_points <- hydro_dens %>% 
-  stars::st_extract(set_spat) %>% 
-  st_drop_geometry() %>% 
-  rename(hydro_kde = density_raster.tif) %>% 
-  bind_cols(select(st_drop_geometry(set_spat), id))
+  group_by(ID) %>% 
+  summarise(hydro_kde = mean(density_raster)) %>% 
+  bind_cols(id = set_spat$id) %>% 
+  select(-ID)
 
@@
 
-ggplot() +
-  stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
-  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
-  theme_void() +
-  labs(fill = "KDE")
-```
+# stars
+# ggplot() +
+#   stars::geom_stars(data = st_crop(hydro_dens, mask), sf = TRUE) +
+#   geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
+#   theme_void() +
+#   labs(fill = "KDE")
+```
+
+```{r water-clean}
+rm(list = c("hydro_dens", "hydro_dens_at_points"))
+```
+
 <!--- ====================================================================== -->
@@
 
-We focus on slope around the settlements.
-
+```{r dem}
+dem_terra <- terra::rast(here::here(derived_data, "/dem/dem_aster_sjtsk_stars.tif"))
+```
+
+## Slope
+
+```{r slope}
+dem_slope <- terra::terrain(dem_terra, v = "slope")
+
+# slope in the vicinity of settlements - 2km
+slope <- terra::extract(dem_slope, 
+                        terra::vect(st_buffer(set_spat, dist = 2e3)))
+
+# derive median aspect
+mean_slope <- slope %>% 
+  tibble() %>% 
+  group_by(ID) %>% 
+  summarise(mean_slope = mean(slope)) %>% 
+  bind_cols(id = set_spat$id) %>% 
+  mutate(reg = str_extract(id, "^."))
+
+# shapiro.test(filter(mean_slope, reg == "B")$mean_slope)
+# hist(filter(mean_slope, reg == "B")$mean_slope)
+# qqnorm(filter(mean_slope, reg == "B")$mean_slope)
+# 
+# shapiro.test(filter(mean_slope, reg == "M")$mean_slope)
+# hist(filter(mean_slope, reg == "M")$mean_slope)
+# qqnorm(filter(mean_slope, reg == "M")$mean_slope)
+
+# mean_slope %>% 
+#   inner_join(settlements2, by = "id") %>% 
+#   ggplot(aes(mean_slope)) + 
+#   geom_histogram(color = "black", fill = "white") +
+#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+
+# mean_slope %>% 
+#   inner_join(settlements2, by = "id") %>% 
+#   ggplot(aes(mean_slope, period_label)) + 
+#   geom_boxplot(color = "black", fill = "white") +
+#   facet_wrap(vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+```
+
+```{r slope-to-res}
+res1_long <- mean_slope %>% 
+  select(-ID, -reg) %>% 
+  inner_join(st_drop_geometry(select(settlements1, id, period))) %>% 
+  transmute(id, 
+            period,
+            value = mean_slope, 
+            variable = "slope") %>% 
+  bind_rows(res1_long)
+
+res2_long <- mean_slope %>% 
+  select(-ID, -reg) %>% 
+  inner_join(st_drop_geometry(select(settlements2, id, period))) %>% 
+  transmute(id, 
+            period,
+            value = mean_slope, 
+            variable = "slope") %>% 
+  bind_rows(res2_long)
+```
+
+```{r aspect-clean}
+rm(list = c("dem_slope", "slope", "mean_slope"))
+gc()
+```
+
+<!--
 ```{r dem-data}
@@
 ```
+-->
+
+## Exploring aspect
+
+Aspect is calculated from the DEM, 0 means the slope is facing North, 
+90 East, 180 South and 270 West. The settlements are buffered 
+by 200 meters and the median aspect is taken.
+The vicinity of settlements is thus of interest, not the given point.
+
+In both Morava river drainage basin and Eastern part of 
+Bohemia, median aspect is normally distributed.
+
+```{r aspect-norm}
+shapiro.test(filter(med_asp, reg == "B")$med_asp)
+shapiro.test(filter(med_asp, reg == "M")$med_asp)
+```
+
+```{r aspect-plot-reg, fig.cap="Histogram of aspect distibution on polar coordinates."}
+
+# polar plot
+med_asp %>% 
+  ggplot(aes(med_asp)) + 
+  geom_histogram(color = "black", fill = "white") +
+  scale_x_continuous(limits = c(0, 359), 
+                     breaks = c(0, 90, 180, 270)) +
+  coord_polar(start = 0) + 
+  facet_wrap(vars(reg)) +
+  theme_bw() +
+  labs(x = "median aspect (°)")
+```
+
+```{r aspect-plot-reg-p2}
+med_asp %>% 
+  inner_join(settlements2, by = "id") %>% 
+  ggplot(aes(med_asp)) + 
+  geom_histogram(color = "black", fill = "white") +
+  scale_x_continuous(limits = c(0, 360), 
+                     breaks = c(0, 90, 180, 270, 360)) +
+  facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+  theme_bw() +
+  labs(x = "median aspect (°)")
+```
+
+```{r aspect-clean}
+rm(list = c("dem_asp", "asp", "med_asp"))
+```
+
+## TPI
+
+> Terrain / Topographic roughness index
+
+```{r tpi}
+dem_tpi <- terra::terrain(dem_terra, v = "TPI")
+
+# tpi in the vicinity of settlements - 2km
+tpi <- terra::extract(dem_tpi, 
+                      terra::vect(st_buffer(set_spat, dist = 2e3)))
+
+# derive median aspect
+mean_slope <- slope %>% 
+  tibble() %>% 
+  group_by(ID) %>% 
+  summarise(mean_slope = mean(slope)) %>% 
+  bind_cols(id = set_spat$id) %>% 
+  mutate(reg = str_extract(id, "^."))
+
+# shapiro.test(filter(mean_slope, reg == "B")$mean_slope)
+# hist(filter(mean_slope, reg == "B")$mean_slope)
+# qqnorm(filter(mean_slope, reg == "B")$mean_slope)
+# 
+# shapiro.test(filter(mean_slope, reg == "M")$mean_slope)
+# hist(filter(mean_slope, reg == "M")$mean_slope)
+# qqnorm(filter(mean_slope, reg == "M")$mean_slope)
+
+# mean_slope %>% 
+#   inner_join(settlements2, by = "id") %>% 
+#   ggplot(aes(mean_slope)) + 
+#   geom_histogram(color = "black", fill = "white") +
+#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+
+# mean_slope %>% 
+#   inner_join(settlements2, by = "id") %>% 
+#   ggplot(aes(mean_slope, period_label)) + 
+#   geom_boxplot(color = "black", fill = "white") +
+#   facet_wrap(vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "slope (°)")
+```
 
@@
 
+<!--
 # Settlement 'continuity'
@@
 ```
+-->
 
@@
   "line_water" = "Along water line",
-  "cont" = "Settlement continuity",
+  # "cont" = "Settlement continuity",
   "hydro_kde" = "Watercourse density",
@@
   plot_nested_pca("period_label")
+
+# ggsave(here::here("analysis/paper/temp/pca_trad_1.png"), width = 10, height = 21)
 ```
@@
   plot_nested_pca("period_label", c("PC3", "PC4"))
+
+# ggsave(here::here("analysis/paper/temp/pca_trad_2.png"), width = 10, height = 14)
 ```
@@
   plot_nested_pca("period_label")
+
+ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
 ```
@@
   plot_nested_pca("period_label", c("PC3", "PC4"))
+
+ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
 ```
@@
 
+```{r package-mclust}
+library(mclust)
+```
+
 ```{r mclust2-neo}
@@
 
-mb_clust2 %>% pull(classif)
-
-mb_clust2 %>%
-    select(classif) %>% 
-    ungroup(c(reg, neo)) %>%
-  unnest() %>% View()
-    mutate(g = str_c(reg, g))
-
-mb_classif2_neoB <- mb_clust2 %>% mbc_classify("NeoB")
-mb_classif2_neoC <- mb_clust2 %>% mbc_classify("NeoC")
+x <- mb_clust2 %>%
+  dplyr::select(reg, neo, classif) %>% 
+  tidyr::unnest() %>% 
+  dplyr::ungroup(c(reg, neo)) %>% 
+  dplyr::mutate(g = str_c(reg, g))
+
+set_spat %>% 
+  inner_join(x, by = "id") %>% 
+  ggplot() +
+  geom_sf(aes(color = g)) +
+  facet_grid(vars(g), vars(neo))
+
+```
+
+mb_classif2_neoB <- mb_clust2 %>% mbc_classify(neo = "NeoB")
+mb_classif2_neoC <- mb_clust2 %>% mbc_classify(neo = "NeoC")
 

### 2022-03-23 analysis/traditions/traditions.Rmd
add new docs edit orig paper

--- 2022-03-23 analysis/groups/groups.Rmd
@@
 ---
-title: "SI: Settlements / Pottery groups"
+title: "SI: Settlements / Pottery traditions"
 author: Petr Pajdla
@@
 
-**This document describes results for Pottery groups.**
+**This document describes results for Pottery traditions.**
 
@@
 ```{r tab-set-nr}
-set_base$period2 %>%
+set_base$period1 %>%
   table_nr_set()
@@
 ```{r map-set-nr, fig.cap = 'Settlements across time periods.'}
-settlements2 %>%
+settlements1 %>%
   plot_nr_set()
@@
 ```{r map-kde, fig.cap = 'Kernel density estimation for different periods.'}
-settlements2 %>% 
+settlements1 %>% 
   plot_kde()
@@
 ```{r set-kde}
-kde <- settlements2 %>% 
+kde <- settlements1 %>% 
   est_kde() %>% 
@@
 ```{r alt-data}
-alt <- settlements2 %>% 
+alt <- settlements1 %>% 
   select(id, altitude) %>% 
@@
 ```{r alt}
-settlements2 %>% 
+settlements1 %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                              "slope_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% settlements1$id) %>% 
   mutate(variable = "slope")
@@
 slope %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(settlements1, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                            "tpi_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% settlements1$id) %>% 
   mutate(variable = "tpi")
@@
 tpi %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(settlements1, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
                              "hydro_at_points.csv")) %>% 
-  filter(id %in% settlements2$id) %>% 
+  filter(id %in% settlements1$id) %>% 
   mutate(variable = "hydro") %>% 
@@
 hydro %>% 
-  left_join(settlements2, by = "id") %>% 
+  left_join(settlements1, by = "id") %>% 
   mutate(reg = str_extract(id, "^.")) %>% 
@@
 
-along_lines <- along_lines$groups %>% 
+along_lines <- along_lines$traditions %>% 
   mutate(chrono = str_remove(chrono, ".$")) %>% 
@@
   mutate(rm = str_to_lower(rm)) %>% 
-  filter(id %in% settlements2$id)
+  filter(id %in% settlements1$id)
 
@@
 
-rm_dist_rel <- set_base$period2 %>% 
+rm_dist_rel <- set_base$period1 %>% 
   filter_rel_dist_rm()
@@
   bind_rows(hydro) %>% 
-  left_join(set_base$period2) %>% 
+  left_join(set_base$period1) %>% 
   select(id, variable, value, period) %>% 
@@
   bind_rows(select(kde, -reg)) %>% 
-  filter(id %in% set_base$period2$id) %>% 
+  filter(id %in% set_base$period1$id) %>% 
   group_by(variable, id, period) %>% 
@@
 ```{r spat-out}
-# set_spat %>% 
-#   select(id) %>% 
-#   inner_join(classif, by = "id") %>% 
-#   write_sf(here::here("analysis/groups/groups.geojson"), delete_layer = TRUE)
+set_spat %>% 
+  select(id) %>% 
+  inner_join(classif, by = "id") %>% 
+  write_sf(here::here("analysis/traditions/traditions.geojson"), delete_layer = TRUE)
 ```

### 2022-03-23 analysis/code/fuubar/lines.R
update gitignore clean mistake

@@ -1,160 +1,0 @@
-# Project "Settlements"
-# Script nr.
-# ARRANGEMENT ON LINES
-# author: Petr Pajdla
-#
-
-library(tidyverse)
-library(sf)
-
-derived_data <- "analysis/data/derived_data"
-path_maps <- here::here(derived_data, "maps/")
-
-mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
-
-# database
-set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
-
-set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"),
-                    quiet = TRUE)
-
-chrono_labels <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
-
-# # settlements spatial by periods
-# set_spat_1 <- set_spat %>%
-#   right_join(set_base$period1) %>%
-#   mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
-#
-# set_spat_2 <- set_spat %>%
-#   right_join(set_base$period2) %>%
-#   mutate(reg = if_else(str_detect(id, "^B"), "b", "m"))
-
-# arrangement on lines
-ids <- set_base$base %>% pull(id)
-names(ids) <- set_base$base %>% pull(site)
-
-add_id <- function(x) {
-  x %>%
-    filter(site %in% set_base$base$site) %>%
-    mutate(id = unname(ids[site]))
-}
-
-pivot_terrain_and_water <- function(x) {
-  x %>%
-    pivot_longer(c(starts_with("w"), starts_with("t")), values_to = "line_len") %>%
-    filter(!is.na(line_len)) %>%
-    separate(name, into = c("type", "chrono"), sep = "_") %>%
-    group_by(type)
-}
-
-pivot_period <- function(x) {
-  x %>%
-    pivot_longer(cols = starts_with("p"), names_to = "period") %>%
-    filter(value) %>%
-    select(-value) %>%
-    mutate(period_label = chrono_labels$periods[period],
-           period_label = factor(period_label, levels = chrono_labels$periods))
-}
-
-# east part of bohemia
-# traditions
-b1 <- readxl::read_excel(
[…]

### 2022-04-27 R/pca.R
add wrapper to summarise factor loadings

@@ -74,7 +74,7 @@ plot_nested_pca <- function(x, var, pc = c("PC1", "PC2")) {
 #'
 #' @param x Nested PCA object.
 #'
-#' @return Nested object with hierarchcal clustering results and dendrogram.
+#' @return Nested object with hierarchical clustering results and dendrogram.
 #' @export
 #'
 #' @examples
@@ -97,4 +97,33 @@ hclust_plot_dendro <- function(x, vars) {
     pull(dendro)
 
   (plots[[3]] / plots[4]) | (plots[[2]] / plots[[1]])
+}
+
+#' Summarise variable loadings
+#'
+#' @param x Nested PCA object.
+#' @param prob A threshold of cumulative variability explained.
+#'     Only PCs that cumulatively explain at least the given value are kept.
+#'     Defaults to 0.75.
+#'
+#' @return A tibble with variable loadings summarised across different PCAs.
+#' @export
+#'
+#' @examples
+summarise_variables <- function(x, prob = 0.75) {
+  x %>%
+    mutate(
+      # select only PCs that cumulatively explain given variability
+      overProb = purrr::map(sdev, \(x) x[(cumsum(x[, "var_prop"]) >= prob), ]$pc[1]),
+      overProb = purrr::map(overProb, \(x) 1:x),
+      # select PCs
+      rotation2 = purrr::map2(rotation, overProb, \(x, y) select(x, c("column", paste0("PC", y))))
+    ) %>%
+    select(rotation2) %>%
+    unnest(cols = c(rotation2)) %>%
+    ungroup() %>%
+    select(-period) %>%
+    group_by(reg, column) %>%
+    mutate(across(where(is.numeric), abs)) %>%
+    summarise(across(where(is.numeric), sum, na.rm = TRUE))
 }

### 2022-04-27 analysis/groups/groups.Rmd
add separate reports for various time resolutions

@@
 ```{r nested-pca}
-# traditions
 pca_per <- res_wide %>%
@@
 pca_per %>% plot_nested_pca(var = "period")
+```
+
+```{r pca-summations}
+loadings_sum <- pca_per %>% 
+  summarise_variables()
+```
+
+```{r pca-summations-hm, eval=FALSE, echo=FALSE}
+# loadings_sum_m <- as.matrix(loadings_sum[, paste0("PC", 1:(ncol(loadings_sum) - 1))])
+# rownames(loadings_sum_m) <- loadings_sum$column
+# heatmap(loadings_sum_m)
+```
+```{r pca-summations-tb}
+loadings_sum %>% 
+  mutate(r = (PC1 + PC2 + PC3 + PC4 + PC5)/(ncol(loadings_sum) - 1)) %>% 
+  select(column, r) %>% 
+  arrange(reg, desc(r)) %>% 
+  knitr::kable()
 ```

### 2022-04-27 analysis/code/results_to_gd.R
do not update results tables

@@ -11,31 +11,37 @@ library(googledrive)
 
 # upload results to GD -----------------------------------------------------
 
-# paths
-dt_result <- "analysis/data/derived_data/results"
-gd_path <- "~/settlements/data/results/"
-
-now <- paste("Updated", Sys.time())
-write_lines(now, here(dt_result, "README"), append = TRUE)
-
-# delete previous files
-drive_trash(filter(drive_ls(gd_path), str_detect(name, "^results.$")))
-
-# list files to upload
-files_local <- list.files(dt_result, pattern = "csv$", full.names = TRUE)
-readme <- list.files(dt_result, pattern = "^README$", full.names = TRUE)
-
-drive_results <- map(files_local, ~ drive_upload(.x, gd_path,
-                                                 type = "spreadsheet",
-                                                 overwrite = TRUE))
-drive_readme <- drive_upload(readme, gd_path, overwrite = TRUE)
+# # paths
+# dt_result <- "analysis/data/derived_data/results"
+# gd_path <- "~/settlements/data/results/"
+#
+# now <- paste("Updated", Sys.time())
+# write_lines(now, here(dt_result, "README"), append = TRUE)
+#
+# # delete previous files
+# drive_trash(filter(drive_ls(gd_path), str_detect(name, "^results.$")))
+#
+# # list files to upload
+# files_local <- list.files(dt_result, pattern = "csv$", full.names = TRUE)
+# readme <- list.files(dt_result, pattern = "^README$", full.names = TRUE)
+#
+# drive_results <- map(files_local, ~ drive_upload(.x, gd_path,
+#                                                  type = "spreadsheet",
+#                                                  overwrite = TRUE))
+# drive_readme <- drive_upload(readme, gd_path, overwrite = TRUE)
 
 
 # upload SI to GD ---------------------------------------------------------
 
-si <- here("analysis/paper/paper.html")
+# si <- here("analysis/paper/paper.html")
+neo <- here("analysis/neo/neo.html")
+traditions <- here("analysis/traditions/traditions.html")
+groups <- here("analysis/groups/groups.html")
 
-drive_si <- drive_upload(si, "~/settlements/SI/", overwrite = TRUE)
+# drive_si <- drive_upload(si, "~/settlements/SI/", overwrite = TRUE)
+drive_neo <- drive_upload(neo, "~/settlements/SI/", overwrite = TRUE)
[…]

### 2022-06-23 R/minions.R
fix code

@@ -78,6 +78,8 @@ relabel_variable <- function(x) {
     column == "rm_dist_chipped" ~ "d. ch.",
     column == "line_water" ~ "w. line",
     column == "rm_dist_polished" ~ "d. pol.",
+    column == "hydro" ~ "hydro",
+    column == "tpi" ~ "tpi"
   ))
 }
 

### 2022-06-23 R/pca.R

@@ -125,5 +125,5 @@ summarise_variables <- function(x, prob = 0.75) {
     select(-period) %>%
     group_by(reg, column) %>%
     mutate(across(where(is.numeric), abs)) %>%
-    summarise(across(where(is.numeric), sum, na.rm = TRUE))
+    summarise(across(where(is.numeric), mean, na.rm = TRUE))
 }

### 2022-06-23 analysis/code/dem_slope.R
site neigborhood unified to 0.3 km

@@ -9,7 +9,7 @@ library(sf)
 devtools::load_all()
 
 # size of a buffer zone around the settlement (meters)
-buffer_dist <- units::set_units(0.5, "km")
+buffer_dist <- units::set_units(0.3, "km")
 
 # data --------------------------------------------------------------------
 
@@ -56,6 +56,11 @@ mean_slope %>%
   theme_bw() +
   coord_flip() +
   labs(x = "slope (°)")
+
+ggplot() +
+  geom_sf(data = st_as_sf(mean_slope), aes(color = value),
+          size = 2, alpha = 0.8) +
+  theme_void()
 
 # mean_slope %>%
 #   mutate(reg = stringr::str_extract(id, ".")) %>%

### 2022-06-23 analysis/code/dem_tpi.R

@@ -9,7 +9,7 @@ library(sf)
 devtools::load_all()
 
 # size of a buffer zone around the settlement (meters)
-buffer_dist <- units::set_units(0.2, "km")
+buffer_dist <- units::set_units(0.3, "km")
 
 # data --------------------------------------------------------------------
 
@@ -45,6 +45,15 @@ mean_tpi %>%
   theme_bw() +
   coord_flip() +
   labs(x = "TPI (terrain position index)")
+
+mean_tpi %>%
+  st_as_sf() %>%
+  ggplot() +
+  geom_sf(aes(color = value), size = 1.8)
+
+mean_tpi %>%
+  ggplot(aes(value)) +
+  geom_histogram()
 
 # mean_slope %>%
 #   mutate(reg = stringr::str_extract(id, ".")) %>%

### 2022-06-23 analysis/code/hydro.R

@@ -9,7 +9,7 @@ library(sf)
 library(terra)
 
 # size of a buffer zone around the settlement (meters)
-buffer_dist <- units::set_units(0.2, "km")
+buffer_dist <- units::set_units(0.3, "km")
 
 # data --------------------------------------------------------------------
 

### 2022-06-23 analysis/groups/groups.Rmd
analysis for pottery groups

@@
 ```{r plot-kde}
-kde %>% ggplot(aes(period, value)) +
+kde %>% 
+  add_period_label() %>% 
+  mutate(reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>% 
+  add_period_label() %>% 
+  ggplot(aes(period_label, value)) +
   geom_violin() +
@@
   theme_bw() +
-  labs(y = "KDE")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "KDE (Density of settlements per square km)", x = "Period")
+
+# ggsave(here::here("groups_kde.svg"), width = 10, height = 5)
 ```
@@
 settlements2 %>% 
-  mutate(reg = str_extract(id, "^.")) %>% 
-  ggplot(aes(period, altitude)) +
+  mutate(reg = str_extract(id, "^."),
+         reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>% 
+  add_period_label() %>% 
+  ggplot(aes(period_label, altitude)) +
   geom_violin() +
@@
   theme_bw() +
-  labs(y = "Altitude (m)")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "Altitude (m)", x = "Period")
+
+# ggsave(here::here("groups_altitude.svg"), width = 10, height = 5)
 ```
@@
   left_join(settlements2, by = "id") %>% 
-  mutate(reg = str_extract(id, "^.")) %>% 
-  ggplot(aes(period, value)) +
+  mutate(reg = str_extract(id, "^."),
+         reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>% 
+  add_period_label() %>% 
+  ggplot(aes(period_label, value)) +
   geom_violin() +
@@
   theme_bw() +
-  labs(y = "Slope (°)")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "Slope (°)", x = "Period")
+
+# ggsave(here::here("groups_slope.svg"), width = 10, height = 5)
 ```
@@
   theme_bw() +
-  labs(y = "TPI")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "TPI", x = "Period")
+```
+
+```{r}
+# ggsave(here::here("groups_tpi.svg"), width = 10, height = 5)
 ```
@@
   theme_bw() +
-  labs(y = "Hydrology KDE")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "Hydrology KDE", x = "Period")
+```
+```{r}
+# ggsave(here::here("groups_hydro.svg"), width = 10, height = 5)
 ```
@@
 
+```{r}
+rm_dist_fin %>% 
+  filter(variable == "rm_dist_chipped") %>% 
+  add_period_label() %>% 
+  mutate(reg = str_extract(id, "^."),
+         reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>% 
+  ggplot(aes(period_label, value)) +
+  geom_violin() +
+  geom_boxplot(alpha = 0.4, width = 0.2) +
+  geom_jitter(alpha = 0.05, width = 0.4) +
+  facet_wrap(vars(reg)) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
+  labs(y = "Distance to lithic raw. mat. (km)", x = "Period")
+
+# ggsave(here::here("groups_distSI.svg"), width = 10, height = 5)
+```
+
+<!--- ====================================================================== -->
+
+# Continuity
+
+```{r continuity-calc}
+cont <- set_base$period2 %>%
+  select(-period_label, -chrono, - facet) %>%
+  distinct(id, period) %>%
+  mutate(pres = 1L) %>%
+  pivot_wider(names_from = period, values_from = pres, values_fill = 0L) %>%
+  pivot_longer(starts_with("p"), names_to = "period") %>%
+  mutate(period = factor(period, levels = levels(set_base$period2$period))) %>%
+  arrange(id, period) %>%
+  group_by(id) %>%
+  mutate(prev = if_else(value == 1, lag(value), 0L),
+         prev = if_else(is.na(prev), 0L, prev)) %>%
+  ungroup(id) %>%
+  select(-value) %>%
+  mutate(variable = "cont") %>%
+  rename(value = prev)
+
+n_set <- set_base$period2 %>% 
+  add_region() %>% 
+  group_by(period_label, reg) %>% 
+  summarise(sum = n())
+
+cont %>%
+  filter(variable == "cont") %>%
+  add_period_label() %>%
+  add_region() %>%
+  group_by(period_label, reg) %>%
+  summarise(n = sum(value, na.rm = TRUE)) %>%
+  full_join(n_set) %>%
+  mutate(n = if_else(is.na(n), 0L, n),
+         perc = (n / sum) * 100,
+         reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>%
+  ggplot(aes(period_label, perc)) +
+  geom_col(color = "black", fill = "white", 
+           position = "dodge", show.legend = FALSE) +
+  facet_wrap(vars(reg), ncol = 1) +
+  labs(x = "Period", y = "Continuity of settlements (%)") +
+  theme_bw()
+
+# ggsave(here::here("groups_continuity.svg"), width = 10, height = 5)
+```
+
 <!--- ====================================================================== -->
@@
 pca_per %>% plot_nested_pca(var = "period")
+# ggsave(here::here("analysis/groups/pca.pdf"), width = 21, height = 35)
+```
+
+```{r pca-plot2}
+pca_per %>% plot_nested_pca(var = "period", pc = c("PC3", "PC4"))
+# ggsave(here::here("analysis/groups/pca2.pdf"), width = 21, height = 35)
 ```

### 2022-06-23 analysis/traditions/traditions.Rmd
add separate reports for various time resolutions
analysis for pottery traditions

@@
 
+```{r pca-summations}
+loadings_sum <- pca_per %>% 
+  summarise_variables()
+```
+
+```{r pca-summations-tb}
+loadings_sum %>% 
+  mutate(r = (PC1 + PC2 + PC3 + PC4 + PC5)/(ncol(loadings_sum) - 1)) %>% 
+  select(column, r) %>% 
+  arrange(reg, desc(r)) %>% 
+  knitr::kable()
+```
+
 ## Groups in the data
@@
 ```{r spat-out}
-set_spat %>% 
-  select(id) %>% 
-  inner_join(classif, by = "id") %>% 
-  write_sf(here::here("analysis/traditions/traditions.geojson"), delete_layer = TRUE)
+# set_spat %>% 
+#   select(id) %>% 
+#   inner_join(classif, by = "id") %>% 
+#   write_sf(here::here("analysis/traditions/traditions.geojson"), delete_layer = TRUE)
 ```

### 2022-06-23 analysis/neo/neo.Rmd
add separate reports for various time resolutions
analysis for Neolithic B and C

@@
 ```{r pca-plot}
-pca_per %>% plot_nested_pca(var = "period")
+pca_per %>% 
+  mutate(reg = if_else(reg == "B", "East Bohemia", "Morava river catchment")) %>% 
+  plot_nested_pca(var = "period")
+```
+
+```{r}
+# ggsave(here::here("neo_pca.svg"), width = 10, height = 7)
+```
+
+```{r pca-summations}
+loadings_sum <- pca_per %>% 
+  summarise_variables()
+```
+
+```{r pca-summations-tb}
+loadings_sum %>% 
+  mutate(r = (PC1 + PC2 + PC3 + PC4 + PC5)/(ncol(loadings_sum) - 1)) %>% 
+  select(column, r) %>% 
+  arrange(reg, desc(r)) %>% 
+  knitr::kable()
+```
+
+```{r}
+# loadings_sum %>% 
+#   mutate(r = (PC1 + PC2 + PC3 + PC4 + PC5)/(ncol(loadings_sum) - 1)) %>% 
+#   select(column, r) %>% 
+#   arrange(reg, desc(r)) %>% 
+#   write_csv(here::here("neo_faktory.csv"))
 ```

### 2022-06-23 analysis/code/continuity.R
update derived data, code for continuity plot

@@ -1,0 +1,48 @@
+cont <- set_base$period2 %>%
+  select(-period_label, -period, -chrono) %>%
+  distinct(id, facet) %>%
+  mutate(pres = 1L) %>%
+  pivot_wider(names_from = facet, values_from = pres, values_fill = 0L) %>%
+  pivot_longer(c(starts_with("TRB"), starts_with("SBK"), starts_with("LgK"), "Proto Ene."),
+               names_to = "facet") %>%
+  mutate(facet = factor(facet, levels = labs_chrono$facets)) %>%
+  arrange(id, facet) %>%
+  group_by(id) %>%
+  mutate(prev = if_else(value == 1, lag(value), 0L),
+         prev = if_else(is.na(prev), 0L, prev)) %>%
+  ungroup(id) %>%
+  select(-value) %>%
+  mutate(variable = "cont") %>%
+  rename(value = prev)
+
+n_set <- set_base$period2 %>%
+  select(-starts_with("period"), -chrono) %>%
+  add_region() %>%
+  group_by(reg, facet) %>%
+  summarise(sum = n())
+
+cont %>%
+  add_region() %>%
+  select(-variable) %>%
+  group_by(reg, facet) %>%
+  summarise(n = sum(value, na.rm = TRUE)) %>%
+  full_join(n_set) %>%
+  mutate(
+    n = if_else(is.na(n), 0L, n),
+    sum = if_else(is.na(sum), 0L, sum),
+    perc = (n / sum) * 100,
+    perc = if_else(perc == 0 | is.nan(perc), NA_real_, perc),
+    reg = if_else(reg == "B", "East Bohemia", "Morava river catchment"),
+    # chrono_label = labs_chrono$chrono2[chrono],
+    # chrono_label = factor(chrono_label, levels = labs_chrono$chrono2)
+    facet = factor(facet, levels = labs_chrono$facets)
+  ) %>%
+  ggplot(aes(facet, perc)) +
+  geom_col(color = "black", fill = "white",
+           position = "dodge", show.legend = FALSE) +
+  facet_wrap(vars(reg), ncol = 1) +
+  labs(x = "Pottery group", y = "Continuity of settlements (%)") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = -90, hjust = 0))
+
+ggsave(here::here("groups_chrono_continuity2.svg"), width = 7, height = 5)

### 2022-06-23 analysis/paper/paper.Rmd
update whole paper fix lua filters

@@
   echo = FALSE,
-  cache = FALSE,
+  cache = TRUE,
   comment = "#>",
@@
 derived_data <- "analysis/data/derived_data"
+input_data <- "analysis/data/input_data"
 path_maps <- here::here(derived_data, "maps/")
@@
 # labels
-labs_chrono <- read_rds(here::here(derived_data, "chrono_labels.RDS"))
+labs_chrono <- read_rds(here::here(input_data, "chrono_labels.RDS"))
 
 # database
-set_base <- read_rds(here::here(derived_data, "settlements.RDS"))
-
-set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"), 
+set_base <- read_rds(here::here(input_data, "settlements.RDS"))
+
+set_spat <- st_read(here::here(input_data, "settlements_sf.geojson"), 
                     quiet = TRUE)
@@
 
-<!-- ```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."} -->
-<!-- rm_dist %>% -->
-<!--   ggplot(aes(x = min, y = rm, fill = region)) + -->
-<!--   geom_violin() + -->
-<!--   scale_fill_brewer(palette = "Accent") + -->
-<!--   facet_wrap(~type, scales = "free_y") + -->
-<!--   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") + -->
-<!--   theme_minimal() + -->
-<!--   theme(legend.position = "bottom",  -->
-<!--         panel.border = element_rect(color = "black", fill = NA)) -->
-<!-- ``` -->
+```{r plot-rm-dist-min, fig.cap="Min. distance to raw material sources."}
+# rm_dist %>%
+#   ggplot(aes(x = min, y = rm, fill = region)) +
+#   geom_violin() +
+#   scale_fill_brewer(palette = "Accent") +
+#   facet_wrap(~type, scales = "free_y") +
+#   labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
+#   theme_minimal() +
+#   theme(legend.position = "bottom",
+#         panel.border = element_rect(color = "black", fill = NA))
+```
 
@@
 ```{r dem}
-dem_terra <- terra::rast(here::here(derived_data, "/dem/dem_aster_sjtsk_stars.tif"))
+dem_terra <- terra::rast(here::here(derived_data, "archive/dem/dem_aster_sjtsk_stars.tif"))
 ```
@@
 
-<!--
 ```{r dem-data}
-dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))
-
-# bind to results
-res1_long <- set_base$period1 %>% 
-prep_dem_res() %>% 
-bind_rows(res1_long)
-
-res2_long <- set_base$period2 %>% 
-select(-facet) %>% 
-prep_dem_res() %>% 
-bind_rows(res2_long)
+# dem_data <- read_csv(here::here(derived_data, "dem/vals_dem.csv"))
+# 
+# # bind to results
+# res1_long <- set_base$period1 %>% 
+# prep_dem_res() %>% 
+# bind_rows(res1_long)
+# 
+# res2_long <- set_base$period2 %>% 
+# select(-facet) %>% 
+# prep_dem_res() %>% 
+# bind_rows(res2_long)
 ```
@@
 ```{r altitude-to-res}
-altitude <- set_spat %>% st_drop_geometry() %>% 
-as_tibble() %>% 
-select(id, altitude)
-
-# bind to results
-res1_long <- set_base$period1 %>% 
-prep_alt_res() %>% 
-bind_rows(res1_long)
-
-res2_long <- set_base$period2 %>% 
-select(-facet) %>% 
-prep_alt_res() %>% 
-bind_rows(res2_long)
-
+# altitude <- set_spat %>% st_drop_geometry() %>% 
+# as_tibble() %>% 
+# select(id, altitude)
+# 
+# # bind to results
+# res1_long <- set_base$period1 %>% 
+# prep_alt_res() %>% 
+# bind_rows(res1_long)
+# 
+# res2_long <- set_base$period2 %>% 
+# select(-facet) %>% 
+# prep_alt_res() %>% 
+# bind_rows(res2_long)
 ```
@@
 ```{r dem-analysis-correlation, fig.cap='Correlation between altitude derived from DEM and altitude reading in the topographic map.'}
-alt_plot <- set_spat %>% st_drop_geometry() %>% 
-as_tibble() %>% 
-select(id, altitude) %>% 
-left_join(dem_data)
-
-alt_cor <- alt_plot %>% 
-summarise(cor = cor(altitude, dem)) %>% 
-pull(cor) %>% 
-round(2)
-
-alt_plot %>% 
-ggplot(aes(altitude, dem)) +
-geom_point() +
-theme_minimal() +
-annotate(geom = "text", x = 175, y = 525, 
-label = paste("Cor. ", alt_cor))
-```
--->
+# alt_plot <- set_spat %>% st_drop_geometry() %>% 
+# as_tibble() %>% 
+# select(id, altitude) %>% 
+# left_join(dem_data)
+# 
+# alt_cor <- alt_plot %>% 
+# summarise(cor = cor(altitude, dem)) %>% 
+# pull(cor) %>% 
+# round(2)
+# 
+# alt_plot %>% 
+# ggplot(aes(altitude, dem)) +
+# geom_point() +
+# theme_minimal() +
+# annotate(geom = "text", x = 175, y = 525, 
+# label = paste("Cor. ", alt_cor))
+```
 
@@
 ```{r aspect-norm}
-shapiro.test(filter(med_asp, reg == "B")$med_asp)
-shapiro.test(filter(med_asp, reg == "M")$med_asp)
+# shapiro.test(filter(med_asp, reg == "B")$med_asp)
+# shapiro.test(filter(med_asp, reg == "M")$med_asp)
 ```
@@
 ```{r aspect-plot-reg, fig.cap="Histogram of aspect distibution on polar coordinates."}
-
-# polar plot
-med_asp %>% 
-  ggplot(aes(med_asp)) + 
-  geom_histogram(color = "black", fill = "white") +
-  scale_x_continuous(limits = c(0, 359), 
-                     breaks = c(0, 90, 180, 270)) +
-  coord_polar(start = 0) + 
-  facet_wrap(vars(reg)) +
-  theme_bw() +
-  labs(x = "median aspect (°)")
+# # polar plot
+# med_asp %>% 
+#   ggplot(aes(med_asp)) + 
+#   geom_histogram(color = "black", fill = "white") +
+#   scale_x_continuous(limits = c(0, 359), 
+#                      breaks = c(0, 90, 180, 270)) +
+#   coord_polar(start = 0) + 
+#   facet_wrap(vars(reg)) +
+#   theme_bw() +
+#   labs(x = "median aspect (°)")
 ```
@@
 ```{r aspect-plot-reg-p2}
-med_asp %>% 
-  inner_join(settlements2, by = "id") %>% 
-  ggplot(aes(med_asp)) + 
-  geom_histogram(color = "black", fill = "white") +
-  scale_x_continuous(limits = c(0, 360), 
-                     breaks = c(0, 90, 180, 270, 360)) +
-  facet_grid(vars(period_label), vars(reg), scales = "free_y") +
-  theme_bw() +
-  labs(x = "median aspect (°)")
-```
-
-```{r aspect-clean}
+# med_asp %>% 
+#   inner_join(settlements2, by = "id") %>% 
+#   ggplot(aes(med_asp)) + 
+#   geom_histogram(color = "black", fill = "white") +
+#   scale_x_continuous(limits = c(0, 360), 
+#                      breaks = c(0, 90, 180, 270, 360)) +
+#   facet_grid(vars(period_label), vars(reg), scales = "free_y") +
+#   theme_bw() +
+#   labs(x = "median aspect (°)")
+```
+
+```{r aspect-clean2}
 rm(list = c("dem_asp", "asp", "med_asp"))
@@
 ```{r tpi}
-dem_tpi <- terra::terrain(dem_terra, v = "TPI")
-
-# tpi in the vicinity of settlements - 2km
-tpi <- terra::extract(dem_tpi, 
-                      terra::vect(st_buffer(set_spat, dist = 2e3)))
-
-# derive median aspect
-mean_slope <- slope %>% 
-  tibble() %>% 
-  group_by(ID) %>% 
-  summarise(mean_slope = mean(slope)) %>% 
-  bind_cols(id = set_spat$id) %>% 
-  mutate(reg = str_extract(id, "^."))
+# dem_tpi <- terra::terrain(dem_terra, v = "TPI")
+# 
+# # tpi in the vicinity of settlements - 2km
+# tpi <- terra::extract(dem_tpi, 
+#                       terra::vect(st_buffer(set_spat, dist = 2e3)))
+# 
+# # derive median aspect
+# mean_slope <- slope %>% 
+#   tibble() %>% 
+#   group_by(ID) %>% 
+#   summarise(mean_slope = mean(slope)) %>% 
+#   bind_cols(id = set_spat$id) %>% 
+#   mutate(reg = str_extract(id, "^."))
 
@@
 
-<!--
 # Settlement 'continuity'
@@
 ```
--->
 
@@
 ```{r res-violin-out}
-ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
-       width = 9, height = 12)
-ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
-       width = 9, height = 12)
+# ggsave(here::here("analysis/figures", "res1_violin.png"), res1_violin, 
+#        width = 9, height = 12)
+# ggsave(here::here("analysis/figures", "res2_violin.png"), res2_violin, 
+#        width = 9, height = 12)
 ```
@@
 ```{r res-line-out}
-ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
-       width = 7, height = 7)
-ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
-       width = 7, height = 7)
+# ggsave(here::here("analysis/figures", "res1_line.png"), res1_line, 
+#        width = 7, height = 7)
+# ggsave(here::here("analysis/figures", "res2_line.png"), res2_line, 
+#        width = 7, height = 7)
 ```
@@
 
-<!-- ```{r pca1-all} -->
-<!-- # fit PCA -->
-<!-- pca1_all_fit <- res1_wide %>%  -->
-<!--   # select(-cont) %>%  -->
-<!--   select(where(is.numeric)) %>%  -->
-<!--   scale(center = TRUE, scale = TRUE) %>%  -->
-<!--   prcomp() -->
-
-<!-- # augmented fitted values -->
-<!-- pca1_all_aug <- pca1_all_fit %>%  -->
-<!--   broom::augment(res1_wide) %>%  -->
-<!--   select(-".rownames") %>%  -->
-<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
-<!--   settlements::add_region() %>%  -->
-<!--   settlements::add_period_label() %>%  -->
-<!--   settlements::add_neo() -->
-
-<!-- # rotation matrix -->
-<!-- pca1_all_rot <- pca1_all_fit %>% -->
-<!--   broom::tidy(matrix = "rotation") %>% -->
-<!--   mutate(PC = str_c("PC", PC)) %>% -->
-<!--   pivot_wider(names_from = PC) -->
-
-<!-- # var explained -->
-<!-- pca1_all_var_exp <- pca1_all_fit %>%  -->
-<!--   broom::augment(res1_wide) %>% -->
-<!--   summarise(across(contains("PC"), var)) %>%  -->
-<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
-<!--   mutate(var_exp = variance/sum(variance), -->
-<!--          cum_var_exp = cumsum(var_exp), -->
-<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
-<!--   select(pc, var_exp, cum_var_exp) -->
-
-<!-- # PC cum var explained over 0.9 -->
-<!-- pca1_all_var_exp_over <- pca1_all_var_exp %>%  -->
-<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
-<!--   pull(n) %>%  -->
-<!--   sum() + 1 -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-var-exp-tab} -->
-<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
-<!--                                              "Prop. of variance",  -->
-<!--                                              "Cumulative prop.")) -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-loadings-plot} -->
-<!-- # PCA loadings -->
-<!-- pca1_all_loadings <- pca1_all_fit %>%  -->
-<!--   broom::tidy(matrix = "rotation") %>%  -->
-<!--   mutate(component = str_c("PC", PC), -->
-<!--          component = fct_reorder(component, PC)) %>%  -->
-<!--   filter(PC <= pca1_all_var_exp_over) %>%  -->
-<!--   group_by(PC) %>%  -->
-<!--   top_n(8, abs(value)) %>%  -->
-<!--   ungroup() %>%  -->
-<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
-<!--          column = tidytext::reorder_within(column, abs(value), component)) -->
-
-<!-- pca1_all_loadings %>%  -->
-<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
-<!--   geom_col(color = "black") + -->
-<!--   scale_fill_manual(values = c("white", "gray80")) + -->
-<!--   facet_wrap(vars(component), scales = "free_y") + -->
-<!--   tidytext::scale_y_reordered() + -->
-<!--   theme_bw() + -->
-<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
-<!-- ``` -->
-
-<!-- ```{r plot-pca1-results} -->
-<!-- plot_pca1_full <- function(x = "PC1", y = "PC2") { -->
-
-<!--   pcs <- c(x, y) -->
-
-<!--   pca1_all_plot <- pca1_all_aug %>%  -->
-<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
-<!--     geom_point(data = select(pca1_all_aug, -period_label),  -->
-<!--                alpha = 0.05, size = 0.4) + -->
-<!--     geom_point(aes(color = reg),  -->
-<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
-<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
-<!--     facet_grid(vars(period_label), vars(reg)) + -->
-<!--     theme_minimal() -->
-
-<!--   pca1_all_rot <-  pca1_all_rot %>% -->
-<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
-<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
-<!--     geom_text( -->
-<!--       aes(label = column), -->
-<!--       hjust = 1, nudge_x = -0.02, -->
-<!--       color = "#904C2F", -->
-<!--       size = 2.6 -->
-<!--     ) + -->
-<!--     theme_minimal() -->
-
-<!--   pca1_all_load <- pca1_all_loadings %>%  -->
-<!--     filter(component %in% pcs) %>%  -->
-<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
-<!--     geom_col(color = "black", show.legend = FALSE) + -->
-<!--     scale_fill_manual(values = c("white", "gray80")) + -->
-<!--     facet_wrap(vars(component), scales = "free_y") + -->
-<!--     tidytext::scale_y_reordered() + -->
-<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
-<!--     theme_minimal() -->
-
-<!--   pca1_all_plot / (pca1_all_rot | pca1_all_load) +  -->
-<!--     plot_layout(heights = c(6, 1)) -->
-<!-- } -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-plots-pc1-pc2} -->
-<!-- plot_pca1_full() -->
-<!-- ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-plots-pc3-pc4} -->
-<!-- plot_pca1_full("PC3", "PC4") -->
-<!-- ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-plots-pc5-pc6} -->
-<!-- plot_pca1_full("PC5", "PC6") -->
-<!-- ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca1-all-plots-pc7-pc8} -->
-<!-- plot_pca1_full("PC7", "PC8") -->
-<!-- ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ### Pottery groups -->
-
-<!-- ```{r pca2-all} -->
-<!-- # fit PCA -->
-<!-- pca2_all_fit <- res2_wide %>%  -->
-<!--   # select(-cont) %>%  -->
-<!--   select(where(is.numeric)) %>%  -->
-<!--   scale(center = TRUE, scale = TRUE) %>%  -->
-<!--   prcomp() -->
-
-<!-- # augmented fitted values -->
-<!-- pca2_all_aug <- pca2_all_fit %>%  -->
-<!--   broom::augment(res2_wide) %>%  -->
-<!--   select(-".rownames") %>%  -->
-<!--   rename_with(\(x) str_remove(x, ".fitted")) %>%  -->
-<!--   settlements::add_region() %>%  -->
-<!--   settlements::add_period_label() %>%  -->
-<!--   settlements::add_neo() -->
-
-<!-- # rotation matrix -->
-<!-- pca2_all_rot <- pca2_all_fit %>% -->
-<!--   broom::tidy(matrix = "rotation") %>% -->
-<!--   mutate(PC = str_c("PC", PC)) %>% -->
-<!--   pivot_wider(names_from = PC) -->
-
-<!-- # var explained -->
-<!-- pca2_all_var_exp <- pca2_all_fit %>%  -->
-<!--   broom::augment(res2_wide) %>% -->
-<!--   summarise(across(contains("PC"), var)) %>%  -->
-<!--   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%  -->
-<!--   mutate(var_exp = variance/sum(variance), -->
-<!--          cum_var_exp = cumsum(var_exp), -->
-<!--          pc = str_replace(pc, ".fitted", "")) %>%  -->
-<!--   select(pc, var_exp, cum_var_exp) -->
-
-<!-- # PC cum var explained over 0.9 -->
-<!-- pca2_all_var_exp_over <- pca2_all_var_exp %>%  -->
-<!--   mutate(n = !(cum_var_exp >= 0.9)) %>%  -->
-<!--   pull(n) %>%  -->
-<!--   sum() + 1 -->
-<!-- ``` -->
-
-<!-- <!-- -->
-<!-- ```{r} -->
-<!-- fuu <- res2_wide %>% mutate(across(where(is.numeric), \(x) x == 0)) %>%  -->
-<!-- select(where(is.logical)) %>%  -->
-<!-- rowSums() %>%  -->
-<!-- as_tibble() -->
-
-<!-- shit <- res2_wide %>% select(id, period) %>% bind_cols(fuu) %>% arrange(desc(value), id, period) %>%  -->
-<!-- filter(value == 8) %>% pull(id) %>% factor() %>% levels() -->
-
-<!-- settlements2 %>%  -->
-<!-- filter(id %in% shit) %>%  -->
-<!-- sf::st_transform(4326) %>%  -->
-<!-- leaflet::leaflet() %>%  -->
-<!-- leaflet::addTiles() %>%  -->
-<!-- leaflet::addMarkers() -->
-
-<!-- res2_wide %>% filter(id %in% shit) %>% View() -->
-<!-- ``` -->
-<!-- -->
-
-<!-- ```{r pca2-all-var-exp-tab} -->
-<!-- knitr::kable(pca1_all_var_exp, col.names = c("Comp.", -->
-<!--                                              "Prop. of variance",  -->
-<!--                                              "Cumulative prop.")) -->
-<!-- ``` -->
-
-<!-- ```{r pca2-all-loadings-plot} -->
-<!-- # PCA loadings -->
-<!-- pca2_all_loadings <- pca2_all_fit %>%  -->
-<!--   broom::tidy(matrix = "rotation") %>%  -->
-<!--   mutate(component = str_c("PC", PC), -->
-<!--          component = fct_reorder(component, PC)) %>%  -->
-<!--   filter(PC <= pca2_all_var_exp_over) %>%  -->
-<!--   group_by(PC) %>%  -->
-<!--   top_n(8, abs(value)) %>%  -->
-<!--   ungroup() %>%  -->
-<!--   mutate(cor = if_else(value > 0, "positive", "negative"), -->
-<!--          column = tidytext::reorder_within(column, abs(value), component)) -->
-
-<!-- pca2_all_loadings %>%  -->
-<!--   ggplot(aes(abs(value), column, fill = cor)) + -->
-<!--   geom_col(color = "black") + -->
-<!--   scale_fill_manual(values = c("white", "gray80")) + -->
-<!--   facet_wrap(vars(component), scales = "free_y") + -->
-<!--   tidytext::scale_y_reordered() + -->
-<!--   theme_bw() + -->
-<!--   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") -->
-<!-- ``` -->
-
-<!-- ```{r plot-pca2-results} -->
-<!-- plot_pca2_full <- function(x = "PC1", y = "PC2") { -->
-
-<!--   pcs <- c(x, y) -->
-
-<!--   pca2_all_plot <- pca2_all_aug %>%  -->
-<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
-<!--     geom_point(data = select(pca2_all_aug, -period_label),  -->
-<!--                alpha = 0.05, size = 0.4) + -->
-<!--     geom_point(aes(color = reg),  -->
-<!--                alpha = 0.8, size = 0.5, show.legend = FALSE) + -->
-<!--     scale_color_manual(values = mycol[c(1, 2)]) + -->
-<!--     facet_grid(vars(period_label), vars(reg)) + -->
-<!--     theme_minimal() -->
-
-<!--   pca2_all_rot <-  pca2_all_rot %>% -->
-<!--     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) + -->
-<!--     geom_segment(xend = 0, yend = 0, alpha = 0.4) + -->
-<!--     geom_text( -->
-<!--       aes(label = column), -->
-<!--       hjust = 1, nudge_x = -0.02, -->
-<!--       color = "#904C2F", -->
-<!--       size = 2.6 -->
-<!--     ) + -->
-<!--     theme_minimal() -->
-
-<!--   pca2_all_load <- pca2_all_loadings %>%  -->
-<!--     filter(component %in% pcs) %>%  -->
-<!--     ggplot(aes(abs(value), column, fill = cor)) + -->
-<!--     geom_col(color = "black", show.legend = FALSE) + -->
-<!--     scale_fill_manual(values = c("white", "gray80")) + -->
-<!--     facet_wrap(vars(component), scales = "free_y") + -->
-<!--     tidytext::scale_y_reordered() + -->
-<!--     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") + -->
-<!--     theme_minimal() -->
-
-<!--   pca2_all_plot / (pca2_all_rot | pca2_all_load) +  -->
-<!--     plot_layout(heights = c(6, 1)) -->
-<!-- } -->
-<!-- ``` -->
-
-<!-- ```{r pca2-all-plots-pc1-pc2} -->
-<!-- plot_pca2_full() -->
-<!-- ggsave(here::here("analysis/figures/pca2_full_pc12.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca2-all-plots-pc3-pc4} -->
-<!-- plot_pca2_full("PC3", "PC4") -->
-<!-- ggsave(here::here("analysis/figures/pca2_full_pc34.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca2-all-plots-pc5-pc6} -->
-<!-- plot_pca2_full("PC5", "PC6") -->
-<!-- ggsave(here::here("analysis/figures/pca2_full_pc56.png"), height = 14) -->
-<!-- ``` -->
-
-<!-- ```{r pca2-all-plots-pc7-pc8} -->
-<!-- plot_pca2_full("PC7", "PC8") -->
-<!-- ggsave(here::here("analysis/figures/pca2_full_pc78.png"), height = 14) -->
-<!-- ``` -->
+```{r pca1-all}
+# # fit PCA
+# pca1_all_fit <- res1_wide %>%
+#   # select(-cont) %>%
+#   select(where(is.numeric)) %>%
+#   scale(center = TRUE, scale = TRUE) %>%
+#   prcomp()
+# 
+# # augmented fitted values
+# pca1_all_aug <- pca1_all_fit %>%
+#   broom::augment(res1_wide) %>%
+#   select(-".rownames") %>%
+#   rename_with(\(x) str_remove(x, ".fitted")) %>%
+#   settlements::add_region() %>%
+#   settlements::add_period_label() %>%
+#   settlements::add_neo()
+# 
+# # rotation matrix
+# pca1_all_rot <- pca1_all_fit %>%
+#   broom::tidy(matrix = "rotation") %>%
+#   mutate(PC = str_c("PC", PC)) %>%
+#   pivot_wider(names_from = PC)
+# 
+# # var explained
+# pca1_all_var_exp <- pca1_all_fit %>%
+#   broom::augment(res1_wide) %>%
+#   summarise(across(contains("PC"), var)) %>%
+#   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%
+#   mutate(var_exp = variance/sum(variance),
+#          cum_var_exp = cumsum(var_exp),
+#          pc = str_replace(pc, ".fitted", "")) %>%
+#   select(pc, var_exp, cum_var_exp)
+# 
+# # PC cum var explained over 0.9
+# pca1_all_var_exp_over <- pca1_all_var_exp %>%
+#   mutate(n = !(cum_var_exp >= 0.9)) %>%
+#   pull(n) %>%
+#   sum() + 1
+```
+
+```{r pca1-all-var-exp-tab}
+# knitr::kable(pca1_all_var_exp, col.names = c("Comp.",
+#                                              "Prop. of variance",
+#                                              "Cumulative prop."))
+```
+
+```{r pca1-all-loadings-plot}
+# # PCA loadings
+# pca1_all_loadings <- pca1_all_fit %>%
+#   broom::tidy(matrix = "rotation") %>%
+#   mutate(component = str_c("PC", PC),
+#          component = fct_reorder(component, PC)) %>%
+#   filter(PC <= pca1_all_var_exp_over) %>%
+#   group_by(PC) %>%
+#   top_n(8, abs(value)) %>%
+#   ungroup() %>%
+#   mutate(cor = if_else(value > 0, "positive", "negative"),
+#          column = tidytext::reorder_within(column, abs(value), component))
+# 
+# pca1_all_loadings %>%
+#   ggplot(aes(abs(value), column, fill = cor)) +
+#   geom_col(color = "black") +
+#   scale_fill_manual(values = c("white", "gray80")) +
+#   facet_wrap(vars(component), scales = "free_y") +
+#   tidytext::scale_y_reordered() +
+#   theme_bw() +
+#   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value")
+```
+
+```{r plot-pca1-results}
+# plot_pca1_full <- function(x = "PC1", y = "PC2") {
+# 
+#   pcs <- c(x, y)
+# 
+#   pca1_all_plot <- pca1_all_aug %>%
+#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
+#     geom_point(data = select(pca1_all_aug, -period_label),
+#                alpha = 0.05, size = 0.4) +
+#     geom_point(aes(color = reg),
+#                alpha = 0.8, size = 0.5, show.legend = FALSE) +
+#     scale_color_manual(values = mycol[c(1, 2)]) +
+#     facet_grid(vars(period_label), vars(reg)) +
+#     theme_minimal()
+# 
+#   pca1_all_rot <-  pca1_all_rot %>%
+#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
+#     geom_segment(xend = 0, yend = 0, alpha = 0.4) +
+#     geom_text(
+#       aes(label = column),
+#       hjust = 1, nudge_x = -0.02,
+#       color = "#904C2F",
+#       size = 2.6
+#     ) +
+#     theme_minimal()
+# 
+#   pca1_all_load <- pca1_all_loadings %>%
+#     filter(component %in% pcs) %>%
+#     ggplot(aes(abs(value), column, fill = cor)) +
+#     geom_col(color = "black", show.legend = FALSE) +
+#     scale_fill_manual(values = c("white", "gray80")) +
+#     facet_wrap(vars(component), scales = "free_y") +
+#     tidytext::scale_y_reordered() +
+#     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") +
+#     theme_minimal()
+# 
+#   pca1_all_plot / (pca1_all_rot | pca1_all_load) +
+#     plot_layout(heights = c(6, 1))
+# }
+```
+
+```{r pca1-all-plots-pc1-pc2}
+# plot_pca1_full()
+# ggsave(here::here("analysis/figures/pca1_full_pc12.png"), height = 14)
+```
+
+```{r pca1-all-plots-pc3-pc4}
+# plot_pca1_full("PC3", "PC4")
+# ggsave(here::here("analysis/figures/pca1_full_pc34.png"), height = 14)
+```
+
+```{r pca1-all-plots-pc5-pc6}
+# plot_pca1_full("PC5", "PC6")
+# ggsave(here::here("analysis/figures/pca1_full_pc56.png"), height = 14)
+```
+
+```{r pca1-all-plots-pc7-pc8}
+# plot_pca1_full("PC7", "PC8")
+# ggsave(here::here("analysis/figures/pca1_full_pc78.png"), height = 14)
+```
+
+### Pottery groups
+
+```{r pca2-all}
+# # fit PCA
+# pca2_all_fit <- res2_wide %>%
+#   # select(-cont) %>%
+#   select(where(is.numeric)) %>%
+#   scale(center = TRUE, scale = TRUE) %>%
+#   prcomp()
+# 
+# # augmented fitted values
+# pca2_all_aug <- pca2_all_fit %>%
+#   broom::augment(res2_wide) %>%
+#   select(-".rownames") %>%
+#   rename_with(\(x) str_remove(x, ".fitted")) %>%
+#   settlements::add_region() %>%
+#   settlements::add_period_label() %>%
+#   settlements::add_neo()
+# 
+# # rotation matrix
+# pca2_all_rot <- pca2_all_fit %>%
+#   broom::tidy(matrix = "rotation") %>%
+#   mutate(PC = str_c("PC", PC)) %>%
+#   pivot_wider(names_from = PC)
+# 
+# # var explained
+# pca2_all_var_exp <- pca2_all_fit %>%
+#   broom::augment(res2_wide) %>%
+#   summarise(across(contains("PC"), var)) %>%
+#   pivot_longer(cols = everything(), names_to = "pc", values_to = "variance") %>%
+#   mutate(var_exp = variance/sum(variance),
+#          cum_var_exp = cumsum(var_exp),
+#          pc = str_replace(pc, ".fitted", "")) %>%
+#   select(pc, var_exp, cum_var_exp)
+# 
+# # PC cum var explained over 0.9
+# pca2_all_var_exp_over <- pca2_all_var_exp %>%
+#   mutate(n = !(cum_var_exp >= 0.9)) %>%
+#   pull(n) %>%
+#   sum() + 1
+```
+
+```{r pca2-all-var-exp-tab}
+# knitr::kable(pca1_all_var_exp, col.names = c("Comp.",
+#                                              "Prop. of variance",
+#                                              "Cumulative prop."))
+```
+
+```{r pca2-all-loadings-plot}
+# # PCA loadings
+# pca2_all_loadings <- pca2_all_fit %>%
+#   broom::tidy(matrix = "rotation") %>%
+#   mutate(component = str_c("PC", PC),
+#          component = fct_reorder(component, PC)) %>%
+#   filter(PC <= pca2_all_var_exp_over) %>%
+#   group_by(PC) %>%
+#   top_n(8, abs(value)) %>%
+#   ungroup() %>%
+#   mutate(cor = if_else(value > 0, "positive", "negative"),
+#          column = tidytext::reorder_within(column, abs(value), component))
+# 
+# pca2_all_loadings %>%
+#   ggplot(aes(abs(value), column, fill = cor)) +
+#   geom_col(color = "black") +
+#   scale_fill_manual(values = c("white", "gray80")) +
+#   facet_wrap(vars(component), scales = "free_y") +
+#   tidytext::scale_y_reordered() +
+#   theme_bw() +
+#   labs(x = "Abs. contribution value", y = element_blank(), fill = "Value")
+```
+
+```{r plot-pca2-results}
+# plot_pca2_full <- function(x = "PC1", y = "PC2") {
+# 
+#   pcs <- c(x, y)
+# 
+#   pca2_all_plot <- pca2_all_aug %>%
+#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
+#     geom_point(data = select(pca2_all_aug, -period_label),
+#                alpha = 0.05, size = 0.4) +
+#     geom_point(aes(color = reg),
+#                alpha = 0.8, size = 0.5, show.legend = FALSE) +
+#     scale_color_manual(values = mycol[c(1, 2)]) +
+#     facet_grid(vars(period_label), vars(reg)) +
+#     theme_minimal()
+# 
+#   pca2_all_rot <-  pca2_all_rot %>%
+#     ggplot(aes(!!sym(pcs[1]), !!sym(pcs[2]))) +
+#     geom_segment(xend = 0, yend = 0, alpha = 0.4) +
+#     geom_text(
+#       aes(label = column),
+#       hjust = 1, nudge_x = -0.02,
+#       color = "#904C2F",
+#       size = 2.6
+#     ) +
+#     theme_minimal()
+# 
+#   pca2_all_load <- pca2_all_loadings %>%
+#     filter(component %in% pcs) %>%
+#     ggplot(aes(abs(value), column, fill = cor)) +
+#     geom_col(color = "black", show.legend = FALSE) +
+#     scale_fill_manual(values = c("white", "gray80")) +
+#     facet_wrap(vars(component), scales = "free_y") +
+#     tidytext::scale_y_reordered() +
+#     labs(x = "Abs. contribution value", y = element_blank(), fill = "Value") +
+#     theme_minimal()
+# 
+#   pca2_all_plot / (pca2_all_rot | pca2_all_load) +
+#     plot_layout(heights = c(6, 1))
+# }
+```
+
+```{r pca2-all-plots-pc1-pc2}
+# plot_pca2_full()
+# ggsave(here::here("analysis/figures/pca2_full_pc12.png"), height = 14)
+```
+
+```{r pca2-all-plots-pc3-pc4}
+# plot_pca2_full("PC3", "PC4")
+# ggsave(here::here("analysis/figures/pca2_full_pc34.png"), height = 14)
+```
+
+```{r pca2-all-plots-pc5-pc6}
+# plot_pca2_full("PC5", "PC6")
+# ggsave(here::here("analysis/figures/pca2_full_pc56.png"), height = 14)
+```
+
+```{r pca2-all-plots-pc7-pc8}
+# plot_pca2_full("PC7", "PC8")
+# ggsave(here::here("analysis/figures/pca2_full_pc78.png"), height = 14)
+```
 
@@
 
-ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
+# ggsave(here::here("analysis/paper/temp/pca_skup_1.png"), width = 10, height = 21)
 ```
@@
 
-ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
+# ggsave(here::here("analysis/paper/temp/pca_skup_2.png"), width = 10, height = 21)
 ```
@@
 
-## Groups in the data
-
-This is not done for pottery traditions due to an error in one of the periods/
-regions that is caused by not enough variance in the PCA data. Weird...
-
-### Pottery groups
-
-```{r package-mclust}
-library(mclust)
-```
-
-```{r mclust2-neo}
-mb_clust2 <- pca2_neo %>% mbc_cluster()
-
-x <- mb_clust2 %>%
-  dplyr::select(reg, neo, classif) %>% 
-  tidyr::unnest() %>% 
-  dplyr::ungroup(c(reg, neo)) %>% 
-  dplyr::mutate(g = str_c(reg, g))
-
-set_spat %>% 
-  inner_join(x, by = "id") %>% 
-  ggplot() +
-  geom_sf(aes(color = g)) +
-  facet_grid(vars(g), vars(neo))
-
-```
-
-mb_classif2_neoB <- mb_clust2 %>% mbc_classify(neo = "NeoB")
-mb_classif2_neoC <- mb_clust2 %>% mbc_classify(neo = "NeoC")
-
-(mb_classif2_neoB == mb_classif2_neoC) %>% all()
-
-# bind_cols(mb_clustering_neo2$x[[2]][, 1:2], mb_clustering_neo2$classif[[2]]) %>% 
-#   ggplot(aes(PC1, PC2, color = factor(g))) +
-#   stat_ellipse() +
-#   geom_point() +
-#   theme_bw() +
-#   facet_wrap(vars(g))
-
-mb_map2_neoB <- mb_classif2 %>% mbc_map(sf = settlements2, neo = "NeoB")
-mb_map2_neoC <- mb_classif2 %>% mbc_map(sf = settlements2, neo = "NeoC")
-
-# boxploty pro proměnné ve skupinách...
-```
-
-```{r mclust2-map-neoB, fig.cap="Clusters in pottery groups in Neolithic B."}
-mb_map2_neoB
-```
-
-```{r mclust2-map-neoC, fig.cap="Clusters in pottery groups in Neolithic C."}
-mb_map2_neoC
-```
-
-```{r}
-res2_long %>% 
-filter(variable != "cont") %>% 
-add_neo() %>% 
-right_join(mb_classif2, by = c("id", "neo")) %>% 
-filter(neo == "NeoC") %>% 
-ggplot(aes(g, value, color = g))+
-geom_boxplot(show.legend = FALSE) +
-facet_wrap(vars(variable), scales = "free_y", ncol = 2) +
-theme_minimal()
-```
-
 <!--- ====================================================================== -->
 
-# Results
+<!-- # Results -->
 
 ```{r results-export}
-res1_long %>%
-distinct(id, period, variable, .keep_all = TRUE) %>% 
-pivot_wider(names_from = variable, values_from = value) %>% 
-arrange(id, period) %>% 
-write_csv(here::here(derived_data, "results", "results1.csv"))
-
-res2_long %>% 
-distinct(id, period, variable, .keep_all = TRUE) %>% 
-pivot_wider(names_from = variable, values_from = value) %>% 
-arrange(id, period) %>% 
-write_csv(here::here(derived_data, "results", "results2.csv"))
+# res1_long %>%
+# distinct(id, period, variable, .keep_all = TRUE) %>% 
+# pivot_wider(names_from = variable, values_from = value) %>% 
+# arrange(id, period) %>% 
+# write_csv(here::here(derived_data, "results", "results1.csv"))
+# 
+# res2_long %>% 
+# distinct(id, period, variable, .keep_all = TRUE) %>% 
+# pivot_wider(names_from = variable, values_from = value) %>% 
+# arrange(id, period) %>% 
+# write_csv(here::here(derived_data, "results", "results2.csv"))
 ```

### 2022-06-23 README.Rmd
update readme, namespace/desc after build

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Spatial analysis of Neolithic settlement patterns in central Europe"
 ```
@@
 
-> Trampota F., Pajdla P., (2021). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Trampota F., Pajdla P., (2022). _`r Title`_. Journal? <https://doi.org/xxx/xxx>
 
-Our pre-print is online here:
+The data paper accompanying this article:
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Pajdla P..
+Trampota F.
+(2021).
+_Neolithic Settlements in Central Europe: Data from the Project 'Lifestyle as an Unintentional Identity in the Neolithic'_.
+Journal of Open Archaeology Data 9, p.13.
+<http://doi.org/10.5334/joad.88>.
+
+Data is deposited in a Zenodo repository: <https://doi.org/10.5281/zenodo.5653180>.
 
@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+    generated by the analysis. Further, the analysis is divided into three 
+    parts with different granularity/detail in time.
+  - [:file\_folder: neo](/analysis/neo): Analysis for Neolithic B and C periods.
+  - [:file\_folder: neo](/analysis/traditions): Analysis for pottery traditions.
+  - [:file\_folder: neo](/analysis/groups): Analysis for pottery groups.
   - [:file\_folder: data](/analysis/data): Data used in the analysis. 
-  - [:file\_folder: figures](/analysis/figures): Plots and other
-    illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
+    We were not able to publish all our source data due to copyrights etc.
+    There is also an individual data paper: <http://doi.org/10.5334/joad.88>.
 
@@
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse, please refer to the individidual data publication at <http://doi.org/10.5334/joad.88>.
 

### 2022-06-23 README.Rmd
fix folder structure

@@
   - [:file\_folder: neo](/analysis/neo): Analysis for Neolithic B and C periods.
-  - [:file\_folder: neo](/analysis/traditions): Analysis for pottery traditions.
-  - [:file\_folder: neo](/analysis/groups): Analysis for pottery groups.
+  - [:file\_folder: traditions](/analysis/traditions): Analysis for pottery traditions.
+  - [:file\_folder: groups](/analysis/groups): Analysis for pottery groups.
   - [:file\_folder: data](/analysis/data): Data used in the analysis. 
